---
title: "Momentum-Driven Sector Rotation via Kalman Filtering and K-Means Clustering"
subtitle: "FIN452 Quantitative Trading Project"
author: "Carter Vereschagin"
date: today
format:
  html:
    embed-resources: true
    toc: false
    theme: flatly
    fig-width: 10
    fig-height: 6
    df-print: paged
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

```{r}
#| label: setup
#| include: false

# --- Data & Finance ---
library(tidyquant)          # tq_get() — Yahoo Finance + FRED wrapper
library(xts)                # time-series objects
library(PerformanceAnalytics) # charts.PerformanceSummary(), maxDrawdown()

# --- Kalman Filter ---
library(dlm)                # dlmModPoly(), dlmFilter()

# --- Parallelism ---
library(doParallel)         # makeCluster(), registerDoParallel()
library(foreach)            # foreach() %dopar%

# --- Visualization ---
library(plotly)             # interactive charts
library(ggplot2)
library(scales)

# --- Data Wrangling ---
library(dplyr)
library(tidyr)
library(purrr)
library(lubridate)
library(tibble)
library(zoo)                # na.locf() for forward-filling FRED data

# --- Display ---
library(knitr)
library(kableExtra)

# Global options
options(scipen = 999)
theme_set(theme_minimal(base_size = 12))

# ─── Global Constants ──────────────────────────────────────────────────────────
TRAIN_START    <- as.Date("2000-06-01")
TRAIN_END      <- as.Date("2024-12-31")
TEST_START     <- as.Date("2025-01-01")
TEST_END       <- as.Date("2026-03-31")
DOWNLOAD_START <- TRAIN_START - 30   # small buffer; no indicator warm-up needed

TC_RATE        <- 0.0005   # 0.05% one-way transaction cost

# All symbols (SPY = benchmark; remaining 11 = sectors)
SYMBOLS        <- c("SPY","XLB","XLC","XLE","XLF","XLI",
                    "XLK","XLP","XLRE","XLV","XLU","XLY")
SECTOR_SYMBOLS <- setdiff(SYMBOLS, "SPY")

# Kalman filter defaults (Q_velocity overridden by optimised value later)
DEFAULT_Q_LEVEL    <- 0.01
DEFAULT_Q_VELOCITY <- 0.001
DEFAULT_R_OBS      <- 1.0

# Strategy parameters
DEFAULT_K         <- 3                         # default number of k-means clusters
K_GRID            <- 2:6                       # grid search range for k
YC_THRESHOLD      <- -0.0025                   # 10Y-2Y < -0.25% → go to cash (all FLAT)

# Known ETF launch dates (important for data coverage disclosure)
XLC_LAUNCH  <- as.Date("2018-06-18")
XLRE_LAUNCH <- as.Date("2015-10-07")
```

---

# Strategy

## Description

This project implements a daily momentum-driven sector rotation strategy applied to the eleven S&P 500 Select Sector SPDR ETFs — XLB, XLC, XLE, XLF, XLI, XLK, XLP, XLRE, XLV, XLU, and XLY — benchmarked against SPY. The strategy runs every trading day. A Kalman filter is applied to each sector's adjusted close price series to extract a noise-filtered trend velocity. Those velocities are then passed into a k-means clustering algorithm that partitions all eleven sectors into momentum regimes. The cluster with the highest average velocity — the Bull cluster — defines which sectors to hold that day. When the FRED 10-year minus 2-year Treasury spread falls below −0.25%, indicating yield curve inversion, the strategy exits all positions entirely and holds cash. Positions are sized equally among Bull cluster sectors and executed at the following morning's market open.

The training period spans June 2000 through December 2024 — 24.5 years including two major bear markets, the global financial crisis, a pandemic crash, and a rate hike cycle. The out-of-sample test period covers January 2025 through March 2026 and was completely sealed during model development. Two free parameters, the number of clusters $k$ and the Kalman velocity noise $Q_2$, are jointly optimized on training data using a 5 × 12 grid search evaluated by Omega ratio.

## Rationale

The core intuition is that equity sectors exhibit persistent momentum. Sectors trending strongly today tend to continue trending strongly in the near term, and sectors in downtrends tend to remain weak. Rather than attempting to forecast which sectors will outperform based on macroeconomic views, this strategy simply measures trend momentum as objectively as possible and holds the strongest-trending sectors each day.

The Kalman filter's role is noise reduction before clustering. Raw daily returns are dominated by single-day shocks — macro headlines, fund flows, temporary imbalances — that do not reflect genuine trend reversals. Clustering on raw returns generates excessive turnover as sectors bounce in and out of the Bull cluster on noise rather than signal. The Kalman velocity filters out these transient shocks by maintaining a running estimate of each sector's underlying trend speed, updating conservatively when a new price is consistent with the prior trend and more aggressively when a sustained directional move accumulates. Clustering on this denoised velocity produces more stable daily assignments and lower unnecessary turnover.

The yield curve overlay provides a macro-grounded defensive mechanism. Rather than trying to pick which sectors will hold up in a recessionary environment — a second prediction problem layered on top of the momentum signal — I simply exit the market entirely when the probability of a broad economic contraction is elevated. The 10Y-2Y spread has preceded every US recession since the 1960s, making it the cleanest and most widely-validated leading indicator available at daily frequency without data access costs.

## Research

Moskowitz & Grinblatt (1999, *Journal of Finance*) demonstrated that industry-level momentum is statistically significant and economically large, and that the profits are concentrated on the long side — validating a long-only implementation. Faber (2010, SSRN 1585517) tested relative-strength sector rotation on SPDR-equivalent data back to the 1920s and found it profitable across all lookback windows from one to twelve months, with a trend-following overlay reducing drawdowns substantially. Moskowitz, Ooi & Pedersen (2012, *Journal of Financial Economics*) established time-series momentum — each asset's own past return predicting its own future return — providing the theoretical basis for using each sector's individual velocity signal rather than only cross-sectional comparisons.

For the macro overlay, Estrella & Mishkin (1998, *Review of Economics and Statistics*) documented the yield curve's predictive power for US recessions, and the Chicago Fed (2018) confirmed that the 10Y-2Y spread has an essentially perfect historical track record as a leading recession indicator. This body of evidence grounds the yield curve filter in peer-reviewed research rather than ad hoc rule-setting.

---

# Model Implementation

## Data

```{r}
#| label: data_download
#| include: false
#| cache: true

# Download all 12 equity symbols
raw_prices <- tq_get(
  SYMBOLS,
  get  = "stock.prices",
  from = DOWNLOAD_START,
  to   = TEST_END
)

# Download FRED 10Y-2Y Treasury yield spread (T10Y2Y series)
# Positive = normal curve; negative = inverted (recession signal)
yc_raw <- tq_get(
  "T10Y2Y",
  get  = "economic.data",
  from = DOWNLOAD_START,
  to   = TEST_END
) %>%
  rename(yield_spread = price) %>%
  mutate(yield_spread = yield_spread / 100)   # convert from percentage points to decimal
```

I download daily OHLCV price data for all eleven sector ETFs and SPY from Yahoo Finance using `tidyquant::tq_get()`, and the FRED T10Y2Y series for the yield curve overlay. XLC (Communication Services) launched on `r format(XLC_LAUNCH, "%B %d, %Y")` and XLRE (Real Estate) launched on `r format(XLRE_LAUNCH, "%B %d, %Y")`; prior to those dates those sectors are simply absent from the clustering universe, and k-means operates on the available subset.

```{r}
#| label: data_quality

coverage <- raw_prices %>%
  group_by(symbol) %>%
  summarise(
    first_date  = min(date),
    last_date   = max(date),
    n_rows      = n(),
    n_na_adj    = sum(is.na(adjusted)),
    n_na_open   = sum(is.na(open)),
    .groups     = "drop"
  ) %>%
  arrange(first_date)

kable(coverage,
      col.names = c("Symbol","First Date","Last Date","Rows","NA (Adjusted)","NA (Open)"),
      caption   = "Table 1: Data Coverage by Symbol",
      format    = "html") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed")) %>%
  row_spec(which(coverage$symbol %in% c("XLC","XLRE")), background = "#fff3cd")
```

I use the `adjusted` close from Yahoo Finance for the Kalman filter and C2C returns. Yahoo computes adjusted prices using the full forward history of dividends and splits, which introduces a minor look-ahead bias in the feature engineering. This is standard practice in academic finance and the impact is small for sector ETFs, which pay modest quarterly dividends and rarely split.

```{r}
#| label: data_split
#| include: false

# Training data (June 2000 – Dec 2024)
train_all    <- raw_prices %>% filter(date >= TRAIN_START, date <= TRAIN_END)
sector_train <- train_all %>% filter(symbol %in% SECTOR_SYMBOLS)
spy_train    <- train_all %>% filter(symbol == "SPY")

# Testing data (Jan 2025 – Mar 2026)
test_all    <- raw_prices %>% filter(date >= TEST_START, date <= TEST_END)
sector_test <- test_all %>% filter(symbol %in% SECTOR_SYMBOLS)
spy_test    <- test_all %>% filter(symbol == "SPY")

# Yield curve split — FRED data has gaps (weekends/holidays); forward-fill to trading days
# The full series is kept unsplit here; filtering happens inside signal generation
yc_filled <- yc_raw %>%
  arrange(date) %>%
  mutate(yield_spread = zoo::na.locf(yield_spread, na.rm = FALSE))

cat("Training rows (sectors):", nrow(sector_train), "\n")
cat("Testing rows (sectors):", nrow(sector_test), "\n")
cat("Training date range:", format(min(train_all$date)), "–", format(max(train_all$date)), "\n")
cat("Testing date range:",  format(min(test_all$date)),  "–", format(max(test_all$date)),  "\n")
```

The training and test splits are created here and the test set is not accessed again until the Performance section. All model fitting, parameter selection, and normalization are derived exclusively from training data.

## Indicators

```{r}
#| label: helper_functions
#| include: false

# ─── Kalman Filter ─────────────────────────────────────────────────────────────
run_kalman <- function(price_vec, q_level = DEFAULT_Q_LEVEL,
                        q_velocity = DEFAULT_Q_VELOCITY, r_obs = DEFAULT_R_OBS) {
  # Local level + slope state-space model
  # State: [level, velocity]; Observation: price
  # IMPORTANT: uses dlmFilter (causal), NOT dlmSmooth (uses future data)
  mod  <- dlmModPoly(order = 2, dV = r_obs, dW = c(q_level, q_velocity))
  filt <- dlmFilter(price_vec, mod)
  # filt$m is (n+1) x 2: row 1 = prior state; rows 2:(n+1) = posterior after each obs
  tibble(
    level    = filt$m[-1, 1],  # filtered level aligned with price_vec
    velocity = filt$m[-1, 2]   # filtered velocity (daily rate of change of level)
  )
}

# ─── Omega Ratio ───────────────────────────────────────────────────────────────
# Non-parametric risk-reward measure; accounts for all moments of the distribution
# Omega > 1: expected gains exceed expected losses at threshold t
# More robust than Sharpe when return distribution is non-normal
omega_ratio <- function(returns, threshold = 0) {
  r <- returns[!is.na(returns)]
  excess <- r - threshold
  gains  <- sum(excess[excess > 0])
  losses <- abs(sum(excess[excess < 0]))
  if (losses == 0) return(Inf)
  gains / losses
}

# ─── Daily Signal Generation ──────────────────────────────────────────────────
# Runs k-means on Kalman velocity every trading day (single feature, k clusters).
# Bull cluster = cluster with highest mean kal_velocity → LONG signal.
# Yield curve override: if 10Y-2Y < threshold → ALL sectors FLAT (go to cash).
# Signal at Close_T → executed at Open_{T+1} via 1-day lag in build_portfolio.
# Returns one row per sector per trading day.
generate_signals_clustered <- function(kal_df, yc_df,
                                       k_clusters = DEFAULT_K,
                                       yc_thresh  = YC_THRESHOLD) {
  # Forward-fill yield curve to every trading day
  all_trading_days <- sort(unique(kal_df$date))
  yc_daily <- tibble(date = all_trading_days) %>%
    left_join(yc_df %>% select(date, yield_spread), by = "date") %>%
    arrange(date) %>%
    mutate(yield_spread = zoo::na.locf(yield_spread, na.rm = FALSE))

  # Join yield curve into Kalman data
  kal_sector <- kal_df %>%
    filter(symbol %in% SECTOR_SYMBOLS) %>%
    left_join(yc_daily, by = "date")

  # For each trading day: k-means on kal_velocity, identify Bull cluster
  result_list <- lapply(all_trading_days, function(d) {
    df_d <- kal_sector %>% filter(date == d)
    if (nrow(df_d) < 2) return(NULL)

    yc_inv <- isTRUE(!is.na(df_d$yield_spread[1]) && df_d$yield_spread[1] < yc_thresh)

    # Yield curve inverted → all flat (cash), skip clustering
    if (yc_inv) {
      return(df_d %>%
        mutate(cluster = NA_integer_, yc_inverted = TRUE, signal = "FLAT") %>%
        select(symbol, date, signal, cluster, kal_velocity, yield_spread, yc_inverted))
    }

    # Guard: cannot have more clusters than observations
    k_actual <- min(k_clusters, nrow(df_d))
    km       <- kmeans(matrix(df_d$kal_velocity, ncol = 1),
                       centers = k_actual, nstart = 25, iter.max = 50)
    df_d$cluster <- as.integer(km$cluster)

    cluster_means <- tapply(df_d$kal_velocity, df_d$cluster, mean)
    bull_cluster  <- as.integer(names(which.max(cluster_means)))

    df_d %>%
      mutate(
        yc_inverted = FALSE,
        signal      = if_else(cluster == bull_cluster, "LONG", "FLAT")
      ) %>%
      select(symbol, date, signal, cluster, kal_velocity, yield_spread, yc_inverted)
  })

  bind_rows(result_list)
}

# ─── Portfolio Construction ────────────────────────────────────────────────────
# Signal at Close_T → executed at Open_{T+1} via 1-day lag.
# Return type depends on position transition:
#   FLAT→LONG (entry) : O2C — bought at open, held to close
#   LONG→LONG (hold)  : C2C — held from previous close to this close
#   LONG→FLAT (exit)  : C2O — held from previous close, sold at open
build_portfolio <- function(signals_df, returns_df, tc_rate = TC_RATE) {
  # Lag signal: today's exec_signal = yesterday's cluster signal
  daily <- signals_df %>%
    arrange(symbol, date) %>%
    group_by(symbol) %>%
    mutate(exec_signal = dplyr::lag(signal, default = "FLAT")) %>%
    ungroup() %>%
    left_join(returns_df %>% select(symbol, date, o2c_ret, c2c_ret, c2o_ret),
              by = c("symbol", "date")) %>%
    filter(!is.na(c2c_ret))

  # Compute equal weights based on exec_signal
  daily <- daily %>%
    group_by(date) %>%
    mutate(
      n_long = sum(exec_signal == "LONG"),
      weight = if_else(exec_signal == "LONG" & n_long > 0L, 1 / n_long, 0)
    ) %>%
    ungroup()

  # Determine position transition and select correct return type
  # prev_exec = what we were doing yesterday (did we already hold this position?)
  daily <- daily %>%
    group_by(symbol) %>%
    arrange(date, .by_group = TRUE) %>%
    mutate(
      prev_exec   = dplyr::lag(exec_signal, default = "FLAT"),
      prev_weight = dplyr::lag(weight, default = 0),
      position_ret = case_when(
        prev_exec == "FLAT" & exec_signal == "LONG" ~ o2c_ret,  # entry: Open→Close
        prev_exec == "LONG" & exec_signal == "LONG" ~ c2c_ret,  # hold:  Close→Close
        prev_exec == "LONG" & exec_signal == "FLAT" ~ c2o_ret,  # exit:  Close→Open
        TRUE                                        ~ 0
      ),
      # Exit days: position was held at prev_weight, now closing; use prev_weight for return
      eff_weight = if_else(prev_exec == "LONG" & exec_signal == "FLAT",
                           prev_weight, weight),
      tc_cost = abs(weight - prev_weight) * tc_rate
    ) %>%
    ungroup()

  daily %>%
    group_by(date) %>%
    summarise(
      port_ret_gross = sum(eff_weight * replace_na(position_ret, 0), na.rm = TRUE),
      total_tc       = sum(tc_cost, na.rm = TRUE),
      n_positions    = sum(weight > 0),
      .groups        = "drop"
    ) %>%
    mutate(port_ret_net = port_ret_gross - total_tc)
}

# ─── Performance Metrics ───────────────────────────────────────────────────────
compute_metrics <- function(port_ret_vec, spy_ret_vec = NULL, n_sectors = 11,
                            label = "", tc_rate = TC_RATE) {
  r    <- port_ret_vec[!is.na(port_ret_vec)]
  n    <- length(r)
  ann  <- 252

  cagr    <- prod(1 + r)^(ann / n) - 1
  sr      <- if (isTRUE(sd(r) > 0)) mean(r) / sd(r) * sqrt(ann) else NA_real_
  om      <- omega_ratio(r)
  cum_r   <- cumprod(1 + r)
  peak    <- cummax(cum_r)
  dds     <- (cum_r - peak) / peak
  max_dd  <- min(dds)
  in_dd   <- dds < -1e-8
  dd_rle  <- rle(in_dd)
  dd_lens <- dd_rle$lengths[dd_rle$values]
  max_dd_len <- if (length(dd_lens) > 0) max(dd_lens) else 0L
  pct_win    <- mean(r > 0, na.rm = TRUE)

  cor_spy <- if (!is.null(spy_ret_vec)) {
    s <- na.omit(cbind(r, spy_ret_vec[seq_along(r)]))
    cor(s[, 1], s[, 2])
  } else NA_real_

  tibble(
    Period          = label,
    `CAGR (%)`      = round(cagr * 100, 2),
    `Sharpe`        = round(sr, 3),
    `Omega`         = round(om, 3),
    `Max DD (%)`    = round(max_dd * 100, 2),
    `Max DD (days)` = max_dd_len,
    `% Win`         = round(pct_win * 100, 1),
    `Cor SPY`       = round(cor_spy, 3)
  )
}
```

I apply a Kalman filter to each sector's daily adjusted close price history to extract a noise-filtered trend velocity. The model specification is a local level plus slope (second-order polynomial) state-space model. The two hidden states are the sector's true underlying price level and its velocity — the rate at which that level is changing. Each new daily price updates both estimates in proportion to how surprising the observation was relative to the prior expectation.

Three key assumptions underpin the model. First, linearity: the observation is simply the true level plus noise, which holds well for daily prices. Second, Gaussian noise: both the process disturbances and the measurement error are normally distributed. Equity returns have fat tails, so this is an approximation — but the Kalman filter remains effective as a smoother even when normality is imperfect. Third, stationarity of noise variances: the process noise ($Q_1$, $Q_2$) and observation noise ($R$) are assumed constant over time. In practice volatility regimes shift, which is why $Q_2$ is treated as a free parameter to be optimized rather than fixed arbitrarily.

Three noise variance parameters control filter behavior. $Q_1$ governs how much random variation is allowed in the underlying price level day to day — fixed at a small value since the true level changes smoothly. $Q_2$ controls how quickly the estimated trend velocity can change, and is the most important tuning parameter: a small $Q_2$ produces a slow-reacting velocity with long memory; a large $Q_2$ makes the velocity react more quickly to recent moves. $R$ is the observation noise, fixed at 1.0 to establish the scaling relationship. The ratio $Q_2 / R$ ultimately drives how responsive the velocity estimate is.

A critical distinction: I use `dlmFilter`, the forward-only algorithm, not `dlmSmooth`. The filter estimates the state on day $t$ using only prices observed through day $t$ — exactly how a real-time system would operate. The smoother runs a backward pass through the full series after it has been observed, incorporating future prices to improve past estimates. Using the smoother in a backtest introduces look-ahead bias. Only `dlmFilter` is used here.

The velocity output is the best estimate at each point in time of how fast the sector's underlying trend is moving. A positive and rising velocity means the sector is accelerating upward; a negative velocity means the trend is pointed downward. Compared to raw returns, the Kalman velocity has three advantages as a clustering feature: it absorbs single-day shocks gradually rather than reversing the signal on one bad headline; it places more weight on recent observations when uncertainty is high and more weight on the prior when conditions are stable; and it produces one clean interpretable number per sector per day, avoiding the arbitrary horizon selection required when constructing lookback return features.

```{r}
#| label: apply_kalman_train
#| cache: true
#| include: false

apply_kalman <- function(df, q_vel = DEFAULT_Q_VELOCITY) {
  kal <- run_kalman(df$adjusted, q_velocity = q_vel)
  df %>%
    mutate(
      kal_level    = kal$level,
      kal_velocity = kal$velocity,
      # Return types needed for portfolio construction
      c2c_ret = adjusted / dplyr::lag(adjusted) - 1,
      o2c_ret = close    / open - 1,
      c2o_ret = open     / dplyr::lag(close)    - 1
    )
}

kalman_train <- sector_train %>%
  group_by(symbol) %>%
  arrange(date, .by_group = TRUE) %>%
  group_modify(~ apply_kalman(.x)) %>%
  ungroup()

# Return table for portfolio construction (training period only)
returns_train <- kalman_train %>%
  filter(date >= TRAIN_START, date <= TRAIN_END) %>%
  select(symbol, date, c2c_ret, o2c_ret, c2o_ret)

# SPY benchmark returns
spy_ret_train <- spy_train %>%
  arrange(date) %>%
  mutate(spy_c2c = adjusted / dplyr::lag(adjusted) - 1) %>%
  filter(!is.na(spy_c2c)) %>%
  select(date, spy_c2c)
```

## Signals

On each trading day, I run k-means clustering on the eleven Kalman velocity values. K-means partitions them into $k$ groups by minimizing total within-cluster variance on that single dimension. The group with the highest mean velocity is the Bull cluster — those sectors receive a LONG signal executed at the following day's open. All other sectors receive a FLAT signal.

Using only the Kalman velocity as the clustering feature keeps the regime classification economically interpretable. The algorithm answers one question: which sectors have the strongest upward trend momentum right now? There is no risk of clustering artifacts from mixing features at different scales, and the result — a hard Bull/non-Bull partition — is straightforward to audit and explain. For a one-dimensional feature with eleven observations per day, k-means is both optimal and efficient; the cluster count $k$ is treated as a free parameter optimized jointly with $Q_2$.

When the 10-year minus 2-year Treasury spread (FRED `T10Y2Y`) falls below −0.25%, the strategy skips clustering entirely and holds 100% cash. Rather than forecasting which sectors will be resilient in a recession, I simply remove all market exposure. This is the most conservative response to a yield curve inversion — it avoids the drawdown from holding risky sectors during a broad market selldown, and the historical evidence for inversion preceding recessions is robust enough to justify the rule without further elaboration.

```{r}
#| label: default_ranking_inspection

# Default model: DEFAULT_K clusters, DEFAULT_Q_VELOCITY
train_signals_default <- generate_signals_clustered(
  kalman_train, yc_filled,
  k_clusters = DEFAULT_K, yc_thresh = YC_THRESHOLD
)

# Sector frequency in Bull cluster by year
rank_freq <- train_signals_default %>%
  mutate(year = year(date)) %>%
  group_by(symbol, year) %>%
  summarise(
    pct_selected = mean(signal == "LONG", na.rm = TRUE),
    .groups = "drop"
  )

hm_wide <- rank_freq %>%
  pivot_wider(names_from = year, values_from = pct_selected, values_fill = 0)
hm_mat  <- as.matrix(hm_wide[, -1])
rownames(hm_mat) <- hm_wide$symbol

plot_ly(
  x          = colnames(hm_mat),
  y          = rownames(hm_mat),
  z          = hm_mat,
  type       = "heatmap",
  colorscale = list(c(0,"#f5f5f5"), c(0.5,"#42a5f5"), c(1,"#1565c0")),
  zmin = 0, zmax = 1,
  hovertemplate = "%{y} | %{x}<br>Selected: %{z:.0%}<extra></extra>"
) %>%
  plotly::layout(
    title  = "Figure 1: Bull-Cluster Selection Frequency by Year (Training, Default Parameters)",
    xaxis  = list(title = ""),
    yaxis  = list(title = ""),
    margin = list(l = 60)
  )
```

Figure 1 shows which sectors were in the Bull cluster most frequently in each year of the training period under default parameters ($k$ = `r DEFAULT_K`, $Q_2$ = `r DEFAULT_Q_VELOCITY`). Warmer colors indicate more frequent selection. Sectors that dominate in specific years — Energy during oil bull markets, Technology during the 2010s growth cycle — reflect the strategy's momentum-driven nature: it concentrates in whatever is actually trending, rather than allocating based on fixed sector views.

The complete daily decision process is summarized in the diagrams below.

```{mermaid}
%%{init: {'theme': 'base', 'themeVariables': {'primaryColor': '#e3f2fd', 'primaryTextColor': '#0d1b2a', 'primaryBorderColor': '#1565c0', 'lineColor': '#1565c0', 'secondaryColor': '#fff3e0', 'tertiaryColor': '#ffffff'}}}%%
flowchart TD
    A([Daily Close]):::start --> B[Apply Kalman Filter\ndlmFilter — forward pass only]
    B --> C[Extract kal_velocity\nnoise-filtered trend slope]
    C --> D{Yield curve inverted?\n10Y-2Y < -0.25%}
    D -- Yes --> E[ALL FLAT\nGo to cash]:::exit
    D -- No --> F[K-Means Clustering\nk clusters on kal_velocity]
    F --> G[Identify Bull Cluster\nhighest mean velocity]:::bull
    G --> H[LONG sectors in Bull Cluster\nequal weight]:::bull
    E --> I[Apply TC: 0.05%\nper unit weight change]
    H --> I
    I --> J([Execute at Open T+1]):::start
    J --> A
    classDef start fill:#1565c0,stroke:#0d47a1,color:#ffffff
    classDef bull  fill:#c8e6c9,stroke:#2e7d32,color:#1b5e20
    classDef exit  fill:#ffecb3,stroke:#f57f17,color:#4e3400
```

```{mermaid}
%%{init: {'theme': 'base', 'themeVariables': {'primaryColor': '#e3f2fd', 'primaryTextColor': '#0d1b2a', 'primaryBorderColor': '#1565c0', 'lineColor': '#555555'}}}%%
flowchart LR
    D1[Daily OHLCV\ntq_get]:::data --> K[Kalman Filter\ndlmFilter]:::kalman
    D2[FRED T10Y2Y\nyield spread]:::data --> YC{YC inverted?}:::decision
    K --> CL[K-Means Clustering\nk clusters on kal_velocity]:::cluster
    CL --> YC
    YC -- Yes --> FL[ALL FLAT\nCash]:::exit
    YC -- No  --> LG[LONG Bull Cluster\nEqual weight]:::bull
    FL --> PT[Portfolio\nLag 1 day · TC 0.05%]:::port
    LG --> PT
    PT --> OP[Grid Search\nk x Q · 60 combos]:::opt
    OP --> EV[Evaluate\nOmega · Sharpe · MaxDD]:::eval
    classDef data     fill:#fff8e1,stroke:#f9a825,color:#4e3400
    classDef kalman   fill:#e3f2fd,stroke:#1565c0,color:#0d47a1
    classDef cluster  fill:#f3e5f5,stroke:#6a1b9a,color:#4a148c
    classDef decision fill:#ffffff,stroke:#555555,color:#000000
    classDef bull     fill:#c8e6c9,stroke:#2e7d32,color:#1b5e20
    classDef exit     fill:#ffecb3,stroke:#f57f17,color:#4e3400
    classDef port     fill:#e8eaf6,stroke:#283593,color:#1a237e
    classDef opt      fill:#fce4ec,stroke:#880e4f,color:#4a0020
    classDef eval     fill:#e0f2f1,stroke:#00695c,color:#003d36
```

| Condition | Signal | Execution Timing | Return Captured |
|---|---|---|---|
| Yield curve inverted (10Y-2Y < -0.25%) | ALL FLAT | Exit all positions at Open$_{T+1}$ | None (hold cash) |
| Yield curve normal AND sector in Bull cluster | LONG | Buy at Open$_{T+1}$ | Open-to-Close on $T+1$ |
| Yield curve normal AND sector NOT in Bull cluster | FLAT | Sell (if held) at Open$_{T+1}$ | None (cash = 0%) |

## Trades

Signals are generated at each day's close and executed at the following day's market open. I observe the cluster assignment at close, place orders overnight, and fill at the open. A position is only entered or exited when the signal actually changes — if a sector stays in the Bull cluster for ten consecutive days, there is no trade on days two through ten.

The return type applied on each day depends on the position transition:

| Transition | Return Type | Formula | Rationale |
|---|---|---|---|
| FLAT → LONG (entry) | Open-to-Close (O2C) | Close / Open − 1 | Bought at open, held to close. No overnight exposure on entry day. |
| LONG → LONG (hold) | Close-to-Close (C2C) | Adj Close / Lag(Adj Close) − 1 | Already held at previous close. Earns full overnight gap plus intraday move. |
| LONG → FLAT (exit) | Close-to-Open (C2O) | Open / Lag(Close) − 1 | Signal flipped overnight. Position sold at open next day — earns only the overnight gap, not the intraday move. |

This three-way decomposition is important for backtest realism. A strategy that always uses C2C returns overstates performance by crediting entry days with overnight gaps that were not captured. Using O2C for hold days misses the overnight gap that a real held position earns. O2C and C2O use raw unadjusted open and close prices; C2C uses adjusted close for accurate long-horizon compounding.

Transaction costs are 0.05% one-way, applied as $|\Delta w_i| \times 0.0005$ per dollar rebalanced. This covers approximately 0.03% bid-ask spread and 0.02% commission, consistent with liquid sector ETFs. Total round-trip cost per entry/exit is 0.10%. Results are reported both gross and net.

## Training Period

The strategy has two free parameters — $k$ (number of clusters) and $Q_2$ (Kalman velocity noise) — optimized jointly on training data by evaluating the Omega ratio across a 5 × 12 = 60 combination grid. $k$ ranges from 2 to 6; $Q_2$ is evaluated at 12 log-spaced values spanning five orders of magnitude from 0.00001 to 0.1. This fine grid captures subtle differences in filter responsiveness across the full spectrum from very slow (long-memory) to very fast (reactive). The grid runs in parallel across available CPU cores.

I optimize for Omega ratio rather than cumulative return because cumulative return would simply reward parameters that kept the portfolio most invested during the 2009–2021 bull market, penalizing the yield curve cash periods that are the strategy's core risk management mechanism. Omega ratio measures whether the strategy generates more dollar-weighted upside than downside at zero return, accounting for all moments of the distribution rather than just mean and variance as Sharpe does.

```{r}
#| label: param_grid_setup
#| include: false

k_grid     <- K_GRID                                                                              # 2, 3, 4, 5, 6
q_grid     <- c(0.00001, 0.00005, 0.0001, 0.0003, 0.0005, 0.001, 0.003, 0.005, 0.01, 0.03, 0.05, 0.1)  # 12 log-spaced values
param_grid <- expand.grid(k = k_grid, q = q_grid)                                                # 5 × 12 = 60 combinations
```

```{r}
#| label: parallel_optimization
#| cache: true
#| include: false

cl <- makeCluster(parallel::detectCores() - 1)
registerDoParallel(cl)

clusterExport(cl, varlist = c(
  "run_kalman", "apply_kalman", "generate_signals_clustered", "build_portfolio",
  "omega_ratio",
  "sector_train", "yc_filled",
  "TRAIN_START", "TRAIN_END", "TC_RATE",
  "SECTOR_SYMBOLS", "YC_THRESHOLD",
  "DEFAULT_Q_LEVEL", "DEFAULT_R_OBS", "DEFAULT_K"
), envir = .GlobalEnv)

opt_results <- foreach(
  idx      = 1:nrow(param_grid),
  .combine = "rbind",
  .packages = c("dlm", "dplyr", "tidyr", "tibble", "purrr", "lubridate", "zoo")
) %dopar% {

  ki <- param_grid$k[idx]
  qi <- param_grid$q[idx]

  tryCatch({
    # 1. Apply Kalman with this Q_velocity
    kal_df <- sector_train %>%
      dplyr::group_by(symbol) %>%
      dplyr::arrange(date, .by_group = TRUE) %>%
      dplyr::group_modify(~ apply_kalman(.x, q_vel = qi)) %>%
      dplyr::ungroup()

    # 2. Generate daily signals (k-means clustering on kal_velocity + YC filter)
    sigs <- generate_signals_clustered(kal_df, yc_filled,
                                       k_clusters = ki, yc_thresh = YC_THRESHOLD)

    # 3. Returns table for this Kalman parameterisation
    ret_df <- kal_df %>%
      dplyr::filter(date >= TRAIN_START, date <= TRAIN_END) %>%
      dplyr::select(symbol, date, c2c_ret, o2c_ret, c2o_ret)

    # 4. Build portfolio and evaluate
    port <- build_portfolio(sigs, ret_df, tc_rate = TC_RATE)
    r    <- na.omit(port$port_ret_net)

    if (length(r) < 50) {
      tibble::tibble(k = ki, q = qi, omega = NA_real_, sharpe = NA_real_,
                     max_dd = NA_real_, pct_win = NA_real_, n_obs = length(r))
    } else {
      cr <- cumprod(1 + r); pk <- cummax(cr)
      tibble::tibble(
        k       = ki,
        q       = qi,
        omega   = omega_ratio(r),
        sharpe  = mean(r) / sd(r) * sqrt(252),
        max_dd  = min((cr - pk) / pk),
        pct_win = mean(r > 0),
        n_obs   = length(r)
      )
    }
  }, error = function(e) {
    tibble::tibble(k = ki, q = qi, omega = NA_real_, sharpe = NA_real_,
                   max_dd = NA_real_, pct_win = NA_real_, n_obs = 0L)
  })
}

stopCluster(cl)
```

```{r}
#| label: surface_plot_omega

surface_wide <- opt_results %>%
  mutate(log_q = log10(q)) %>%
  select(k, log_q, omega) %>%
  pivot_wider(names_from = log_q, values_from = omega)

z_mat <- as.matrix(surface_wide[, -1])
rownames(z_mat) <- surface_wide$k

plot_ly(
  x          = sort(unique(log10(opt_results$q))),
  y          = sort(unique(opt_results$k)),
  z          = z_mat,
  type       = "surface",
  colorscale = list(c(0,"#d32f2f"), c(0.5,"#ffeb3b"), c(1,"#1565c0")),
  showscale  = TRUE
) %>%
  plotly::layout(
    title = list(text = "Figure 2: Omega Ratio Surface — Parameter Grid (Training Period)"),
    scene = list(
      xaxis = list(title = "log\u2081\u2080(Kalman Q\u2082 velocity noise)"),
      yaxis = list(title = "Clusters (k)"),
      zaxis = list(title = "Omega Ratio")
    )
  )
```

Figure 2 shows the Omega ratio surface across all 60 parameter combinations. A broad plateau indicates robustness — the strategy performs similarly across a range of parameter values and is not sensitive to the exact grid location of the optimum. A sharp isolated spike would suggest overfitting to a specific historical period.

```{r}
#| label: optimal_params
#| include: false

best_row <- opt_results %>%
  filter(!is.na(omega)) %>%
  slice_max(omega, n = 1)

OPTIMAL_K <- best_row$k[1]
OPTIMAL_Q <- best_row$q[1]

cat("Optimal parameters (training Omega criterion):\n")
cat("  k (clusters)      :", OPTIMAL_K, "\n")
cat("  Q (velocity noise):", OPTIMAL_Q, "\n")
cat("  Training Omega    :", round(best_row$omega[1], 3), "\n")
cat("  Training Sharpe   :", round(best_row$sharpe[1], 3), "\n")
```

The optimizer selects $k$ = `r OPTIMAL_K` clusters and $Q_2$ = `r OPTIMAL_Q`. These parameters are frozen and applied without adjustment to the test period.

```{r}
#| label: zscored_metrics

zscored <- opt_results %>%
  filter(!is.na(omega)) %>%
  mutate(
    z_omega   = as.vector(scale(omega)),
    z_sharpe  = as.vector(scale(sharpe)),
    z_max_dd  = as.vector(scale(-max_dd)),
    z_pct_win = as.vector(scale(pct_win))
  ) %>%
  arrange(desc(z_omega)) %>%
  select(k, q, omega, sharpe, max_dd, pct_win, z_omega, z_sharpe, z_max_dd, z_pct_win)

kable(zscored,
      digits  = 3,
      caption = "Table 3: Z-Scored Risk Metrics Across Parameter Grid (Training)",
      format  = "html",
      col.names = c("k (clusters)","Q","Omega","Sharpe","Max DD","% Win",
                    "z(Omega)","z(Sharpe)","z(Max DD)","z(% Win)")) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed")) %>%
  row_spec(1, bold = TRUE, background = "#e3f2fd")
```

Table 3 z-scores all metrics relative to their grid-wide mean and standard deviation, allowing comparison across metrics that are on different scales. Values above zero are above the grid average; values above 1.0 are more than one standard deviation above average. The top-ranked row is the selected parameter combination.

```{r}
#| label: refit_optimal
#| cache: true
#| include: false

# Apply Kalman with OPTIMAL_Q to full training data
kalman_train_opt <- sector_train %>%
  group_by(symbol) %>%
  arrange(date, .by_group = TRUE) %>%
  group_modify(~ apply_kalman(.x, q_vel = OPTIMAL_Q)) %>%
  ungroup()

# Generate daily signals with OPTIMAL_K and OPTIMAL_Q
train_signals_opt <- generate_signals_clustered(
  kalman_train_opt, yc_filled,
  k_clusters = OPTIMAL_K, yc_thresh = YC_THRESHOLD
)

# Returns for training period
returns_train_opt <- kalman_train_opt %>%
  filter(date >= TRAIN_START, date <= TRAIN_END) %>%
  select(symbol, date, c2c_ret, o2c_ret, c2o_ret)

# Portfolio
portfolio_train <- build_portfolio(train_signals_opt, returns_train_opt, tc_rate = TC_RATE)

# Merge with SPY benchmark (spy_ret_train built in apply_kalman_train chunk)
train_combined <- portfolio_train %>%
  left_join(spy_ret_train, by = "date") %>%
  filter(!is.na(spy_c2c))
```

The in-sample results below are presented for transparency and model calibration only. Because the model was fit on this same data, in-sample performance is expected to look favorable and should not be interpreted as a forward-looking forecast.

```{r}
#| label: train_metrics_table

# Count daily entry trades — days where a sector's signal flips from FLAT to LONG
n_trades_train <- train_signals_opt %>%
  group_by(symbol) %>%
  arrange(date) %>%
  mutate(entry = signal == "LONG" & dplyr::lag(signal, default = "FLAT") == "FLAT") %>%
  ungroup() %>%
  summarise(trades = sum(entry)) %>%
  pull(trades)

train_metrics_strategy <- compute_metrics(
  port_ret_vec = train_combined$port_ret_net,
  spy_ret_vec  = train_combined$spy_c2c,
  label        = "Strategy (Train)"
)
train_metrics_spy <- compute_metrics(
  port_ret_vec = train_combined$spy_c2c,
  label        = "SPY Buy & Hold (Train)"
)

bind_rows(train_metrics_strategy, train_metrics_spy) %>%
  mutate(
    `% In Market` = c(
      round(mean(portfolio_train$n_positions > 0) * 100, 1),
      100
    ),
    `# Trades` = c(n_trades_train, 0)
  ) %>%
  kable(caption = "Table 4: In-Sample Performance Metrics (Training Period — June 2000 – Dec 2024)",
        format = "html") %>%
  kable_styling(bootstrap_options = c("striped","hover"))
```

```{r}
#| label: train_cum_chart

train_cr <- train_combined %>%
  mutate(
    strategy = (cumprod(1 + port_ret_net) - 1) * 100,
    spy      = (cumprod(1 + spy_c2c)      - 1) * 100
  )

plot_ly(train_cr, x = ~date) %>%
  add_lines(y = ~strategy, name = "Strategy (net)",
            line = list(color = "#1565c0", width = 1.5)) %>%
  add_lines(y = ~spy,      name = "SPY Buy & Hold",
            line = list(color = "#b71c1c", width = 1.5)) %>%
  plotly::layout(
    title     = "Figure 3: Cumulative Returns — Training Period (In-Sample)",
    xaxis     = list(title = ""),
    yaxis     = list(title = "Cumulative Return (%)", ticksuffix = "%"),
    hovermode = "x unified",
    legend    = list(x = 0.01, y = 0.99)
  )
```

## Performance

The testing period — January 2025 through March 2026 — is the true out-of-sample evaluation. The Kalman filter state carries forward naturally from training because `dlmFilter` processes data sequentially; applying it to the full concatenated price series involves no look-ahead. The optimal parameters ($k$ = `r OPTIMAL_K`, $Q_2$ = `r OPTIMAL_Q`) are fixed from training. No refitting or parameter adjustment occurs. The test set was not examined at any point prior to this section.

```{r}
#| label: build_test_features
#| cache: true
#| include: false

# Apply Kalman to full price history (train + test concatenated).
# dlmFilter is sequential — filter state at t uses only y_1..y_t; no look-ahead.
full_sector_prices <- bind_rows(sector_train, sector_test) %>%
  arrange(symbol, date)

kalman_full_opt <- full_sector_prices %>%
  group_by(symbol) %>%
  arrange(date, .by_group = TRUE) %>%
  group_modify(~ apply_kalman(.x, q_vel = OPTIMAL_Q)) %>%
  ungroup()
```

```{r}
#| label: test_clustering_and_portfolio
#| include: false

# Generate daily signals for the test period using optimal parameters
# The Kalman state from training carries forward naturally via the full concatenation
kalman_test_only <- kalman_full_opt %>%
  filter(date >= TEST_START, date <= TEST_END)

test_signals <- generate_signals_clustered(
  kalman_test_only, yc_filled,
  k_clusters = OPTIMAL_K, yc_thresh = YC_THRESHOLD
)

# Returns for test period
returns_test <- kalman_full_opt %>%
  filter(symbol %in% SECTOR_SYMBOLS, date >= TEST_START, date <= TEST_END) %>%
  select(symbol, date, c2c_ret, o2c_ret, c2o_ret)

# Build test portfolio
portfolio_test <- build_portfolio(test_signals, returns_test, tc_rate = TC_RATE)

# SPY test benchmark returns
spy_ret_test <- spy_test %>%
  arrange(date) %>%
  mutate(spy_c2c = adjusted / dplyr::lag(adjusted) - 1) %>%
  filter(!is.na(spy_c2c)) %>%
  select(date, spy_c2c)

test_combined <- portfolio_test %>%
  left_join(spy_ret_test, by = "date") %>%
  filter(!is.na(spy_c2c))

cat("Test portfolio constructed.\n")
cat("Test period:", format(min(test_combined$date)), "to", format(max(test_combined$date)), "\n")
cat("Trading days:", nrow(test_combined), "\n")
cat("Avg daily positions:", round(mean(test_combined$n_positions), 2), "\n")
cat("% Days in market:", round(mean(test_combined$n_positions > 0) * 100, 1), "%\n")
```

```{r}
#| label: test_metrics_table

n_trades_test <- test_signals %>%
  group_by(symbol) %>%
  arrange(date) %>%
  mutate(entry = signal == "LONG" & dplyr::lag(signal, default = "FLAT") == "FLAT") %>%
  ungroup() %>%
  summarise(trades = sum(entry)) %>%
  pull(trades)

test_metrics_strategy <- compute_metrics(
  port_ret_vec = test_combined$port_ret_net,
  spy_ret_vec  = test_combined$spy_c2c,
  label        = "Strategy (Test)"
)
test_metrics_spy <- compute_metrics(
  port_ret_vec = test_combined$spy_c2c,
  label        = "SPY Buy & Hold (Test)"
)

bind_rows(test_metrics_strategy, test_metrics_spy) %>%
  mutate(
    `% In Market` = c(
      round(mean(portfolio_test$n_positions > 0) * 100, 1),
      100
    ),
    `# Trades` = c(n_trades_test, 0),
    `Gross CAGR` = c(
      round((prod(1 + na.omit(test_combined$port_ret_gross))^(252/nrow(test_combined)) - 1) * 100, 2),
      NA
    )
  ) %>%
  kable(caption = "Table 5: Out-of-Sample Performance Metrics (Testing Period — Jan 2025 – Mar 2026)",
        format = "html") %>%
  kable_styling(bootstrap_options = c("striped","hover")) %>%
  footnote(general = "Strategy return = Open-to-Close of execution day. SPY = Close-to-Close. TC = 0.05% one-way.")
```

```{r}
#| label: cum_return_chart

test_cr <- test_combined %>%
  mutate(
    net   = (cumprod(1 + port_ret_net)   - 1) * 100,
    gross = (cumprod(1 + port_ret_gross) - 1) * 100,
    spy   = (cumprod(1 + spy_c2c)        - 1) * 100
  )

plot_ly(test_cr, x = ~date) %>%
  add_lines(y = ~net,   name = "Strategy (net)",
            line = list(color = "#1565c0", width = 2)) %>%
  add_lines(y = ~gross, name = "Strategy (gross)",
            line = list(color = "#42a5f5", width = 1.5, dash = "dash")) %>%
  add_lines(y = ~spy,   name = "SPY Buy & Hold",
            line = list(color = "#b71c1c", width = 2)) %>%
  plotly::layout(
    title     = "Figure 4: Cumulative Returns — Testing Period (Out-of-Sample)",
    xaxis     = list(title = ""),
    yaxis     = list(title = "Cumulative Return (%)", ticksuffix = "%"),
    hovermode = "x unified",
    legend    = list(x = 0.01, y = 0.99)
  )
```

```{r}
#| label: pa_charts

# Build xts for PerformanceAnalytics functions that need it
port_xts <- xts(test_combined$port_ret_net, order.by = test_combined$date)
spy_xts  <- xts(test_combined$spy_c2c,      order.by = test_combined$date)
both_xts <- na.omit(merge(port_xts, spy_xts))
colnames(both_xts) <- c("Strategy", "SPY")

# ── Plotly performance dashboard: daily returns + drawdown ──────────────────
dd_df <- test_combined %>%
  mutate(
    strat_cum = cumprod(1 + port_ret_net),
    spy_cum   = cumprod(1 + spy_c2c),
    strat_dd  = (strat_cum - cummax(strat_cum)) / cummax(strat_cum) * 100,
    spy_dd    = (spy_cum   - cummax(spy_cum))   / cummax(spy_cum)   * 100,
    daily_pct = port_ret_net * 100,
    bar_color = if_else(daily_pct >= 0, "gain", "loss")
  )

p_ret <- plot_ly(dd_df, x = ~date) %>%
  add_bars(y = ~daily_pct, color = ~bar_color,
           colors = c("gain" = "#1b5e20", "loss" = "#b71c1c"),
           showlegend = FALSE,
           hovertemplate = "%{x}<br>Daily: %{y:.2f}%<extra></extra>") %>%
  plotly::layout(
    yaxis     = list(title = "Daily Return (%)", ticksuffix = "%"),
    xaxis     = list(title = "", showticklabels = FALSE),
    hovermode = "x unified",
    bargap    = 0
  )

p_dd <- plot_ly(dd_df, x = ~date) %>%
  add_lines(y = ~strat_dd, name = "Strategy",
            line = list(color = "#1565c0", width = 1.5)) %>%
  add_lines(y = ~spy_dd,   name = "SPY",
            line = list(color = "#b71c1c", width = 1.5)) %>%
  plotly::layout(
    yaxis     = list(title = "Drawdown (%)", ticksuffix = "%"),
    xaxis     = list(title = ""),
    hovermode = "x unified",
    legend    = list(x = 0.01, y = 0.05)
  )

subplot(p_ret, p_dd, nrows = 2, shareX = TRUE, titleY = TRUE,
        heights = c(0.45, 0.55)) %>%
  plotly::layout(
    title = "Figure 5: Performance Dashboard — Daily Returns & Drawdown (Testing Period)"
  )
```

```{r}
#| label: monthly_holdings

# Which sectors were held each month (test period)
monthly_holdings <- test_signals %>%
  filter(signal == "LONG") %>%
  mutate(ym = floor_date(date, "month")) %>%
  group_by(ym) %>%
  summarise(
    held_sectors = paste(sort(unique(symbol)), collapse = ", "),
    n_held       = n_distinct(symbol),
    .groups      = "drop"
  )

kable(monthly_holdings,
      col.names = c("Month","Sectors Held","# Sectors"),
      caption   = "Table 6: Monthly Holdings — Testing Period",
      format    = "html") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"))
```

```{r}
#| label: positions_chart

# Heatmap: which sectors were LONG on each day of the test period
pos_df <- test_signals %>%
  mutate(held = as.integer(signal == "LONG"))

pos_wide <- pos_df %>%
  select(date, symbol, held) %>%
  pivot_wider(names_from = symbol, values_from = held, values_fill = 0)

pos_mat  <- as.matrix(pos_wide[, -1])
rownames(pos_mat) <- as.character(pos_wide$date)

plot_ly(
  x          = colnames(pos_mat),
  y          = rownames(pos_mat),
  z          = t(pos_mat),
  type       = "heatmap",
  colorscale = list(c(0, "#f5f5f5"), c(1, "#1565c0")),
  showscale  = FALSE,
  hovertemplate = "<b>%{x}</b><br>%{y}<br>%{z:d} (1=LONG, 0=FLAT)<extra></extra>"
) %>%
  plotly::layout(
    title  = "Figure 6: Sector Positions Held — Testing Period",
    xaxis  = list(title = ""),
    yaxis  = list(title = "", autorange = "reversed",
                  tickfont = list(size = 9))
  )
```

```{r}
#| label: daily_pnl_chart

pnl_df <- test_combined %>%
  mutate(
    daily_pct = port_ret_net * 100,
    cum_pct   = (cumprod(1 + port_ret_net) - 1) * 100,
    bar_color = if_else(daily_pct >= 0, "gain", "loss")
  )

plot_ly(pnl_df, x = ~date) %>%
  add_bars(
    y             = ~daily_pct,
    color         = ~bar_color,
    colors        = c("gain" = "#1b5e20", "loss" = "#b71c1c"),
    yaxis         = "y",
    name          = "Daily Return",
    showlegend    = TRUE,
    hovertemplate = "%{x}<br>Daily: %{y:.2f}%<extra></extra>",
    bargap        = 0
  ) %>%
  add_lines(
    y             = ~cum_pct,
    yaxis         = "y2",
    name          = "Cumulative Return",
    line          = list(color = "#1565c0", width = 2),
    hovertemplate = "%{x}<br>Cumulative: %{y:.1f}%<extra></extra>"
  ) %>%
  plotly::layout(
    title     = "Figure 7: Daily P&L and Cumulative Return — Testing Period",
    xaxis     = list(title = ""),
    yaxis     = list(title = "Daily Return (%)", ticksuffix = "%", side = "left"),
    yaxis2    = list(title = "Cumulative Return (%)", ticksuffix = "%",
                     overlaying = "y", side = "right"),
    hovermode = "x unified",
    legend    = list(x = 0.01, y = 0.99),
    bargap    = 0
  )
```

I also examine rolling correlation to SPY, which measures the strategy's diversification value relative to a passive index. A strategy with low or negative correlation adds something beyond what SPY already provides.

```{r}
#| label: rolling_correlation

# 60-day rolling Pearson correlation between strategy and SPY returns
min_window <- 30  # require at least 30 observations for correlation

roll_cor_vec <- rollapply(
  cbind(test_combined$port_ret_net, test_combined$spy_c2c),
  width     = 60,
  FUN       = function(x) {
    cc <- stats::complete.cases(x)
    if (sum(cc) < min_window) return(NA_real_)
    stats::cor(x[cc, 1], x[cc, 2])
  },
  by.column = FALSE,
  align     = "right",
  fill      = NA
)

roll_cor_df <- tibble(
  date    = test_combined$date,
  cor_60d = as.numeric(roll_cor_vec)
)

plot_ly(
  data = roll_cor_df,
  x = ~date, y = ~cor_60d,
  type = "scatter", mode = "lines",
  line = list(color = "#1565c0", width = 2),
  name = "60-day Rolling Correlation"
) %>%
  add_segments(
    x = min(roll_cor_df$date, na.rm=TRUE), xend = max(roll_cor_df$date, na.rm=TRUE),
    y = 0, yend = 0,
    line = list(color = "grey60", dash = "dash"),
    showlegend = FALSE
  ) %>%
  plotly::layout(
    title  = "Figure 8: 60-Day Rolling Correlation — Strategy vs SPY (Testing Period)",
    xaxis  = list(title = ""),
    yaxis  = list(title = "Pearson Correlation", range = c(-1, 1)),
    hovermode = "x unified"
  )

cat("Average correlation to SPY (test period):",
    round(mean(roll_cor_df$cor_60d, na.rm=TRUE), 3), "\n")
```

To probe robustness, I run three stress tests on the results. First, doubling transaction costs to 0.10% one-way tests sensitivity to the TC assumption. Second and third, excluding the GFC (2008–2009) and COVID shock (February–June 2020) from the training backtest shows whether the strategy's training-period performance depends critically on those extreme episodes.

```{r}
#| label: stress_test_computations
#| include: false

# ── Stress Test 1: Double Transaction Costs (0.10% one-way) ──────────────────
portfolio_stress_2x_tc <- build_portfolio(test_signals, returns_test, tc_rate = TC_RATE * 2)
stress_combined_2x <- portfolio_stress_2x_tc %>%
  left_join(spy_ret_test, by = "date") %>% filter(!is.na(spy_c2c))

# ── Stress Test 2: Exclude GFC (2008–2009) from TRAINING period ───────────────
returns_no_gfc  <- returns_train_opt %>%
  filter(!(date >= "2008-01-01" & date <= "2009-12-31"))
signals_no_gfc  <- train_signals_opt %>%
  filter(!(date >= "2008-01-01" & date <= "2009-12-31"))
portfolio_no_gfc <- build_portfolio(signals_no_gfc, returns_no_gfc, tc_rate = TC_RATE)
spy_no_gfc <- spy_ret_train %>%
  filter(!(date >= "2008-01-01" & date <= "2009-12-31"))
combined_no_gfc <- portfolio_no_gfc %>%
  left_join(spy_no_gfc, by = "date") %>% filter(!is.na(spy_c2c))

# ── Stress Test 3: Exclude COVID Shock (Feb–Jun 2020) from TRAINING ───────────
returns_no_covid <- returns_train_opt %>%
  filter(!(date >= "2020-02-01" & date <= "2020-06-30"))
signals_no_covid <- train_signals_opt %>%
  filter(!(date >= "2020-02-01" & date <= "2020-06-30"))
portfolio_no_covid <- build_portfolio(signals_no_covid, returns_no_covid, tc_rate = TC_RATE)
spy_no_covid <- spy_ret_train %>%
  filter(!(date >= "2020-02-01" & date <= "2020-06-30"))
combined_no_covid <- portfolio_no_covid %>%
  left_join(spy_no_covid, by = "date") %>% filter(!is.na(spy_c2c))
```

```{r}
#| label: stress_metrics_table

stress_table <- bind_rows(
  # Training period baseline
  compute_metrics(train_combined$port_ret_net, train_combined$spy_c2c,
                  label = "Training (Baseline)"),
  # Test period baseline
  compute_metrics(test_combined$port_ret_net,  test_combined$spy_c2c,
                  label = "Testing (Baseline)"),
  # 2x TC (test period)
  compute_metrics(stress_combined_2x$port_ret_net, stress_combined_2x$spy_c2c,
                  label = "Testing (2× TC)"),
  # Exclude GFC (training period analysis)
  compute_metrics(combined_no_gfc$port_ret_net, combined_no_gfc$spy_c2c,
                  label = "Training (excl. GFC 2008-09)"),
  # Exclude COVID (training period analysis)
  compute_metrics(combined_no_covid$port_ret_net, combined_no_covid$spy_c2c,
                  label = "Training (excl. COVID 2020)")
)

kable(stress_table,
      digits  = 2,
      caption = "Table 7: Stress Test Comparison — Strategy Performance Across Scenarios",
      format  = "html") %>%
  kable_styling(bootstrap_options = c("striped","hover")) %>%
  row_spec(2, bold = TRUE, background = "#e3f2fd") %>%
  footnote(general = "Testing scenarios apply the frozen model trained on full training data. Training scenarios exclude specific periods from the training backtest.")
```

```{r}
#| label: stress_cum_chart

sc_std <- test_combined      %>% mutate(cum = (cumprod(1 + port_ret_net) - 1) * 100)
sc_2x  <- stress_combined_2x %>% mutate(cum = (cumprod(1 + port_ret_net) - 1) * 100)

plot_ly(x = ~sc_std$date) %>%
  add_lines(y = ~sc_std$cum, name = "Standard TC (0.05%)",
            line = list(color = "#1565c0", width = 2)) %>%
  add_lines(y = ~sc_2x$cum,  name = "Double TC (0.10%)",
            line = list(color = "#e65100", width = 2, dash = "dash")) %>%
  plotly::layout(
    title     = "Figure 9: Transaction Cost Sensitivity — Testing Period",
    xaxis     = list(title = ""),
    yaxis     = list(title = "Cumulative Return (%)", ticksuffix = "%"),
    hovermode = "x unified",
    legend    = list(x = 0.01, y = 0.99)
  )
```

---

# Learnings

## Summary

This project implemented a Kalman-filter-based sector rotation strategy using k-means momentum clustering with a yield curve macro overlay. The strategy was trained on 24.5 years of daily SPDR sector ETF data and evaluated out-of-sample over 15 months. The Kalman filter extracted a noise-filtered trend velocity from daily prices, and k-means partitioned the eleven sectors daily into economically interpretable momentum regimes. The yield curve overlay provided a macro-grounded defensive mechanism, exiting the market entirely during inversion periods rather than requiring a separate sector selection forecast under stress conditions.

The two-parameter optimization — $k$ and $Q_2$ jointly via Omega ratio on training data — produced a clean and interpretable optimal configuration. The Omega ratio objective avoided the trap of optimizing for cumulative return, which would have overfit to the 2009–2021 bull market at the expense of the defensive cash periods the yield curve filter is designed to produce.

## Limitations

The yield curve overlay uses one FRED series with a fixed threshold. Yield curve inversion is a strong but lagging recession predictor — the inversion-to-recession lag historically ranges from 6 to 24 months, meaning the defensive rotation may activate prematurely or persist too long relative to actual market drawdowns.

K-means produces hard cluster assignments, and sectors near cluster boundaries can flip between Bull and non-Bull on small velocity changes, generating unnecessary trades even with `nstart = 25` to stabilize centroid initialization. Soft clustering methods like Gaussian Mixture Models would provide probabilistic membership and could reduce this boundary sensitivity.

The parameters are optimized once on the full training period and applied statically to the test period. A walk-forward approach — re-optimizing quarterly on a rolling window — would be more realistic and reduce the risk of regime-change invalidation as market conditions evolve.

Held sectors are equally weighted regardless of velocity magnitude within the Bull cluster. Velocity-proportional weighting would concentrate exposure on the strongest-trending sectors within the cluster rather than distributing it uniformly.

The 15-month test window is short. A single bull or bear market regime during this period will dominate the results. Definitive conclusions about strategy robustness require multiple market cycles.

## Future Work

- **Walk-forward backtesting**: Re-optimize $k$ and $Q_2$ quarterly on rolling 36-month windows to adapt to regime shifts
- **Soft clustering**: Replace k-means with Gaussian Mixture Models; weight positions by Bull cluster membership probability
- **Multi-feature clustering**: Extend beyond single-feature velocity clustering by adding yield-spread sensitivity or realized volatility as additional dimensions
- **Additional macro filters**: Credit spreads (HYG/LQD ratio), ISM PMI, or Fed funds futures to complement the yield curve signal
- **Volatility scaling**: Scale position size inversely to realized volatility, as in Moskowitz, Ooi & Pedersen (2012), to stabilize portfolio risk through regimes
