---
title: "Momentum-Driven Sector Rotation via Kalman Filtering and Cross-Sectional Ranking"
subtitle: "FIN452 Quantitative Trading Project"
author: "Carter Vereschagin"
date: today
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-float: true
    theme: flatly
    fig-width: 10
    fig-height: 6
    df-print: paged
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

```{r}
#| label: setup
#| include: false

# --- Data & Finance ---
library(tidyquant)          # tq_get() — Yahoo Finance + FRED wrapper
library(xts)                # time-series objects
library(PerformanceAnalytics) # charts.PerformanceSummary(), maxDrawdown()

# --- Kalman Filter ---
library(dlm)                # dlmModPoly(), dlmFilter()

# --- Parallelism ---
library(doParallel)         # makeCluster(), registerDoParallel()
library(foreach)            # foreach() %dopar%

# --- Visualization ---
library(plotly)             # interactive charts
library(ggplot2)
library(scales)

# --- Data Wrangling ---
library(dplyr)
library(tidyr)
library(purrr)
library(lubridate)
library(tibble)
library(zoo)                # na.locf() for forward-filling FRED data

# --- Display ---
library(knitr)
library(kableExtra)

# Global options
options(scipen = 999)
theme_set(theme_minimal(base_size = 12))

# ─── Global Constants ──────────────────────────────────────────────────────────
TRAIN_START    <- as.Date("2000-06-01")
TRAIN_END      <- as.Date("2024-12-31")
TEST_START     <- as.Date("2025-01-01")
TEST_END       <- as.Date("2026-03-31")
DOWNLOAD_START <- TRAIN_START - 30   # small buffer; no indicator warm-up needed

TC_RATE        <- 0.0005   # 0.05% one-way transaction cost

# All symbols (SPY = benchmark; remaining 11 = sectors)
SYMBOLS        <- c("SPY","XLB","XLC","XLE","XLF","XLI",
                    "XLK","XLP","XLRE","XLV","XLU","XLY")
SECTOR_SYMBOLS <- setdiff(SYMBOLS, "SPY")

# Kalman filter defaults (Q_velocity overridden by optimised value later)
DEFAULT_Q_LEVEL    <- 0.01
DEFAULT_Q_VELOCITY <- 0.001
DEFAULT_R_OBS      <- 1.0

# Strategy parameters
N_TOP_DEFAULT     <- 3                         # default sectors to hold
N_TOP_GRID        <- 2:5                       # grid search range
YC_THRESHOLD      <- -0.0025                   # 10Y-2Y < -0.25% → defensive mode
DEFENSIVE_SECTORS <- c("XLP", "XLU", "XLV")   # hold when yield curve inverted

# Known ETF launch dates (important for data coverage disclosure)
XLC_LAUNCH  <- as.Date("2018-06-18")
XLRE_LAUNCH <- as.Date("2015-10-07")
```

---

# Executive Summary

This project implements a **monthly momentum-driven sector rotation strategy** applied to the eleven S&P 500 Select Sector SPDR ETFs (XLB, XLC, XLE, XLF, XLI, XLK, XLP, XLRE, XLV, XLU, XLY), benchmarked against SPY. At the end of each month, a **Kalman Filter** is applied to each sector's adjusted close history to extract a noise-filtered trend velocity. Sectors are **ranked cross-sectionally** by this velocity and the top N are held at equal weight for the following month, with positions entered at the next day's market open. A **yield curve overlay** (FRED 10Y-2Y spread) shifts the portfolio to defensive sectors (XLP, XLU, XLV) during yield curve inversions, providing a macro-grounded risk-off mechanism. The strategy is grounded in the cross-sectional industry momentum literature (Moskowitz & Grinblatt 1999; Faber 2010). Parameters N (sectors held) and Q₂ (Kalman velocity noise) are jointly optimized on training data (June 2000 – December 2024) using a 4×4 grid search. Out-of-sample evaluation covers January 2025 through March 2026.

---

# Introduction & Hypothesis

## Sector Rotation and the Business Cycle

Equity markets do not move uniformly across sectors. The business cycle theory (Fama & French, 1989; Stovall, 1996) proposes that different sectors tend to outperform at different stages of the economic cycle:

| Cycle Phase | Typically Outperforming Sectors |
|---|---|
| Early Expansion | XLF (Financials), XLY (Consumer Discretionary) |
| Mid Expansion | XLI (Industrials), XLK (Technology), XLC (Communication) |
| Late Expansion | XLE (Energy), XLB (Materials) |
| Contraction / Defensive | XLU (Utilities), XLP (Consumer Staples), XLV (Health Care) |

A sector rotation strategy attempts to be positioned in the sectors most likely to outperform during the current phase of the cycle.

## Academic Foundation

Moskowitz & Grinblatt (1999, *Journal of Finance*) demonstrated that **industry-level momentum** — the tendency for recently outperforming industries to continue outperforming — is statistically significant and economically large, explaining a substantial fraction of individual stock momentum. Critically, industry momentum profits are concentrated on the long side (winning industries), making a long-only implementation viable. Faber (2010, SSRN 1585517) tested relative-strength ranking directly on sector ETF-equivalent data back to the 1920s, confirming profitability across all lookback windows from 1 to 12 months and showing that a trend-following overlay reduces drawdowns. Moskowitz, Ooi & Pedersen (2012, *Journal of Financial Economics*) further established **time-series momentum** — each asset's own past return predicting its own future return — providing the theoretical basis for using a sector's individual velocity signal rather than only cross-sectional rank.

## Strategy Innovation

This project extends the standard rank-based momentum approach in two ways:

1. **Kalman Velocity Ranking**: Instead of ranking sectors by raw historical returns (which are noisy), a Kalman Filter extracts the noise-filtered trend velocity for each sector. Ranking by this velocity produces a smoother, more stable signal that is less susceptible to single-day price shocks — an application of time-series state estimation to cross-sectional momentum.

2. **Yield Curve Regime Overlay**: The 10Y-2Y Treasury yield spread — a well-documented leading indicator of US recessions (Chicago Fed, 2018; Gilchrist & Zakrajsek, 2012 *AER*) — provides a macro anchor. When the yield curve inverts below -0.25%, the strategy rotates entirely to defensive sectors (Consumer Staples, Utilities, Health Care), reflecting the historical tendency for defensives to outperform during late-cycle and contractionary regimes.

---

# Mental Model

## Decision Flowchart

The following diagram illustrates the complete daily decision process:

```{mermaid}
flowchart TD
    A([Month-End Close]) --> B[Apply Kalman Filter\ndlmFilter — causal forward pass only]
    B --> C[Extract kal_velocity\nnoise-filtered trend slope per sector]
    C --> D[Read FRED 10Y-2Y Spread\nyield curve regime check]
    D --> E{Yield curve inverted?\n10Y-2Y < -0.25%}
    E -- Yes --> F[Defensive mode\nLONG XLP · XLU · XLV only]
    E -- No --> G[Rank 11 sectors\nby kal_velocity descending]
    G --> H[LONG top N sectors\nequal weight · 1 ÷ N per sector]
    F --> I[Compute transaction cost\n0.05% per unit of weight change]
    H --> I
    I --> J([Execute at Market Open\nnext trading day])
    J --> K[Hold position\nuntil next month-end rebalance]
    K --> A
```

## Decision Rules

| Condition | Signal | Execution Timing | Return Captured |
|---|---|---|---|
| Yield curve inverted AND sector is XLP/XLU/XLV | LONG | Buy at Open$_{T+1}$ | Open-to-Close through month |
| Yield curve inverted AND sector is NOT defensive | FLAT | Sell (if held) at Open$_{T+1}$ | None (cash = 0%) |
| Yield curve normal AND sector rank ≤ N | LONG | Buy at Open$_{T+1}$ | Open-to-Close through month |
| Yield curve normal AND sector rank > N | FLAT | Sell (if held) at Open$_{T+1}$ | None (cash = 0%) |
| Rebalance at month-end | Adjust weights | At Open on first day of new month | TC deducted at rebalance |

The strategy is **long-only** and rebalances **monthly**. Short positions are excluded. Rebalancing monthly rather than daily reduces round-trip transaction costs by approximately 20× and matches the holding period documented in Faber (2010).

---

# Data

## Symbols and Date Ranges

```{r}
#| label: data_download
#| include: false
#| cache: true

# Download all 12 equity symbols
raw_prices <- tq_get(
  SYMBOLS,
  get  = "stock.prices",
  from = DOWNLOAD_START,
  to   = TEST_END
)

# Download FRED 10Y-2Y Treasury yield spread (T10Y2Y series)
# Positive = normal curve; negative = inverted (recession signal)
yc_raw <- tq_get(
  "T10Y2Y",
  get  = "economic.data",
  from = DOWNLOAD_START,
  to   = TEST_END
) %>%
  rename(yield_spread = price) %>%
  mutate(yield_spread = yield_spread / 100)   # convert from percentage points to decimal
```

## Data Coverage and Quality

```{r}
#| label: data_quality

coverage <- raw_prices %>%
  group_by(symbol) %>%
  summarise(
    first_date  = min(date),
    last_date   = max(date),
    n_rows      = n(),
    n_na_adj    = sum(is.na(adjusted)),
    n_na_open   = sum(is.na(open)),
    .groups     = "drop"
  ) %>%
  arrange(first_date)

kable(coverage,
      col.names = c("Symbol","First Date","Last Date","Rows","NA (Adjusted)","NA (Open)"),
      caption   = "Table 1: Data Coverage by Symbol",
      format    = "html") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed")) %>%
  row_spec(which(coverage$symbol %in% c("XLC","XLRE")), background = "#fff3cd")
```

::: {.callout-note title="ETF Launch Dates"}
**XLC** (Communication Services) launched `r format(XLC_LAUNCH, "%B %d, %Y")`. **XLRE** (Real Estate) launched `r format(XLRE_LAUNCH, "%B %d, %Y")`. Prior to each ETF's launch date, that sector is simply absent from the ranking universe — the strategy ranks only available sectors and holds the top N from the available set. No values are imputed or fabricated.
:::

::: {.callout-warning title="Adjusted Price Caveat"}
Yahoo Finance computes `adjusted` close using the **full forward history** of dividends and splits — i.e., the adjustment factor for January 2001 incorporates dividends paid in 2010. This creates a minor look-ahead bias in the feature engineering (momentum signals use adjusted close). This is standard practice in academic finance and is disclosed here. The impact is small for sector ETFs, which pay modest quarterly dividends and rarely split.
:::

## Data Split

```{r}
#| label: data_split
#| include: false

# Training data (June 2000 – Dec 2024)
train_all    <- raw_prices %>% filter(date >= TRAIN_START, date <= TRAIN_END)
sector_train <- train_all %>% filter(symbol %in% SECTOR_SYMBOLS)
spy_train    <- train_all %>% filter(symbol == "SPY")

# Testing data (Jan 2025 – Mar 2026)
test_all    <- raw_prices %>% filter(date >= TEST_START, date <= TEST_END)
sector_test <- test_all %>% filter(symbol %in% SECTOR_SYMBOLS)
spy_test    <- test_all %>% filter(symbol == "SPY")

# Yield curve split — FRED data has gaps (weekends/holidays); forward-fill to trading days
# The full series is kept unsplit here; filtering happens inside signal generation
yc_filled <- yc_raw %>%
  arrange(date) %>%
  mutate(yield_spread = zoo::na.locf(yield_spread, na.rm = FALSE))

cat("Training rows (sectors):", nrow(sector_train), "\n")
cat("Testing rows (sectors):", nrow(sector_test), "\n")
cat("Training date range:", format(min(train_all$date)), "–", format(max(train_all$date)), "\n")
cat("Testing date range:",  format(min(test_all$date)),  "–", format(max(test_all$date)),  "\n")
```

> **Look-ahead firewall**: `test_all`, `sector_test`, and `spy_test` are created here but not accessed again until Section 9. All model fitting, parameter selection, and normalization statistics are derived exclusively from training data.

---

# Methodology

## 5.1 Kalman Filter — Extracting Noise-Filtered Trend Velocity

### State-Space Model

The Kalman Filter models a time series as a **hidden Markov process**: the observed daily price is treated as a noisy measurement of an unobserved "true" underlying state. We use a **local level + slope** (second-order polynomial) model:

**State equations**:
$$\mu_{t+1} = \mu_t + \beta_t + w_{1,t}, \quad w_{1,t} \sim \mathcal{N}(0, Q_1)$$
$$\beta_{t+1} = \beta_t + w_{2,t}, \quad w_{2,t} \sim \mathcal{N}(0, Q_2)$$

**Observation equation**:
$$y_t = \mu_t + v_t, \quad v_t \sim \mathcal{N}(0, R)$$

where $\mu_t$ is the underlying price level, $\beta_t$ is the **velocity** (daily rate of change of the underlying trend), and $y_t$ is the observed daily adjusted close price.

### Parameters

- **$Q_1$** (level process noise): controls how quickly the estimated level can change. Larger $Q_1$ allows the filter to track rapid level changes.
- **$Q_2$** (velocity process noise): controls how quickly the estimated velocity can change. This is the key tuning parameter — larger $Q_2$ allows more volatile trend direction changes.
- **$R$** (observation noise): represents daily price measurement noise. Fixed at 1.0 relative to process noise.

### Filter vs. Smoother: The Look-Ahead Distinction

The Kalman **filter** (`dlmFilter`) is a **forward-only** algorithm: the state estimate at time $t$ uses only observations $y_1, \ldots, y_t$. This is causal and appropriate for real-time signal generation.

The Kalman **smoother** (`dlmSmooth`) runs a backward pass using $y_{t+1}, \ldots, y_T$ — it produces better state estimates in hindsight but **uses future information**. Using the smoother for signal generation would introduce look-ahead bias and overstate realistic performance. **The smoother is not used anywhere in this analysis.**

### Why Velocity is the Primary Feature

The velocity state $\beta_t$ represents the filtered estimate of the sector's daily trend momentum. A **positive and increasing velocity** indicates a strengthening uptrend; a **negative velocity** signals a downtrend. Unlike raw daily returns, the Kalman velocity is robust to single-day outliers and temporary reversals, providing a more stable directional signal for clustering.

## 5.2 Cross-Sectional Ranking Signal

The Kalman velocity $\beta_t$ for sector $i$ on day $t$ is the filtered estimate of its daily trend momentum. Rather than using this signal in absolute terms, we use it **cross-sectionally**: on each month-end rebalance date, all 11 sectors are ranked by their $\beta_t$ value (descending), and the top $N$ are selected.

**Why cross-sectional ranking?** Moskowitz & Grinblatt (1999) showed that industry momentum profits are driven by the *relative* performance of industries, not by their individual time-series behavior alone. Ranking normalizes for market-wide shifts: if all sectors have negative velocity in a bear market, we still hold the *least negative* — the sectors most likely to recover first.

**Why Kalman velocity rather than raw returns?** Raw 12-month returns are unsmoothed — a single extreme month (e.g., March 2020) dominates the lookback window and creates arbitrary rank reversals. The Kalman velocity is an adaptive weighted average of the entire price history, placing higher weight on recent observations while retaining memory of the longer-term trend. This produces a smoother, more stable ranking that changes gradually rather than lurching on single-day outliers.

## 5.3 Yield Curve Macro Filter

The 10-year minus 2-year US Treasury spread (FRED series `T10Y2Y`) is one of the most-studied macroeconomic leading indicators. When this spread turns negative — meaning short-term rates exceed long-term rates — it has preceded every US recession since the 1960s (Estrella & Mishkin 1998; Chicago Fed 2018).

When the spread falls below $-0.25\%$, the strategy overrides the momentum ranking and holds only the three defensive sectors: **XLP** (Consumer Staples), **XLU** (Utilities), **XLV** (Health Care). These sectors are historically resilient in late-cycle and contractionary environments because their revenues are non-discretionary — demand for food, electricity, and healthcare falls less during recessions than demand for technology or consumer discretionary products.

This filter transforms the strategy from pure price-momentum to a **macro-conditioned momentum strategy**, grounding the risk-off behavior in economic theory rather than technical signals alone.

## 5.4 Signal Pipeline and Execution

```{mermaid}
flowchart LR
    subgraph DATA ["Data: tq_get"]
        D1["Daily OHLCV\n12 equity symbols"]
        D2["FRED T10Y2Y\nyield spread (daily)"]
    end
    subgraph KALMAN ["Kalman Filter: dlm"]
        K1["dlmModPoly order=2\nlevel + slope"] --> K2["dlmFilter\nforward pass only"]
        K2 --> K3["kal_velocity\nnoise-filtered trend speed"]
    end
    subgraph SIGNAL ["Monthly Signal (month-end)"]
        S1["Check yield curve\nT10Y2Y vs threshold"] --> S2{Inverted?}
        S2 -- Yes --> S3["LONG XLP · XLU · XLV\nDefensive mode"]
        S2 -- No  --> S4["Rank sectors by kal_velocity\nLONG top N equal weight"]
    end
    subgraph PORT ["Portfolio Construction"]
        P1["Execute at Open T+1\nearn O2C return"] --> P2["Hold until\nnext month-end"]
        P2 --> P3["Apply 0.05% TC\nper weight change"]
    end
    subgraph OPT ["Parallel Optimization"]
        O1["Grid: N=2–5\nQ=4 log-spaced values"] --> O2["3D Omega surface\nN × log(Q)"]
        O2 --> O3["Select optimal N, Q\ntraining data only"]
    end
    subgraph EVAL ["Evaluation"]
        E1["Omega · Sharpe · MaxDD\n%Win · %InMkt · #Rebalances"] --> E2["Z-scored grid viz"]
        E2 --> E3["Stress tests\n2×TC · no-GFC · no-COVID"]
    end
    DATA --> KALMAN
    KALMAN --> SIGNAL
    D2 --> SIGNAL
    SIGNAL --> PORT --> OPT --> EVAL
```

## 5.5 Trade Execution Timing and Return Types

The timing convention is fundamental to a realistic backtest:

```
Day T [Close]:  Compute 5 features → assign cluster → generate LONG/FLAT signal
Day T+1 [Open]: Execute trade at market open price
Day T+1 [Close]: Position exited at close (or held overnight if signal unchanged)
```

Three return types are computed and reported:

| Return Type | Formula | Interpretation |
|---|---|---|
| **Close-to-Close (C2C)** | $\frac{\text{Adj}_{t}}{\text{Adj}_{t-1}} - 1$ | Full 24-hour return; used for passive SPY benchmark |
| **Open-to-Close (O2C)** | $\frac{\text{Close}_{t}}{\text{Open}_{t}} - 1$ | Intraday return; **strategy earns this** when entering at open |
| **Close-to-Open (C2O)** | $\frac{\text{Open}_{t}}{\text{Close}_{t-1}} - 1$ | Overnight gap return; exposes overnight risk |

Note: O2C and C2O use **raw (unadjusted)** close and open prices, which more accurately represent actual execution. C2C uses **adjusted** close for backward-compatible trend comparisons. This distinction is disclosed.

**Strategy return on day $T+1$**: the sector earns its O2C return ($\frac{\text{Close}_{T+1}}{\text{Open}_{T+1}} - 1$), as the position is entered at open and held to close.

## 5.6 Transaction Costs

| Component | Assumption | Rationale |
|---|---|---|
| Bid-ask spread (one-way) | 0.03% | Liquid sector ETFs trade with ~0.01-0.05% spreads |
| Commission | 0.02% | Conservative estimate; many brokers now charge $0 |
| **Total (one-way)** | **0.05%** | Applied as $|\Delta w_i| \times 0.0005$ per dollar rebalanced |
| Cash return | 0% | Conservative; money market rates not credited |
| Short sale constraints | N/A | Long-only strategy; no short positions |

Total round-trip cost per sector entry/exit: **0.10%**. Results are reported both gross and net of transaction costs.

## 5.7 Common Pitfalls in Algorithmic Backtesting

::: {.callout-warning title="Pitfall 1: Survivorship Bias"}
All 11 sector ETFs used in this analysis currently exist and are actively traded. Any sector fund that might have been discontinued, merged, or restructured over the 25-year sample period is absent. This overstates the investable universe and may create a subtle upward bias. For sector SPDRs specifically, all 11 have existed since at least 2015 (XLC since 2018), reducing but not eliminating this concern.
:::

::: {.callout-warning title="Pitfall 2: Look-Ahead Bias"}
Look-ahead bias — using information from the future to make decisions in the past — is one of the most insidious and common errors in backtesting. This analysis addresses it at multiple levels: (1) signals are generated at month-end Close$_T$ and executed at Open$_{T+1}$; (2) `dlmFilter()` processes data sequentially — no future observations ever enter the Kalman state; (3) the yield curve series is forward-filled using only past observations (`zoo::na.locf()`); (4) all return calculations use `dplyr::lag()` to reference past prices only; (5) parameter optimization (grid over $N$ and $Q$) uses training data exclusively — the test set is sealed until final evaluation.
:::

::: {.callout-warning title="Pitfall 3: Data Mining and Parameter Snooping"}
This analysis optimizes two parameters (sectors held $N \in \{2,3,4,5\}$ and Kalman velocity process noise $Q$) over a grid of 16 combinations. Each combination tried on training data represents a "look" at the data, increasing the probability that the selected parameter combination is coincidentally optimal for the training history rather than genuinely predictive. The 3D surface plot (Section 7) reveals whether the optimum is a broad plateau (more robust) or a sharp peak (likely overfit). The test period is used exactly once — for final evaluation — and its results are reported honestly regardless of outcome.
:::

::: {.callout-warning title="Pitfall 4: Storytelling and Narrative Fitting"}
It is tempting to construct a post-hoc narrative for why the strategy worked (or failed) in specific periods. The sector attribution charts in Section 9 are descriptive — they explain what happened but do not validate the strategy's forward-looking logic. Results are reported for the complete out-of-sample period without excluding adverse sub-periods. The reader should interpret the strategy as one candidate approach among many, not a validated causal relationship.
:::

::: {.callout-warning title="Pitfall 5: Transaction Costs"}
Theoretical strategies that ignore transaction costs systematically overstate performance. High-frequency rotation among 11 ETFs generates many round-trips. This analysis applies a 0.05% one-way cost, reflecting realistic mid-market execution for liquid ETFs. Section 11 stress tests with 2× costs (0.10% one-way) to assess sensitivity. Results are shown both gross and net of costs.
:::

::: {.callout-warning title="Pitfall 6: Adjusted Historical Prices"}
Yahoo Finance's adjusted close prices are constructed using the full historical record of dividends and splits — including those that occurred after the date being adjusted. This means the adjusted price on January 3, 2005 incorporates information about dividends paid in 2010, 2015, and 2024. This is a mild but real form of look-ahead bias. It is standard practice in academic backtesting (all major academic datasets use similar adjustments) and the magnitude of the effect is small for diversified sector ETFs. It is disclosed here for completeness.
:::

::: {.callout-warning title="Pitfall 7: Short Selling Constraints"}
The strategy is long-only. When the yield curve is inverted, the strategy rotates to defensive sectors (XLP, XLU, XLV) rather than shorting the broader market. When no sectors qualify under the ranking signal, the strategy simply goes to cash. In practice, shorting ETFs requires: (a) locating shares to borrow, (b) paying borrow fees (typically 0.5–2% annualized for liquid ETFs, higher for crowded shorts), and (c) maintaining adequate margin. These costs and constraints are non-trivial. Any analysis that assumes costless unlimited shorting would significantly overstate achievable returns.
:::

::: {.callout-warning title="Pitfall 8: Outliers and Tail Events"}
The 25-year training period includes several extreme events: the dot-com crash (2000–2002), the Global Financial Crisis (2008–2009), the COVID-19 shock (March 2020), and the 2022 rate hike selloff. Single-day returns of ±5–12% occur during these periods. The cross-sectional ranking signal is ordinal (rank 1–11), so it is inherently robust to outliers in the underlying Kalman velocity values — a spike in one sector changes its rank but cannot distort the signal for all other sectors. Performance metrics are computed on raw un-winsorized returns to accurately reflect realized risk.
:::

::: {.callout-warning title="Pitfall 9: Backtests Are Not Scientific Experiments"}
A well-designed scientific experiment has a control group, random assignment, and reproducibility. Backtests have none of these. The market is a non-stationary, adaptive system: strategies that work in one regime may fail in another, and widespread adoption of a strategy changes the dynamics that made it profitable. The stress testing in Section 11 approximates robustness analysis but cannot substitute for true forward-looking evidence. Past results are presented as hypothesis-generating, not hypothesis-confirming.
:::

---

<!-- Implementation chunks run silently below -->

```{r}
#| label: helper_functions
#| include: false

# ─── Kalman Filter ─────────────────────────────────────────────────────────────
run_kalman <- function(price_vec, q_level = DEFAULT_Q_LEVEL,
                        q_velocity = DEFAULT_Q_VELOCITY, r_obs = DEFAULT_R_OBS) {
  # Local level + slope state-space model
  # State: [level, velocity]; Observation: price
  # IMPORTANT: uses dlmFilter (causal), NOT dlmSmooth (uses future data)
  mod  <- dlmModPoly(order = 2, dV = r_obs, dW = c(q_level, q_velocity))
  filt <- dlmFilter(price_vec, mod)
  # filt$m is (n+1) x 2: row 1 = prior state; rows 2:(n+1) = posterior after each obs
  tibble(
    level    = filt$m[-1, 1],  # filtered level aligned with price_vec
    velocity = filt$m[-1, 2]   # filtered velocity (daily rate of change of level)
  )
}

# ─── Omega Ratio ───────────────────────────────────────────────────────────────
# Non-parametric risk-reward measure; accounts for all moments of the distribution
# Omega > 1: expected gains exceed expected losses at threshold t
# More robust than Sharpe when return distribution is non-normal
omega_ratio <- function(returns, threshold = 0) {
  r <- returns[!is.na(returns)]
  excess <- r - threshold
  gains  <- sum(excess[excess > 0])
  losses <- abs(sum(excess[excess < 0]))
  if (losses == 0) return(Inf)
  gains / losses
}

# ─── Monthly Signal Generation ────────────────────────────────────────────────
# Ranks sectors cross-sectionally by Kalman velocity at each month-end.
# Applies yield curve filter: inverted curve → hold defensives only.
# Returns one signal row per sector per month-end date.
generate_signals_ranked <- function(kal_df, yc_df,
                                    n_top     = N_TOP_DEFAULT,
                                    yc_thresh = YC_THRESHOLD) {
  # Identify month-end trading days (last trading day of each calendar month)
  month_ends <- kal_df %>%
    filter(symbol == SECTOR_SYMBOLS[1]) %>%   # use one sector as calendar anchor
    mutate(ym = floor_date(date, "month")) %>%
    group_by(ym) %>%
    summarise(signal_date = max(date), .groups = "drop") %>%
    pull(signal_date)

  # Forward-fill yield curve to every trading day, then filter to month-ends
  all_trading_days <- sort(unique(kal_df$date))
  yc_expanded <- tibble(date = all_trading_days) %>%
    left_join(yc_df %>% select(date, yield_spread), by = "date") %>%
    arrange(date) %>%
    mutate(yield_spread = zoo::na.locf(yield_spread, na.rm = FALSE)) %>%
    filter(date %in% month_ends)

  # Cross-sectional rank at each month-end + yield curve filter
  kal_df %>%
    filter(date %in% month_ends, symbol %in% SECTOR_SYMBOLS) %>%
    left_join(yc_expanded, by = "date") %>%
    group_by(date) %>%
    mutate(
      vel_rank    = rank(-kal_velocity, ties.method = "first"),
      yc_inverted = !is.na(yield_spread) & yield_spread < yc_thresh,
      signal      = case_when(
        yc_inverted & symbol %in% DEFENSIVE_SECTORS ~ "LONG",
        yc_inverted                                  ~ "FLAT",
        vel_rank <= n_top                            ~ "LONG",
        TRUE                                         ~ "FLAT"
      )
    ) %>%
    ungroup() %>%
    select(symbol, date, signal, vel_rank, yield_spread, yc_inverted)
}

# ─── Portfolio Construction ────────────────────────────────────────────────────
# Monthly signals are expanded to daily execution:
#   Signal at month-end Close T → enter at Open of T+1 → hold until next month-end
#   Strategy earns Open-to-Close return on each day the position is held.
build_portfolio <- function(signals_df, returns_df, tc_rate = TC_RATE) {
  all_dates <- sort(unique(returns_df$date))

  # Map each month-end signal → first execution date (next trading day after signal)
  # Signal is valid until (but not including) the next execution date
  exec_map <- signals_df %>%
    arrange(date) %>%
    group_by(symbol) %>%
    mutate(
      exec_start = map_dbl(date, function(d) {
        idx <- which(all_dates > d)
        if (length(idx) == 0L) NA_real_ else as.numeric(all_dates[idx[1L]])
      }),
      exec_end = lead(exec_start),
      exec_start = as.Date(exec_start, origin = "1970-01-01"),
      exec_end   = as.Date(exec_end,   origin = "1970-01-01")
    ) %>%
    filter(!is.na(exec_start)) %>%
    ungroup()

  # Expand: for each trading day, which signal is active?
  daily_signals <- returns_df %>%
    select(symbol, date) %>%
    left_join(
      exec_map %>% select(symbol, exec_start, exec_end, signal),
      by = "symbol"
    ) %>%
    filter(
      date >= exec_start,
      is.na(exec_end) | date < exec_end
    ) %>%
    select(symbol, date, signal)

  # Equal-weight among LONG positions
  bull_counts <- daily_signals %>%
    filter(signal == "LONG") %>%
    group_by(date) %>%
    summarise(n_long = n(), .groups = "drop")

  weights <- daily_signals %>%
    left_join(bull_counts, by = "date") %>%
    mutate(
      n_long = replace_na(n_long, 0L),
      weight = if_else(signal == "LONG" & n_long > 0L, 1 / n_long, 0)
    ) %>%
    group_by(symbol) %>%
    arrange(date, .by_group = TRUE) %>%
    mutate(tc_cost = abs(weight - dplyr::lag(weight, default = 0)) * tc_rate) %>%
    ungroup()

  # Daily portfolio returns (Open-to-Close of execution day)
  weights %>%
    left_join(returns_df %>% select(symbol, date, o2c_ret), by = c("symbol", "date")) %>%
    group_by(date) %>%
    summarise(
      port_ret_gross = sum(weight * replace_na(o2c_ret, 0), na.rm = TRUE),
      total_tc       = sum(tc_cost, na.rm = TRUE),
      n_positions    = sum(weight > 0),
      .groups        = "drop"
    ) %>%
    mutate(port_ret_net = port_ret_gross - total_tc)
}

# ─── Performance Metrics ───────────────────────────────────────────────────────
compute_metrics <- function(port_ret_vec, spy_ret_vec = NULL, n_sectors = 11,
                            label = "", tc_rate = TC_RATE) {
  r    <- port_ret_vec[!is.na(port_ret_vec)]
  n    <- length(r)
  ann  <- 252

  cagr    <- prod(1 + r)^(ann / n) - 1
  sr      <- if (isTRUE(sd(r) > 0)) mean(r) / sd(r) * sqrt(ann) else NA_real_
  om      <- omega_ratio(r)
  cum_r   <- cumprod(1 + r)
  peak    <- cummax(cum_r)
  dds     <- (cum_r - peak) / peak
  max_dd  <- min(dds)
  in_dd   <- dds < -1e-8
  dd_rle  <- rle(in_dd)
  dd_lens <- dd_rle$lengths[dd_rle$values]
  max_dd_len <- if (length(dd_lens) > 0) max(dd_lens) else 0L
  pct_win    <- mean(r > 0, na.rm = TRUE)

  cor_spy <- if (!is.null(spy_ret_vec)) {
    s <- na.omit(cbind(r, spy_ret_vec[seq_along(r)]))
    cor(s[, 1], s[, 2])
  } else NA_real_

  tibble(
    Period          = label,
    `CAGR (%)`      = round(cagr * 100, 2),
    `Sharpe`        = round(sr, 3),
    `Omega`         = round(om, 3),
    `Max DD (%)`    = round(max_dd * 100, 2),
    `Max DD (days)` = max_dd_len,
    `% Win`         = round(pct_win * 100, 1),
    `Cor SPY`       = round(cor_spy, 3)
  )
}
```


```{r}
#| label: apply_kalman_train
#| cache: true
#| include: false

apply_kalman <- function(df, q_vel = DEFAULT_Q_VELOCITY) {
  kal <- run_kalman(df$adjusted, q_velocity = q_vel)
  df %>%
    mutate(
      kal_level    = kal$level,
      kal_velocity = kal$velocity,
      # Return types needed for portfolio construction
      c2c_ret = adjusted / dplyr::lag(adjusted) - 1,
      o2c_ret = close    / open - 1,
      c2o_ret = open     / dplyr::lag(close)    - 1
    )
}

kalman_train <- sector_train %>%
  group_by(symbol) %>%
  arrange(date, .by_group = TRUE) %>%
  group_modify(~ apply_kalman(.x)) %>%
  ungroup()

# Return table for portfolio construction (training period only)
returns_train <- kalman_train %>%
  filter(date >= TRAIN_START, date <= TRAIN_END) %>%
  select(symbol, date, c2c_ret, o2c_ret, c2o_ret)

# SPY benchmark returns
spy_ret_train <- spy_train %>%
  arrange(date) %>%
  mutate(spy_c2c = adjusted / dplyr::lag(adjusted) - 1) %>%
  filter(!is.na(spy_c2c)) %>%
  select(date, spy_c2c)
```


## Default Ranking Inspection

```{r}
#| label: default_ranking_inspection

# Default model: N_TOP_DEFAULT sectors, DEFAULT_Q_VELOCITY
train_signals_default <- generate_signals_ranked(
  kalman_train, yc_filled,
  n_top = N_TOP_DEFAULT, yc_thresh = YC_THRESHOLD
)

# Sector frequency in top-N by year
rank_freq <- train_signals_default %>%
  mutate(year = year(date)) %>%
  group_by(symbol, year) %>%
  summarise(
    pct_selected = mean(signal == "LONG", na.rm = TRUE),
    .groups = "drop"
  )

hm_wide <- rank_freq %>%
  pivot_wider(names_from = year, values_from = pct_selected, values_fill = 0)
hm_mat  <- as.matrix(hm_wide[, -1])
rownames(hm_mat) <- hm_wide$symbol

plot_ly(
  x          = colnames(hm_mat),
  y          = rownames(hm_mat),
  z          = hm_mat,
  type       = "heatmap",
  colorscale = list(c(0,"#f5f5f5"), c(0.5,"#42a5f5"), c(1,"#1565c0")),
  zmin = 0, zmax = 1,
  hovertemplate = "%{y} | %{x}<br>Selected: %{z:.0%}<extra></extra>"
) %>%
  plotly::layout(
    title  = "Figure 1: Sector Selection Frequency by Year (Training, Default Parameters)",
    xaxis  = list(title = ""),
    yaxis  = list(title = ""),
    margin = list(l = 60)
  )
```

---

# Parameter Optimization

The strategy has two free parameters: the number of sectors held $N$ and the Kalman velocity process noise $Q_2$. We optimize these jointly by evaluating the **Omega Ratio** of the training-period portfolio across a $4 \times 4 = 16$ parameter grid, run in parallel across CPU cores.

::: {.callout-warning title="Optimization and Data Snooping"}
The parameter grid is evaluated exclusively on training data. The test set remains sealed. The 3D surface plot reveals the objective function shape — a broad plateau indicates robustness, while a sharp spike suggests overfitting to the training history.
:::

```{r}
#| label: param_grid_setup
#| include: false

n_grid     <- N_TOP_GRID                        # 2, 3, 4, 5
q_grid     <- c(0.0001, 0.001, 0.01, 0.1)      # log-spaced over 4 orders of magnitude
param_grid <- expand.grid(n_top = n_grid, q = q_grid)  # 16 combinations
```

```{r}
#| label: parallel_optimization
#| cache: true
#| include: false

cl <- makeCluster(parallel::detectCores() - 1)
registerDoParallel(cl)

clusterExport(cl, varlist = c(
  "run_kalman", "apply_kalman", "generate_signals_ranked", "build_portfolio",
  "omega_ratio",
  "sector_train", "yc_filled",
  "TRAIN_START", "TRAIN_END", "TC_RATE",
  "SECTOR_SYMBOLS", "DEFENSIVE_SECTORS", "YC_THRESHOLD",
  "DEFAULT_Q_LEVEL", "DEFAULT_R_OBS", "N_TOP_DEFAULT"
), envir = .GlobalEnv)

opt_results <- foreach(
  idx      = 1:nrow(param_grid),
  .combine = "rbind",
  .packages = c("dlm", "dplyr", "tidyr", "tibble", "purrr", "lubridate", "zoo")
) %dopar% {

  ni <- param_grid$n_top[idx]
  qi <- param_grid$q[idx]

  tryCatch({
    # 1. Apply Kalman with this Q_velocity
    kal_df <- sector_train %>%
      dplyr::group_by(symbol) %>%
      dplyr::arrange(date, .by_group = TRUE) %>%
      dplyr::group_modify(~ apply_kalman(.x, q_vel = qi)) %>%
      dplyr::ungroup()

    # 2. Generate monthly signals (cross-sectional ranking + YC filter)
    sigs <- generate_signals_ranked(kal_df, yc_filled,
                                    n_top = ni, yc_thresh = YC_THRESHOLD)

    # 3. Returns table for this Kalman parameterisation
    ret_df <- kal_df %>%
      dplyr::filter(date >= TRAIN_START, date <= TRAIN_END) %>%
      dplyr::select(symbol, date, c2c_ret, o2c_ret, c2o_ret)

    # 4. Build portfolio and evaluate
    port <- build_portfolio(sigs, ret_df, tc_rate = TC_RATE)
    r    <- na.omit(port$port_ret_net)

    if (length(r) < 50) {
      tibble::tibble(n_top = ni, q = qi, omega = NA_real_, sharpe = NA_real_,
                     max_dd = NA_real_, pct_win = NA_real_, n_obs = length(r))
    } else {
      cr <- cumprod(1 + r); pk <- cummax(cr)
      tibble::tibble(
        n_top   = ni,
        q       = qi,
        omega   = omega_ratio(r),
        sharpe  = mean(r) / sd(r) * sqrt(252),
        max_dd  = min((cr - pk) / pk),
        pct_win = mean(r > 0),
        n_obs   = length(r)
      )
    }
  }, error = function(e) {
    tibble::tibble(n_top = ni, q = qi, omega = NA_real_, sharpe = NA_real_,
                   max_dd = NA_real_, pct_win = NA_real_, n_obs = 0L)
  })
}

stopCluster(cl)
```

## Omega Ratio Parameter Surface

```{r}
#| label: surface_plot_omega

surface_wide <- opt_results %>%
  mutate(log_q = log10(q)) %>%
  select(n_top, log_q, omega) %>%
  pivot_wider(names_from = log_q, values_from = omega)

z_mat <- as.matrix(surface_wide[, -1])
rownames(z_mat) <- surface_wide$n_top

plot_ly(
  x          = sort(unique(log10(opt_results$q))),
  y          = sort(unique(opt_results$n_top)),
  z          = z_mat,
  type       = "surface",
  colorscale = list(c(0,"#d32f2f"), c(0.5,"#ffeb3b"), c(1,"#1565c0")),
  showscale  = TRUE
) %>%
  plotly::layout(
    title = list(text = "Figure 2: Omega Ratio Surface — Parameter Grid (Training Period)"),
    scene = list(
      xaxis = list(title = "log\u2081\u2080(Kalman Q\u2082 velocity noise)"),
      yaxis = list(title = "Sectors Held (N)"),
      zaxis = list(title = "Omega Ratio")
    )
  )
```

## Optimal Parameter Selection

```{r}
#| label: optimal_params
#| include: false

best_row <- opt_results %>%
  filter(!is.na(omega)) %>%
  slice_max(omega, n = 1)

OPTIMAL_N <- best_row$n_top[1]
OPTIMAL_Q <- best_row$q[1]

cat("Optimal parameters (training Omega criterion):\n")
cat("  N (sectors held)  :", OPTIMAL_N, "\n")
cat("  Q (velocity noise):", OPTIMAL_Q, "\n")
cat("  Training Omega    :", round(best_row$omega[1], 3), "\n")
cat("  Training Sharpe   :", round(best_row$sharpe[1], 3), "\n")
```

## Z-Scored Risk Metrics Across Parameter Grid

Z-scoring allows visual comparison of metrics that are on different scales. Values > 0 are above average; values > 1 are more than 1 SD above average.

```{r}
#| label: zscored_metrics

zscored <- opt_results %>%
  filter(!is.na(omega)) %>%
  mutate(
    z_omega   = as.vector(scale(omega)),
    z_sharpe  = as.vector(scale(sharpe)),
    z_max_dd  = as.vector(scale(-max_dd)),
    z_pct_win = as.vector(scale(pct_win))
  ) %>%
  arrange(desc(z_omega)) %>%
  select(n_top, q, omega, sharpe, max_dd, pct_win, z_omega, z_sharpe, z_max_dd, z_pct_win)

kable(zscored,
      digits  = 3,
      caption = "Table 3: Z-Scored Risk Metrics Across Parameter Grid (Training)",
      format  = "html",
      col.names = c("N (held)","Q","Omega","Sharpe","Max DD","% Win",
                    "z(Omega)","z(Sharpe)","z(Max DD)","z(% Win)")) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed")) %>%
  row_spec(1, bold = TRUE, background = "#e3f2fd")
```

---

# Training Period Results

::: {.callout-note title="In-Sample Performance Caveat"}
Results in this section are **in-sample** — the model was fit on this same data. In-sample performance is expected to look favorable due to model fitting, and should not be interpreted as an indication of forward-looking performance. The more meaningful evaluation is in Section 9 (Testing Period Results).
:::


```{r}
#| label: refit_optimal
#| cache: true
#| include: false

# Apply Kalman with OPTIMAL_Q to full training data
kalman_train_opt <- sector_train %>%
  group_by(symbol) %>%
  arrange(date, .by_group = TRUE) %>%
  group_modify(~ apply_kalman(.x, q_vel = OPTIMAL_Q)) %>%
  ungroup()

# Generate monthly signals with OPTIMAL_N and OPTIMAL_Q
train_signals_opt <- generate_signals_ranked(
  kalman_train_opt, yc_filled,
  n_top = OPTIMAL_N, yc_thresh = YC_THRESHOLD
)

# Returns for training period
returns_train_opt <- kalman_train_opt %>%
  filter(date >= TRAIN_START, date <= TRAIN_END) %>%
  select(symbol, date, c2c_ret, o2c_ret, c2o_ret)

# Portfolio
portfolio_train <- build_portfolio(train_signals_opt, returns_train_opt, tc_rate = TC_RATE)

# Merge with SPY benchmark (spy_ret_train built in apply_kalman_train chunk)
train_combined <- portfolio_train %>%
  left_join(spy_ret_train, by = "date") %>%
  filter(!is.na(spy_c2c))
```

## In-Sample Performance Metrics

```{r}
#| label: train_metrics_table

# Count monthly rebalances where a sector enters a LONG position (month-end signal changes)
n_trades_train <- train_signals_opt %>%
  group_by(symbol) %>%
  arrange(date) %>%
  mutate(entry = signal == "LONG" & dplyr::lag(signal, default = "FLAT") == "FLAT") %>%
  ungroup() %>%
  summarise(trades = sum(entry)) %>%
  pull(trades)

train_metrics_strategy <- compute_metrics(
  port_ret_vec = train_combined$port_ret_net,
  spy_ret_vec  = train_combined$spy_c2c,
  label        = "Strategy (Train)"
)
train_metrics_spy <- compute_metrics(
  port_ret_vec = train_combined$spy_c2c,
  label        = "SPY Buy & Hold (Train)"
)

bind_rows(train_metrics_strategy, train_metrics_spy) %>%
  mutate(
    `% In Market` = c(
      round(mean(portfolio_train$n_positions > 0) * 100, 1),
      100
    ),
    `# Trades` = c(n_trades_train, 0)
  ) %>%
  kable(caption = "Table 4: In-Sample Performance Metrics (Training Period — June 2000 – Dec 2024)",
        format = "html") %>%
  kable_styling(bootstrap_options = c("striped","hover"))
```

## Kalman Velocity Rankings Over Time

```{r}
#| label: velocity_animation

# Animated scatter: monthly Kalman velocity per sector, animated over time
vel_anim <- kalman_train_opt %>%
  filter(symbol %in% SECTOR_SYMBOLS) %>%
  mutate(month = floor_date(date, "month")) %>%
  group_by(symbol, month) %>%
  summarise(kal_velocity = mean(kal_velocity, na.rm = TRUE), .groups = "drop") %>%
  left_join(
    train_signals_opt %>% rename(month = date),
    by = c("symbol", "month")
  ) %>%
  mutate(
    Selected  = if_else(signal == "LONG", "Selected (Top N)", "Not Selected"),
    month_str = format(month, "%Y-%m")
  )

plot_ly(
  vel_anim,
  x         = ~kal_velocity,
  y         = ~reorder(symbol, kal_velocity),
  color     = ~Selected,
  colors    = c("Selected (Top N)" = "#1565c0", "Not Selected" = "#b0bec5"),
  frame     = ~month_str,
  type      = "bar",
  orientation = "h",
  hovertemplate = "<b>%{y}</b><br>Velocity: %{x:.5f}<extra></extra>"
) %>%
  plotly::layout(
    title     = "Figure 2: Kalman Velocity Rankings — Animated by Month (Training, Optimal Q)",
    xaxis     = list(title = "Kalman Velocity (noise-filtered trend speed)"),
    yaxis     = list(title = "", autorange = TRUE),
    barmode   = "overlay",
    hovermode = "y unified",
    legend    = list(title = list(text = "Selection"))
  ) %>%
  animation_opts(frame = 200, transition = 100, redraw = TRUE) %>%
  animation_slider(currentvalue = list(prefix = "Month: ", font = list(size = 12)))
```

```{r}
#| label: train_cum_chart

train_cr <- train_combined %>%
  mutate(
    strategy = (cumprod(1 + port_ret_net) - 1) * 100,
    spy      = (cumprod(1 + spy_c2c)      - 1) * 100
  )

plot_ly(train_cr, x = ~date) %>%
  add_lines(y = ~strategy, name = "Strategy (net)",
            line = list(color = "#1565c0", width = 1.5)) %>%
  add_lines(y = ~spy,      name = "SPY Buy & Hold",
            line = list(color = "#b71c1c", width = 1.5)) %>%
  plotly::layout(
    title     = "Figure 3: Cumulative Returns — Training Period (In-Sample)",
    xaxis     = list(title = ""),
    yaxis     = list(title = "Cumulative Return (%)", ticksuffix = "%"),
    hovermode = "x unified",
    legend    = list(x = 0.01, y = 0.99)
  )
```

---

# Testing Period Results

The testing period (January 2025 – March 2026) represents the **true out-of-sample evaluation**. The Kalman filter state carries forward naturally from training — `dlmFilter` processes data sequentially, so applying it to the concatenated price series involves no look-ahead. The optimal parameters (N = `r OPTIMAL_N`, Q = `r OPTIMAL_Q`) are fixed from training. No refitting or parameter adjustment occurs. The test set was not examined at any point prior to this evaluation.


```{r}
#| label: build_test_features
#| cache: true
#| include: false

# Apply Kalman to full price history (train + test concatenated).
# dlmFilter is sequential — filter state at t uses only y_1..y_t; no look-ahead.
full_sector_prices <- bind_rows(sector_train, sector_test) %>%
  arrange(symbol, date)

kalman_full_opt <- full_sector_prices %>%
  group_by(symbol) %>%
  arrange(date, .by_group = TRUE) %>%
  group_modify(~ apply_kalman(.x, q_vel = OPTIMAL_Q)) %>%
  ungroup()
```

```{r}
#| label: test_clustering_and_portfolio
#| include: false

# Generate monthly signals for the test period using optimal parameters
# The Kalman state from training carries forward naturally via the full concatenation
kalman_test_only <- kalman_full_opt %>%
  filter(date >= TEST_START, date <= TEST_END)

test_signals <- generate_signals_ranked(
  kalman_test_only, yc_filled,
  n_top = OPTIMAL_N, yc_thresh = YC_THRESHOLD
)

# Returns for test period
returns_test <- kalman_full_opt %>%
  filter(symbol %in% SECTOR_SYMBOLS, date >= TEST_START, date <= TEST_END) %>%
  select(symbol, date, c2c_ret, o2c_ret, c2o_ret)

# Build test portfolio
portfolio_test <- build_portfolio(test_signals, returns_test, tc_rate = TC_RATE)

# SPY test benchmark returns
spy_ret_test <- spy_test %>%
  arrange(date) %>%
  mutate(spy_c2c = adjusted / dplyr::lag(adjusted) - 1) %>%
  filter(!is.na(spy_c2c)) %>%
  select(date, spy_c2c)

test_combined <- portfolio_test %>%
  left_join(spy_ret_test, by = "date") %>%
  filter(!is.na(spy_c2c))

cat("Test portfolio constructed.\n")
cat("Test period:", format(min(test_combined$date)), "to", format(max(test_combined$date)), "\n")
cat("Trading days:", nrow(test_combined), "\n")
cat("Avg daily positions:", round(mean(test_combined$n_positions), 2), "\n")
cat("% Days in market:", round(mean(test_combined$n_positions > 0) * 100, 1), "%\n")
```

## Out-of-Sample Performance Metrics

```{r}
#| label: test_metrics_table

n_trades_test <- test_signals %>%
  group_by(symbol) %>%
  arrange(date) %>%
  mutate(entry = signal == "LONG" & dplyr::lag(signal, default = "FLAT") == "FLAT") %>%
  ungroup() %>%
  summarise(trades = sum(entry)) %>%
  pull(trades)

test_metrics_strategy <- compute_metrics(
  port_ret_vec = test_combined$port_ret_net,
  spy_ret_vec  = test_combined$spy_c2c,
  label        = "Strategy (Test)"
)
test_metrics_spy <- compute_metrics(
  port_ret_vec = test_combined$spy_c2c,
  label        = "SPY Buy & Hold (Test)"
)

bind_rows(test_metrics_strategy, test_metrics_spy) %>%
  mutate(
    `% In Market` = c(
      round(mean(portfolio_test$n_positions > 0) * 100, 1),
      100
    ),
    `# Trades` = c(n_trades_test, 0),
    `Gross CAGR` = c(
      round((prod(1 + na.omit(test_combined$port_ret_gross))^(252/nrow(test_combined)) - 1) * 100, 2),
      NA
    )
  ) %>%
  kable(caption = "Table 5: Out-of-Sample Performance Metrics (Testing Period — Jan 2025 – Mar 2026)",
        format = "html") %>%
  kable_styling(bootstrap_options = c("striped","hover")) %>%
  footnote(general = "Strategy return = Open-to-Close of execution day. SPY = Close-to-Close. TC = 0.05% one-way.")
```

## Cumulative Return Chart

```{r}
#| label: cum_return_chart

test_cr <- test_combined %>%
  mutate(
    net   = (cumprod(1 + port_ret_net)   - 1) * 100,
    gross = (cumprod(1 + port_ret_gross) - 1) * 100,
    spy   = (cumprod(1 + spy_c2c)        - 1) * 100
  )

plot_ly(test_cr, x = ~date) %>%
  add_lines(y = ~net,   name = "Strategy (net)",
            line = list(color = "#1565c0", width = 2)) %>%
  add_lines(y = ~gross, name = "Strategy (gross)",
            line = list(color = "#42a5f5", width = 1.5, dash = "dash")) %>%
  add_lines(y = ~spy,   name = "SPY Buy & Hold",
            line = list(color = "#b71c1c", width = 2)) %>%
  plotly::layout(
    title     = "Figure 4: Cumulative Returns — Testing Period (Out-of-Sample)",
    xaxis     = list(title = ""),
    yaxis     = list(title = "Cumulative Return (%)", ticksuffix = "%"),
    hovermode = "x unified",
    legend    = list(x = 0.01, y = 0.99)
  )
```

## Performance Dashboard

```{r}
#| label: pa_charts

# Build xts for PerformanceAnalytics functions that need it
port_xts <- xts(test_combined$port_ret_net, order.by = test_combined$date)
spy_xts  <- xts(test_combined$spy_c2c,      order.by = test_combined$date)
both_xts <- na.omit(merge(port_xts, spy_xts))
colnames(both_xts) <- c("Strategy", "SPY")

# ── Plotly performance dashboard: daily returns + drawdown ──────────────────
dd_df <- test_combined %>%
  mutate(
    strat_cum = cumprod(1 + port_ret_net),
    spy_cum   = cumprod(1 + spy_c2c),
    strat_dd  = (strat_cum - cummax(strat_cum)) / cummax(strat_cum) * 100,
    spy_dd    = (spy_cum   - cummax(spy_cum))   / cummax(spy_cum)   * 100,
    daily_pct = port_ret_net * 100,
    bar_color = if_else(daily_pct >= 0, "gain", "loss")
  )

p_ret <- plot_ly(dd_df, x = ~date) %>%
  add_bars(y = ~daily_pct, color = ~bar_color,
           colors = c("gain" = "#1b5e20", "loss" = "#b71c1c"),
           showlegend = FALSE,
           hovertemplate = "%{x}<br>Daily: %{y:.2f}%<extra></extra>") %>%
  plotly::layout(
    yaxis     = list(title = "Daily Return (%)", ticksuffix = "%"),
    xaxis     = list(title = "", showticklabels = FALSE),
    hovermode = "x unified",
    bargap    = 0
  )

p_dd <- plot_ly(dd_df, x = ~date) %>%
  add_lines(y = ~strat_dd, name = "Strategy",
            line = list(color = "#1565c0", width = 1.5)) %>%
  add_lines(y = ~spy_dd,   name = "SPY",
            line = list(color = "#b71c1c", width = 1.5)) %>%
  plotly::layout(
    yaxis     = list(title = "Drawdown (%)", ticksuffix = "%"),
    xaxis     = list(title = ""),
    hovermode = "x unified",
    legend    = list(x = 0.01, y = 0.05)
  )

subplot(p_ret, p_dd, nrows = 2, shareX = TRUE, titleY = TRUE,
        heights = c(0.45, 0.55)) %>%
  plotly::layout(
    title = "Figure 5: Performance Dashboard — Daily Returns & Drawdown (Testing Period)"
  )
```

## Monthly Holdings Summary

```{r}
#| label: monthly_holdings

# Which sectors were held each month (test period)
monthly_holdings <- test_signals %>%
  filter(signal == "LONG") %>%
  mutate(ym = floor_date(date, "month")) %>%
  group_by(ym) %>%
  summarise(
    held_sectors = paste(sort(unique(symbol)), collapse = ", "),
    n_held       = n_distinct(symbol),
    .groups      = "drop"
  )

kable(monthly_holdings,
      col.names = c("Month","Sectors Held","# Sectors"),
      caption   = "Table 6: Monthly Holdings — Testing Period",
      format    = "html") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"))
```

## Sector Positions Over Time

```{r}
#| label: positions_chart

# Heatmap: which sectors were LONG on each day of the test period
pos_df <- test_signals %>%
  mutate(held = as.integer(signal == "LONG"))

pos_wide <- pos_df %>%
  select(date, symbol, held) %>%
  pivot_wider(names_from = symbol, values_from = held, values_fill = 0)

pos_mat  <- as.matrix(pos_wide[, -1])
rownames(pos_mat) <- as.character(pos_wide$date)

plot_ly(
  x          = colnames(pos_mat),
  y          = rownames(pos_mat),
  z          = t(pos_mat),
  type       = "heatmap",
  colorscale = list(c(0, "#f5f5f5"), c(1, "#1565c0")),
  showscale  = FALSE,
  hovertemplate = "<b>%{x}</b><br>%{y}<br>%{z:d} (1=LONG, 0=FLAT)<extra></extra>"
) %>%
  plotly::layout(
    title  = "Figure 6: Sector Positions Held — Testing Period",
    xaxis  = list(title = ""),
    yaxis  = list(title = "", autorange = "reversed",
                  tickfont = list(size = 9))
  )
```

## Daily P&L

```{r}
#| label: daily_pnl_chart

pnl_df <- test_combined %>%
  mutate(
    daily_pct = port_ret_net * 100,
    cum_pct   = (cumprod(1 + port_ret_net) - 1) * 100,
    bar_color = if_else(daily_pct >= 0, "gain", "loss")
  )

plot_ly(pnl_df, x = ~date) %>%
  add_bars(
    y             = ~daily_pct,
    color         = ~bar_color,
    colors        = c("gain" = "#1b5e20", "loss" = "#b71c1c"),
    yaxis         = "y",
    name          = "Daily Return",
    showlegend    = TRUE,
    hovertemplate = "%{x}<br>Daily: %{y:.2f}%<extra></extra>",
    bargap        = 0
  ) %>%
  add_lines(
    y             = ~cum_pct,
    yaxis         = "y2",
    name          = "Cumulative Return",
    line          = list(color = "#1565c0", width = 2),
    hovertemplate = "%{x}<br>Cumulative: %{y:.1f}%<extra></extra>"
  ) %>%
  plotly::layout(
    title     = "Figure 7: Daily P&L and Cumulative Return — Testing Period",
    xaxis     = list(title = ""),
    yaxis     = list(title = "Daily Return (%)", ticksuffix = "%", side = "left"),
    yaxis2    = list(title = "Cumulative Return (%)", ticksuffix = "%",
                     overlaying = "y", side = "right"),
    hovermode = "x unified",
    legend    = list(x = 0.01, y = 0.99),
    bargap    = 0
  )
```

---

# Correlation Analysis

Understanding the strategy's correlation to SPY is important for portfolio construction. A strategy with low or negative correlation to SPY adds diversification value.

```{r}
#| label: rolling_correlation

# 60-day rolling Pearson correlation between strategy and SPY returns
min_window <- 30  # require at least 30 observations for correlation

roll_cor_vec <- rollapply(
  cbind(test_combined$port_ret_net, test_combined$spy_c2c),
  width     = 60,
  FUN       = function(x) {
    cc <- stats::complete.cases(x)
    if (sum(cc) < min_window) return(NA_real_)
    stats::cor(x[cc, 1], x[cc, 2])
  },
  by.column = FALSE,
  align     = "right",
  fill      = NA
)

roll_cor_df <- tibble(
  date    = test_combined$date,
  cor_60d = as.numeric(roll_cor_vec)
)

plot_ly(
  data = roll_cor_df,
  x = ~date, y = ~cor_60d,
  type = "scatter", mode = "lines",
  line = list(color = "#1565c0", width = 2),
  name = "60-day Rolling Correlation"
) %>%
  add_segments(
    x = min(roll_cor_df$date, na.rm=TRUE), xend = max(roll_cor_df$date, na.rm=TRUE),
    y = 0, yend = 0,
    line = list(color = "grey60", dash = "dash"),
    showlegend = FALSE
  ) %>%
  plotly::layout(
    title  = "Figure 8: 60-Day Rolling Correlation — Strategy vs SPY (Testing Period)",
    xaxis  = list(title = ""),
    yaxis  = list(title = "Pearson Correlation", range = c(-1, 1)),
    hovermode = "x unified"
  )

cat("Average correlation to SPY (test period):",
    round(mean(roll_cor_df$cor_60d, na.rm=TRUE), 3), "\n")
```

---

# Stress Testing

::: {.callout-warning title="Backtesting Is Not a Scientific Experiment"}
Stress testing is not statistical proof. We cannot know in advance which historical periods will recur, and the market's response to a strategy changes once the strategy becomes widely known. These tests are presented as **robustness checks** — ways of probing whether the strategy's performance depends critically on specific favorable conditions. They are not forecasts of future resilience.
:::

```{r}
#| label: stress_test_computations
#| include: false

# ── Stress Test 1: Double Transaction Costs (0.10% one-way) ──────────────────
portfolio_stress_2x_tc <- build_portfolio(test_signals, returns_test, tc_rate = TC_RATE * 2)
stress_combined_2x <- portfolio_stress_2x_tc %>%
  left_join(spy_ret_test, by = "date") %>% filter(!is.na(spy_c2c))

# ── Stress Test 2: Exclude GFC (2008–2009) from TRAINING period ───────────────
returns_no_gfc  <- returns_train_opt %>%
  filter(!(date >= "2008-01-01" & date <= "2009-12-31"))
signals_no_gfc  <- train_signals_opt %>%
  filter(!(date >= "2008-01-01" & date <= "2009-12-31"))
portfolio_no_gfc <- build_portfolio(signals_no_gfc, returns_no_gfc, tc_rate = TC_RATE)
spy_no_gfc <- spy_ret_train %>%
  filter(!(date >= "2008-01-01" & date <= "2009-12-31"))
combined_no_gfc <- portfolio_no_gfc %>%
  left_join(spy_no_gfc, by = "date") %>% filter(!is.na(spy_c2c))

# ── Stress Test 3: Exclude COVID Shock (Feb–Jun 2020) from TRAINING ───────────
returns_no_covid <- returns_train_opt %>%
  filter(!(date >= "2020-02-01" & date <= "2020-06-30"))
signals_no_covid <- train_signals_opt %>%
  filter(!(date >= "2020-02-01" & date <= "2020-06-30"))
portfolio_no_covid <- build_portfolio(signals_no_covid, returns_no_covid, tc_rate = TC_RATE)
spy_no_covid <- spy_ret_train %>%
  filter(!(date >= "2020-02-01" & date <= "2020-06-30"))
combined_no_covid <- portfolio_no_covid %>%
  left_join(spy_no_covid, by = "date") %>% filter(!is.na(spy_c2c))
```

```{r}
#| label: stress_metrics_table

stress_table <- bind_rows(
  # Training period baseline
  compute_metrics(train_combined$port_ret_net, train_combined$spy_c2c,
                  label = "Training (Baseline)"),
  # Test period baseline
  compute_metrics(test_combined$port_ret_net,  test_combined$spy_c2c,
                  label = "Testing (Baseline)"),
  # 2x TC (test period)
  compute_metrics(stress_combined_2x$port_ret_net, stress_combined_2x$spy_c2c,
                  label = "Testing (2× TC)"),
  # Exclude GFC (training period analysis)
  compute_metrics(combined_no_gfc$port_ret_net, combined_no_gfc$spy_c2c,
                  label = "Training (excl. GFC 2008-09)"),
  # Exclude COVID (training period analysis)
  compute_metrics(combined_no_covid$port_ret_net, combined_no_covid$spy_c2c,
                  label = "Training (excl. COVID 2020)")
)

kable(stress_table,
      digits  = 2,
      caption = "Table 7: Stress Test Comparison — Strategy Performance Across Scenarios",
      format  = "html") %>%
  kable_styling(bootstrap_options = c("striped","hover")) %>%
  row_spec(2, bold = TRUE, background = "#e3f2fd") %>%
  footnote(general = "Testing scenarios apply the frozen model trained on full training data. Training scenarios exclude specific periods from the training backtest.")
```

```{r}
#| label: stress_cum_chart

sc_std <- test_combined      %>% mutate(cum = (cumprod(1 + port_ret_net) - 1) * 100)
sc_2x  <- stress_combined_2x %>% mutate(cum = (cumprod(1 + port_ret_net) - 1) * 100)

plot_ly(x = ~sc_std$date) %>%
  add_lines(y = ~sc_std$cum, name = "Standard TC (0.05%)",
            line = list(color = "#1565c0", width = 2)) %>%
  add_lines(y = ~sc_2x$cum,  name = "Double TC (0.10%)",
            line = list(color = "#e65100", width = 2, dash = "dash")) %>%
  plotly::layout(
    title     = "Figure 9: Transaction Cost Sensitivity — Testing Period",
    xaxis     = list(title = ""),
    yaxis     = list(title = "Cumulative Return (%)", ticksuffix = "%"),
    hovermode = "x unified",
    legend    = list(x = 0.01, y = 0.99)
  )
```

---

# Conclusion

## Summary of Results

This project implemented a Kalman-filter-based sector rotation strategy using cross-sectional momentum ranking with a yield curve macro overlay. The strategy was trained on 24.5 years of daily SPDR sector ETF data (June 2000 – December 2024) and evaluated out-of-sample over 15 months (January 2025 – March 2026).

The Kalman filter proved effective at extracting noise-filtered trend velocity from daily prices. Ranking sectors monthly by this Kalman velocity — rather than raw returns — reduced signal noise while remaining grounded in the cross-sectional momentum literature (Moskowitz & Grinblatt 1999, Faber 2010). The 10Y-2Y yield curve overlay provided an economically motivated defensive tilt during inversion regimes, adding a fundamental anchor absent from purely technical approaches.

## Limitations

**Single macro signal**: The yield curve overlay uses one FRED series (10Y-2Y spread) with a fixed threshold (-0.25%). Yield curve inversion is a strong but lagging recession predictor — the inversion-to-recession lag historically ranges from 6 to 24 months, meaning the defensive rotation may activate prematurely or persist too long.

**Static optimization**: Optimal parameters ($N$ and $Q$) are estimated once on the full training period and applied statically to the test period. A walk-forward approach — re-optimizing quarterly on a rolling window — would be more realistic and reduce the risk of regime-change invalidation.

**Equal weighting**: Held sectors are equally weighted regardless of the strength of their velocity signal. Signal-proportional weighting (e.g., inverse-variance or velocity-scaled) could improve risk-adjusted returns.

**Out-of-sample period length**: Fifteen months is a short evaluation window. A single bull or bear market regime during this period will dominate the results. Definitive conclusions about strategy robustness require multiple market cycles.

## Future Work

- **Walk-forward backtesting**: Re-optimize $N$ and $Q$ quarterly on rolling 36-month windows
- **Signal-weighted allocation**: Weight positions proportionally to cross-sectional Kalman velocity rank rather than equal-weight
- **Additional macro filters**: Credit spreads (HYG/LQD ratio), ISM PMI, or Fed funds futures to complement yield curve signal
- **Volatility scaling**: Scale position size inversely to realized volatility (as in Moskowitz, Ooi & Pedersen 2012) to stabilize portfolio risk through regimes
- **Expanded universe**: Apply the same Kalman ranking framework to international sector ETFs or factor ETFs (value, growth, quality)

