---
title: "Momentum-Driven Sector Rotation via Kalman Filtering and Cross-Sectional Ranking"
subtitle: "FIN452 Quantitative Trading Project"
author: "Carter Vereschagin"
date: today
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-float: true
    theme: flatly
    fig-width: 10
    fig-height: 6
    df-print: paged
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

```{r}
#| label: setup
#| include: false

# --- Data & Finance ---
library(tidyquant)          # tq_get() — Yahoo Finance + FRED wrapper
library(xts)                # time-series objects
library(PerformanceAnalytics) # charts.PerformanceSummary(), maxDrawdown()

# --- Kalman Filter ---
library(dlm)                # dlmModPoly(), dlmFilter()

# --- Parallelism ---
library(doParallel)         # makeCluster(), registerDoParallel()
library(foreach)            # foreach() %dopar%

# --- Visualization ---
library(plotly)             # interactive charts
library(ggplot2)
library(scales)

# --- Data Wrangling ---
library(dplyr)
library(tidyr)
library(purrr)
library(lubridate)
library(tibble)
library(zoo)                # na.locf() for forward-filling FRED data

# --- Display ---
library(knitr)
library(kableExtra)

# Global options
options(scipen = 999)
theme_set(theme_minimal(base_size = 12))

# ─── Global Constants ──────────────────────────────────────────────────────────
TRAIN_START    <- as.Date("2000-06-01")
TRAIN_END      <- as.Date("2024-12-31")
TEST_START     <- as.Date("2025-01-01")
TEST_END       <- as.Date("2026-03-31")
DOWNLOAD_START <- TRAIN_START - 30   # small buffer; no indicator warm-up needed

TC_RATE        <- 0.0005   # 0.05% one-way transaction cost

# All symbols (SPY = benchmark; remaining 11 = sectors)
SYMBOLS        <- c("SPY","XLB","XLC","XLE","XLF","XLI",
                    "XLK","XLP","XLRE","XLV","XLU","XLY")
SECTOR_SYMBOLS <- setdiff(SYMBOLS, "SPY")

# Kalman filter defaults (Q_velocity overridden by optimised value later)
DEFAULT_Q_LEVEL    <- 0.01
DEFAULT_Q_VELOCITY <- 0.001
DEFAULT_R_OBS      <- 1.0

# Strategy parameters
DEFAULT_K         <- 3                         # default number of k-means clusters
K_GRID            <- 2:5                       # grid search range for k
YC_THRESHOLD      <- -0.0025                   # 10Y-2Y < -0.25% → go to cash (all FLAT)

# Known ETF launch dates (important for data coverage disclosure)
XLC_LAUNCH  <- as.Date("2018-06-18")
XLRE_LAUNCH <- as.Date("2015-10-07")
```

---

# Executive Summary

This project implements a **monthly momentum-driven sector rotation strategy** applied to the eleven S&P 500 Select Sector SPDR ETFs (XLB, XLC, XLE, XLF, XLI, XLK, XLP, XLRE, XLV, XLU, XLY), benchmarked against SPY. At the end of each month, a **Kalman Filter** is applied to each sector's adjusted close history to extract a noise-filtered trend velocity. Sectors are then **clustered daily by k-means** on this single velocity signal into $k$ momentum regimes; the cluster with the highest mean velocity — the "Bull cluster" — is held at equal weight, with positions entered at the next day's market open and adjusted whenever the cluster assignment changes. A **yield curve overlay** (FRED 10Y-2Y spread) moves the portfolio entirely to cash during yield curve inversions, providing a macro-grounded risk-off mechanism. Parameters $k$ (number of clusters) and $Q_2$ (Kalman velocity noise) are jointly optimized on training data (June 2000 – December 2024) using a 4×4 grid search. Out-of-sample evaluation covers January 2025 through March 2026.

---

# Introduction & Hypothesis

## Sector Rotation and the Business Cycle

Equity markets do not move uniformly across sectors. The business cycle theory (Fama & French, 1989; Stovall, 1996) proposes that different sectors tend to outperform at different stages of the economic cycle:

| Cycle Phase | Typically Outperforming Sectors |
|---|---|
| Early Expansion | XLF (Financials), XLY (Consumer Discretionary) |
| Mid Expansion | XLI (Industrials), XLK (Technology), XLC (Communication) |
| Late Expansion | XLE (Energy), XLB (Materials) |
| Contraction / Defensive | XLU (Utilities), XLP (Consumer Staples), XLV (Health Care) |

A sector rotation strategy attempts to be positioned in the sectors most likely to outperform during the current phase of the cycle.

## Academic Foundation

Moskowitz & Grinblatt (1999, *Journal of Finance*) demonstrated that **industry-level momentum** — the tendency for recently outperforming industries to continue outperforming — is statistically significant and economically large, explaining a substantial fraction of individual stock momentum. Critically, industry momentum profits are concentrated on the long side (winning industries), making a long-only implementation viable. Faber (2010, SSRN 1585517) tested relative-strength ranking directly on sector ETF-equivalent data back to the 1920s, confirming profitability across all lookback windows from 1 to 12 months and showing that a trend-following overlay reduces drawdowns. Moskowitz, Ooi & Pedersen (2012, *Journal of Financial Economics*) further established **time-series momentum** — each asset's own past return predicting its own future return — providing the theoretical basis for using a sector's individual velocity signal rather than only cross-sectional rank.

## Strategy Innovation

This project extends the standard rank-based momentum approach in two ways:

1. **Kalman Velocity as Clustering Feature**: Instead of clustering sectors on raw historical returns or arbitrary technical indicators, the Kalman Filter first extracts a noise-filtered trend velocity for each sector. K-means then clusters sectors by this single, economically interpretable signal — partitioning them into momentum regimes (Bull/Bear). Using a denoised signal prevents single-day outliers from distorting cluster assignments.

2. **Yield Curve Regime Overlay**: The 10Y-2Y Treasury yield spread — a well-documented leading indicator of US recessions (Chicago Fed, 2018; Gilchrist & Zakrajsek, 2012 *AER*) — provides a macro anchor. When the yield curve inverts below -0.25%, the strategy exits all positions and holds cash entirely. This preserves capital during late-cycle risk-off conditions without requiring a forecast of which sectors will outperform in the downturn.

---

# Mental Model

## Decision Flowchart

The following diagram illustrates the complete daily decision process:

```{mermaid}
%%{init: {'theme': 'default'}}%%
flowchart LR
    A([Daily Close]) --> B[Apply Kalman Filter\ndlmFilter — causal forward pass only]
    B --> C[Extract kal_velocity\nnoise-filtered trend slope per sector]
    C --> D{Yield curve inverted?\n10Y-2Y < -0.25%}
    D -- Yes --> E[ALL FLAT\nGo to cash]
    D -- No --> F[K-Means Clustering\nk clusters on kal_velocity]
    F --> G[Identify Bull Cluster\nhighest mean kal_velocity]
    G --> H[LONG sectors in Bull Cluster\nequal weight]
    E --> I[TC: 0.05% per unit weight change]
    H --> I
    I --> J([Execute at Open T+1\nearn O2C return])
    J --> A
```

## Decision Rules

| Condition | Signal | Execution Timing | Return Captured |
|---|---|---|---|
| Yield curve inverted (10Y-2Y < -0.25%) | ALL FLAT | Exit all positions at Open$_{T+1}$ | None (hold cash) |
| Yield curve normal AND sector in Bull cluster | LONG | Buy at Open$_{T+1}$ | Open-to-Close on $T+1$ |
| Yield curve normal AND sector NOT in Bull cluster | FLAT | Sell (if held) at Open$_{T+1}$ | None (cash = 0%) |

The strategy is **long-only** with **daily signal generation**. The cluster is re-evaluated every trading day using the latest Kalman velocity; a position is only entered or exited when the signal actually changes, so transaction costs are proportional to realized turnover rather than a fixed calendar schedule.

---

# Data

## Symbols and Date Ranges

```{r}
#| label: data_download
#| include: false
#| cache: true

# Download all 12 equity symbols
raw_prices <- tq_get(
  SYMBOLS,
  get  = "stock.prices",
  from = DOWNLOAD_START,
  to   = TEST_END
)

# Download FRED 10Y-2Y Treasury yield spread (T10Y2Y series)
# Positive = normal curve; negative = inverted (recession signal)
yc_raw <- tq_get(
  "T10Y2Y",
  get  = "economic.data",
  from = DOWNLOAD_START,
  to   = TEST_END
) %>%
  rename(yield_spread = price) %>%
  mutate(yield_spread = yield_spread / 100)   # convert from percentage points to decimal
```

## Data Coverage and Quality

```{r}
#| label: data_quality

coverage <- raw_prices %>%
  group_by(symbol) %>%
  summarise(
    first_date  = min(date),
    last_date   = max(date),
    n_rows      = n(),
    n_na_adj    = sum(is.na(adjusted)),
    n_na_open   = sum(is.na(open)),
    .groups     = "drop"
  ) %>%
  arrange(first_date)

kable(coverage,
      col.names = c("Symbol","First Date","Last Date","Rows","NA (Adjusted)","NA (Open)"),
      caption   = "Table 1: Data Coverage by Symbol",
      format    = "html") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed")) %>%
  row_spec(which(coverage$symbol %in% c("XLC","XLRE")), background = "#fff3cd")
```

::: {.callout-note title="ETF Launch Dates"}
**XLC** (Communication Services) launched `r format(XLC_LAUNCH, "%B %d, %Y")`. **XLRE** (Real Estate) launched `r format(XLRE_LAUNCH, "%B %d, %Y")`. Prior to each ETF's launch date, that sector is simply absent from the clustering universe — k-means operates only on available sectors, and the Bull cluster is identified from the available set. No values are imputed or fabricated.
:::

::: {.callout-warning title="Adjusted Price Caveat"}
Yahoo Finance computes `adjusted` close using the **full forward history** of dividends and splits — i.e., the adjustment factor for January 2001 incorporates dividends paid in 2010. This creates a minor look-ahead bias in the feature engineering (momentum signals use adjusted close). This is standard practice in academic finance and is disclosed here. The impact is small for sector ETFs, which pay modest quarterly dividends and rarely split.
:::

## Data Split

```{r}
#| label: data_split
#| include: false

# Training data (June 2000 – Dec 2024)
train_all    <- raw_prices %>% filter(date >= TRAIN_START, date <= TRAIN_END)
sector_train <- train_all %>% filter(symbol %in% SECTOR_SYMBOLS)
spy_train    <- train_all %>% filter(symbol == "SPY")

# Testing data (Jan 2025 – Mar 2026)
test_all    <- raw_prices %>% filter(date >= TEST_START, date <= TEST_END)
sector_test <- test_all %>% filter(symbol %in% SECTOR_SYMBOLS)
spy_test    <- test_all %>% filter(symbol == "SPY")

# Yield curve split — FRED data has gaps (weekends/holidays); forward-fill to trading days
# The full series is kept unsplit here; filtering happens inside signal generation
yc_filled <- yc_raw %>%
  arrange(date) %>%
  mutate(yield_spread = zoo::na.locf(yield_spread, na.rm = FALSE))

cat("Training rows (sectors):", nrow(sector_train), "\n")
cat("Testing rows (sectors):", nrow(sector_test), "\n")
cat("Training date range:", format(min(train_all$date)), "–", format(max(train_all$date)), "\n")
cat("Testing date range:",  format(min(test_all$date)),  "–", format(max(test_all$date)),  "\n")
```

> **Look-ahead firewall**: `test_all`, `sector_test`, and `spy_test` are created here but not accessed again until Section 9. All model fitting, parameter selection, and normalization statistics are derived exclusively from training data.

---

# Methodology

## 5.1 Kalman Filter — Extracting Noise-Filtered Trend Velocity

### State-Space Model

The Kalman Filter models a time series as a **hidden Markov process**: the observed daily price is treated as a noisy measurement of an unobserved "true" underlying state. We use a **local level + slope** (second-order polynomial) model:

**State equations**:
$$\mu_{t+1} = \mu_t + \beta_t + w_{1,t}, \quad w_{1,t} \sim \mathcal{N}(0, Q_1)$$
$$\beta_{t+1} = \beta_t + w_{2,t}, \quad w_{2,t} \sim \mathcal{N}(0, Q_2)$$

**Observation equation**:
$$y_t = \mu_t + v_t, \quad v_t \sim \mathcal{N}(0, R)$$

where $\mu_t$ is the underlying price level, $\beta_t$ is the **velocity** (daily rate of change of the underlying trend), and $y_t$ is the observed daily adjusted close price.

### Parameters

- **$Q_1$** (level process noise): controls how quickly the estimated level can change. Larger $Q_1$ allows the filter to track rapid level changes.
- **$Q_2$** (velocity process noise): controls how quickly the estimated velocity can change. This is the key tuning parameter — larger $Q_2$ allows more volatile trend direction changes.
- **$R$** (observation noise): represents daily price measurement noise. Fixed at 1.0 relative to process noise.

### Filter vs. Smoother: The Look-Ahead Distinction

The Kalman **filter** (`dlmFilter`) is a **forward-only** algorithm: the state estimate at time $t$ uses only observations $y_1, \ldots, y_t$. This is causal and appropriate for real-time signal generation.

The Kalman **smoother** (`dlmSmooth`) runs a backward pass using $y_{t+1}, \ldots, y_T$ — it produces better state estimates in hindsight but **uses future information**. Using the smoother for signal generation would introduce look-ahead bias and overstate realistic performance. **The smoother is not used anywhere in this analysis.**

### Why Velocity is the Primary Feature

The velocity state $\beta_t$ represents the filtered estimate of the sector's daily trend momentum. A **positive and increasing velocity** indicates a strengthening uptrend; a **negative velocity** signals a downtrend. Unlike raw daily returns, the Kalman velocity is robust to single-day outliers and temporary reversals, providing a more stable directional signal for clustering.

## 5.2 K-Means Clustering on Kalman Velocity

On each trading day, the Kalman velocity $\beta_t$ for each of the 11 sectors is extracted and used as the **sole feature** for k-means clustering. K-means partitions the sectors into $k$ clusters by minimizing within-cluster sum of squares (WCSS):

$$\underset{C_1,\ldots,C_k}{\arg\min} \sum_{j=1}^{k} \sum_{i \in C_j} \left(\beta_i - \bar{\beta}_{C_j}\right)^2$$

The cluster with the highest mean Kalman velocity is designated the **Bull cluster**; all sectors assigned to it receive a LONG signal for the following month. All other sectors receive a FLAT signal.

**Why k-means on a single feature?** Using only $\beta_t$ keeps the clustering economically interpretable: the algorithm is purely partitioning sectors into momentum regimes based on their noise-filtered trend speed. There is no risk of clustering artifacts from mixing features with different scales or units. The result is a hard regime classifier — sectors are either trending up (Bull) or not — which is easier to explain and audit than a soft scoring rule.

**Why Kalman velocity rather than raw returns?** Raw returns are dominated by single-day outliers. The Kalman velocity is a sequential weighted average of the price history — placing higher weight on recent observations while retaining memory of the longer-term trend. Clustering on this denoised signal produces more stable monthly assignments, reducing unnecessary turnover relative to clustering on raw returns.

**Why k-means rather than hierarchical or DBSCAN?** For a one-dimensional feature (a single velocity value per sector per month), k-means is both optimal and efficient. With 11 observations and 1 feature per rebalance date, DBSCAN would be ill-conditioned and hierarchical clustering adds no information beyond what k-means provides. The cluster count $k$ is a free parameter optimized jointly with $Q_2$.

## 5.3 Yield Curve Macro Filter

The 10-year minus 2-year US Treasury spread (FRED series `T10Y2Y`) is one of the most-studied macroeconomic leading indicators. When this spread turns negative — meaning short-term rates exceed long-term rates — it has preceded every US recession since the 1960s (Estrella & Mishkin 1998; Chicago Fed 2018).

When the spread falls below $-0.25\%$, the strategy exits all positions and holds cash. Rather than forecasting which sectors will be resilient — a second prediction problem layered on top of the momentum signal — the strategy simply removes market exposure entirely. This is the most conservative response to a yield curve inversion: it avoids the drawdown from holding risky sectors during a recession, while also avoiding the complexity of defending a specific sector selection in the risk-off regime.

This filter transforms the strategy from pure price-momentum to a **macro-conditioned momentum strategy**, grounding the risk-off behavior in economic theory rather than technical signals alone.

## 5.4 Signal Pipeline and Execution

```{mermaid}
%%{init: {'theme': 'default'}}%%
flowchart LR
    subgraph DATA ["Data: tq_get"]
        D1["Daily OHLCV\n12 equity symbols"]
        D2["FRED T10Y2Y\nyield spread (daily)"]
    end
    subgraph KALMAN ["Kalman Filter: dlm"]
        K1["dlmModPoly order=2\nlevel + slope"] --> K2["dlmFilter\nforward pass only"]
        K2 --> K3["kal_velocity\nnoise-filtered trend speed"]
    end
    subgraph CLUSTER ["K-Means Clustering (month-end)"]
        C1["k-means on kal_velocity\n1 feature · k clusters"] --> C2["Bull cluster =\nhighest mean velocity"]
    end
    subgraph SIGNAL ["Daily Signal"]
        S1["Check yield curve\nT10Y2Y vs threshold"] --> S2{Inverted?}
        S2 -- Yes --> S3["ALL FLAT\nGo to cash"]
        S2 -- No  --> S4["LONG sectors in Bull cluster\nequal weight"]
    end
    subgraph PORT ["Portfolio Construction"]
        P1["Lag signal 1 day\nexecute at Open T+1"] --> P2["Earn O2C return\nwhile signal = LONG"]
        P2 --> P3["Apply 0.05% TC\nonly on weight changes"]
    end
    subgraph OPT ["Parallel Optimization"]
        O1["Grid: k=2–5\nQ=4 log-spaced values"] --> O2["3D Omega surface\nk × log(Q)"]
        O2 --> O3["Select optimal k, Q\ntraining data only"]
    end
    subgraph EVAL ["Evaluation"]
        E1["Omega · Sharpe · MaxDD\n%Win · %InMkt · #Rebalances"] --> E2["Z-scored grid viz"]
        E2 --> E3["Stress tests\n2×TC · no-GFC · no-COVID"]
    end
    DATA --> KALMAN
    KALMAN --> CLUSTER
    CLUSTER --> SIGNAL
    D2 --> SIGNAL
    SIGNAL --> PORT --> OPT --> EVAL
```

## 5.5 Trade Execution Timing and Return Types

The timing convention is fundamental to a realistic backtest:

```
Day T [Close]:   Kalman velocity extracted → k-means cluster → LONG/FLAT signal
Day T+1 [Open]:  Execute: enter new LONG positions, exit positions that flipped to FLAT
Day T+1 [Close]: Earn O2C return on all held positions
Day T+2:         Repeat with updated velocities — position unchanged if cluster unchanged
```

Three return types are computed and reported:

| Return Type | Formula | Interpretation |
|---|---|---|
| **Close-to-Close (C2C)** | $\frac{\text{Adj}_{t}}{\text{Adj}_{t-1}} - 1$ | Full 24-hour return; used for passive SPY benchmark |
| **Open-to-Close (O2C)** | $\frac{\text{Close}_{t}}{\text{Open}_{t}} - 1$ | Intraday return; **strategy earns this** when entering at open |
| **Close-to-Open (C2O)** | $\frac{\text{Open}_{t}}{\text{Close}_{t-1}} - 1$ | Overnight gap return; exposes overnight risk |

Note: O2C and C2O use **raw (unadjusted)** close and open prices, which more accurately represent actual execution. C2C uses **adjusted** close for backward-compatible trend comparisons. This distinction is disclosed.

**Strategy return on day $T+1$**: the sector earns its O2C return ($\frac{\text{Close}_{T+1}}{\text{Open}_{T+1}} - 1$), as the position is entered at open and held to close.

## 5.6 Transaction Costs

| Component | Assumption | Rationale |
|---|---|---|
| Bid-ask spread (one-way) | 0.03% | Liquid sector ETFs trade with ~0.01-0.05% spreads |
| Commission | 0.02% | Conservative estimate; many brokers now charge $0 |
| **Total (one-way)** | **0.05%** | Applied as $|\Delta w_i| \times 0.0005$ per dollar rebalanced |
| Cash return | 0% | Conservative; money market rates not credited |
| Short sale constraints | N/A | Long-only strategy; no short positions |

Total round-trip cost per sector entry/exit: **0.10%**. Results are reported both gross and net of transaction costs.

---

<!-- Implementation chunks run silently below -->

```{r}
#| label: helper_functions
#| include: false

# ─── Kalman Filter ─────────────────────────────────────────────────────────────
run_kalman <- function(price_vec, q_level = DEFAULT_Q_LEVEL,
                        q_velocity = DEFAULT_Q_VELOCITY, r_obs = DEFAULT_R_OBS) {
  # Local level + slope state-space model
  # State: [level, velocity]; Observation: price
  # IMPORTANT: uses dlmFilter (causal), NOT dlmSmooth (uses future data)
  mod  <- dlmModPoly(order = 2, dV = r_obs, dW = c(q_level, q_velocity))
  filt <- dlmFilter(price_vec, mod)
  # filt$m is (n+1) x 2: row 1 = prior state; rows 2:(n+1) = posterior after each obs
  tibble(
    level    = filt$m[-1, 1],  # filtered level aligned with price_vec
    velocity = filt$m[-1, 2]   # filtered velocity (daily rate of change of level)
  )
}

# ─── Omega Ratio ───────────────────────────────────────────────────────────────
# Non-parametric risk-reward measure; accounts for all moments of the distribution
# Omega > 1: expected gains exceed expected losses at threshold t
# More robust than Sharpe when return distribution is non-normal
omega_ratio <- function(returns, threshold = 0) {
  r <- returns[!is.na(returns)]
  excess <- r - threshold
  gains  <- sum(excess[excess > 0])
  losses <- abs(sum(excess[excess < 0]))
  if (losses == 0) return(Inf)
  gains / losses
}

# ─── Daily Signal Generation ──────────────────────────────────────────────────
# Runs k-means on Kalman velocity every trading day (single feature, k clusters).
# Bull cluster = cluster with highest mean kal_velocity → LONG signal.
# Yield curve override: if 10Y-2Y < threshold → ALL sectors FLAT (go to cash).
# Signal at Close_T → executed at Open_{T+1} via 1-day lag in build_portfolio.
# Returns one row per sector per trading day.
generate_signals_clustered <- function(kal_df, yc_df,
                                       k_clusters = DEFAULT_K,
                                       yc_thresh  = YC_THRESHOLD) {
  # Forward-fill yield curve to every trading day
  all_trading_days <- sort(unique(kal_df$date))
  yc_daily <- tibble(date = all_trading_days) %>%
    left_join(yc_df %>% select(date, yield_spread), by = "date") %>%
    arrange(date) %>%
    mutate(yield_spread = zoo::na.locf(yield_spread, na.rm = FALSE))

  # Join yield curve into Kalman data
  kal_sector <- kal_df %>%
    filter(symbol %in% SECTOR_SYMBOLS) %>%
    left_join(yc_daily, by = "date")

  # For each trading day: k-means on kal_velocity, identify Bull cluster
  result_list <- lapply(all_trading_days, function(d) {
    df_d <- kal_sector %>% filter(date == d)
    if (nrow(df_d) < 2) return(NULL)

    yc_inv <- isTRUE(!is.na(df_d$yield_spread[1]) && df_d$yield_spread[1] < yc_thresh)

    # Yield curve inverted → all flat (cash), skip clustering
    if (yc_inv) {
      return(df_d %>%
        mutate(cluster = NA_integer_, yc_inverted = TRUE, signal = "FLAT") %>%
        select(symbol, date, signal, cluster, kal_velocity, yield_spread, yc_inverted))
    }

    # Guard: cannot have more clusters than observations
    k_actual <- min(k_clusters, nrow(df_d))
    km       <- kmeans(matrix(df_d$kal_velocity, ncol = 1),
                       centers = k_actual, nstart = 25, iter.max = 50)
    df_d$cluster <- as.integer(km$cluster)

    cluster_means <- tapply(df_d$kal_velocity, df_d$cluster, mean)
    bull_cluster  <- as.integer(names(which.max(cluster_means)))

    df_d %>%
      mutate(
        yc_inverted = FALSE,
        signal      = if_else(cluster == bull_cluster, "LONG", "FLAT")
      ) %>%
      select(symbol, date, signal, cluster, kal_velocity, yield_spread, yc_inverted)
  })

  bind_rows(result_list)
}

# ─── Portfolio Construction ────────────────────────────────────────────────────
# Daily signals: signal at Close_T → executed at Open_{T+1} via 1-day lag.
# Strategy earns Open-to-Close return on each day the lagged signal is LONG.
build_portfolio <- function(signals_df, returns_df, tc_rate = TC_RATE) {
  # Lag signal by 1 day per sector: today's cluster → tomorrow's position
  daily <- signals_df %>%
    arrange(symbol, date) %>%
    group_by(symbol) %>%
    mutate(exec_signal = dplyr::lag(signal, default = "FLAT")) %>%
    ungroup() %>%
    left_join(returns_df %>% select(symbol, date, o2c_ret), by = c("symbol", "date")) %>%
    filter(!is.na(o2c_ret))

  # Equal-weight among LONG sectors each day; TC on weight change
  daily %>%
    group_by(date) %>%
    mutate(
      n_long = sum(exec_signal == "LONG"),
      weight = if_else(exec_signal == "LONG" & n_long > 0L, 1 / n_long, 0)
    ) %>%
    ungroup() %>%
    group_by(symbol) %>%
    arrange(date, .by_group = TRUE) %>%
    mutate(tc_cost = abs(weight - dplyr::lag(weight, default = 0)) * tc_rate) %>%
    ungroup() %>%
    group_by(date) %>%
    summarise(
      port_ret_gross = sum(weight * replace_na(o2c_ret, 0), na.rm = TRUE),
      total_tc       = sum(tc_cost, na.rm = TRUE),
      n_positions    = sum(weight > 0),
      .groups        = "drop"
    ) %>%
    mutate(port_ret_net = port_ret_gross - total_tc)
}

# ─── Performance Metrics ───────────────────────────────────────────────────────
compute_metrics <- function(port_ret_vec, spy_ret_vec = NULL, n_sectors = 11,
                            label = "", tc_rate = TC_RATE) {
  r    <- port_ret_vec[!is.na(port_ret_vec)]
  n    <- length(r)
  ann  <- 252

  cagr    <- prod(1 + r)^(ann / n) - 1
  sr      <- if (isTRUE(sd(r) > 0)) mean(r) / sd(r) * sqrt(ann) else NA_real_
  om      <- omega_ratio(r)
  cum_r   <- cumprod(1 + r)
  peak    <- cummax(cum_r)
  dds     <- (cum_r - peak) / peak
  max_dd  <- min(dds)
  in_dd   <- dds < -1e-8
  dd_rle  <- rle(in_dd)
  dd_lens <- dd_rle$lengths[dd_rle$values]
  max_dd_len <- if (length(dd_lens) > 0) max(dd_lens) else 0L
  pct_win    <- mean(r > 0, na.rm = TRUE)

  cor_spy <- if (!is.null(spy_ret_vec)) {
    s <- na.omit(cbind(r, spy_ret_vec[seq_along(r)]))
    cor(s[, 1], s[, 2])
  } else NA_real_

  tibble(
    Period          = label,
    `CAGR (%)`      = round(cagr * 100, 2),
    `Sharpe`        = round(sr, 3),
    `Omega`         = round(om, 3),
    `Max DD (%)`    = round(max_dd * 100, 2),
    `Max DD (days)` = max_dd_len,
    `% Win`         = round(pct_win * 100, 1),
    `Cor SPY`       = round(cor_spy, 3)
  )
}
```


```{r}
#| label: apply_kalman_train
#| cache: true
#| include: false

apply_kalman <- function(df, q_vel = DEFAULT_Q_VELOCITY) {
  kal <- run_kalman(df$adjusted, q_velocity = q_vel)
  df %>%
    mutate(
      kal_level    = kal$level,
      kal_velocity = kal$velocity,
      # Return types needed for portfolio construction
      c2c_ret = adjusted / dplyr::lag(adjusted) - 1,
      o2c_ret = close    / open - 1,
      c2o_ret = open     / dplyr::lag(close)    - 1
    )
}

kalman_train <- sector_train %>%
  group_by(symbol) %>%
  arrange(date, .by_group = TRUE) %>%
  group_modify(~ apply_kalman(.x)) %>%
  ungroup()

# Return table for portfolio construction (training period only)
returns_train <- kalman_train %>%
  filter(date >= TRAIN_START, date <= TRAIN_END) %>%
  select(symbol, date, c2c_ret, o2c_ret, c2o_ret)

# SPY benchmark returns
spy_ret_train <- spy_train %>%
  arrange(date) %>%
  mutate(spy_c2c = adjusted / dplyr::lag(adjusted) - 1) %>%
  filter(!is.na(spy_c2c)) %>%
  select(date, spy_c2c)
```


## Default Ranking Inspection

```{r}
#| label: default_ranking_inspection

# Default model: DEFAULT_K clusters, DEFAULT_Q_VELOCITY
train_signals_default <- generate_signals_clustered(
  kalman_train, yc_filled,
  k_clusters = DEFAULT_K, yc_thresh = YC_THRESHOLD
)

# Sector frequency in Bull cluster by year
rank_freq <- train_signals_default %>%
  mutate(year = year(date)) %>%
  group_by(symbol, year) %>%
  summarise(
    pct_selected = mean(signal == "LONG", na.rm = TRUE),
    .groups = "drop"
  )

hm_wide <- rank_freq %>%
  pivot_wider(names_from = year, values_from = pct_selected, values_fill = 0)
hm_mat  <- as.matrix(hm_wide[, -1])
rownames(hm_mat) <- hm_wide$symbol

plot_ly(
  x          = colnames(hm_mat),
  y          = rownames(hm_mat),
  z          = hm_mat,
  type       = "heatmap",
  colorscale = list(c(0,"#f5f5f5"), c(0.5,"#42a5f5"), c(1,"#1565c0")),
  zmin = 0, zmax = 1,
  hovertemplate = "%{y} | %{x}<br>Selected: %{z:.0%}<extra></extra>"
) %>%
  plotly::layout(
    title  = "Figure 1: Bull-Cluster Selection Frequency by Year (Training, Default Parameters)",
    xaxis  = list(title = ""),
    yaxis  = list(title = ""),
    margin = list(l = 60)
  )
```

---

# Parameter Optimization

The strategy has two free parameters: the number of clusters $k$ and the Kalman velocity process noise $Q_2$. We optimize these jointly by evaluating the **Omega Ratio** of the training-period portfolio across a $4 \times 4 = 16$ parameter grid, run in parallel across CPU cores.

::: {.callout-warning title="Optimization and Data Snooping"}
The parameter grid is evaluated exclusively on training data. The test set remains sealed. The 3D surface plot reveals the objective function shape — a broad plateau indicates robustness, while a sharp spike suggests overfitting to the training history.
:::

```{r}
#| label: param_grid_setup
#| include: false

k_grid     <- K_GRID                            # 2, 3, 4, 5
q_grid     <- c(0.0001, 0.001, 0.01, 0.1)      # log-spaced over 4 orders of magnitude
param_grid <- expand.grid(k = k_grid, q = q_grid)      # 16 combinations
```

```{r}
#| label: parallel_optimization
#| cache: true
#| include: false

cl <- makeCluster(parallel::detectCores() - 1)
registerDoParallel(cl)

clusterExport(cl, varlist = c(
  "run_kalman", "apply_kalman", "generate_signals_clustered", "build_portfolio",
  "omega_ratio",
  "sector_train", "yc_filled",
  "TRAIN_START", "TRAIN_END", "TC_RATE",
  "SECTOR_SYMBOLS", "YC_THRESHOLD",
  "DEFAULT_Q_LEVEL", "DEFAULT_R_OBS", "DEFAULT_K"
), envir = .GlobalEnv)

opt_results <- foreach(
  idx      = 1:nrow(param_grid),
  .combine = "rbind",
  .packages = c("dlm", "dplyr", "tidyr", "tibble", "purrr", "lubridate", "zoo")
) %dopar% {

  ki <- param_grid$k[idx]
  qi <- param_grid$q[idx]

  tryCatch({
    # 1. Apply Kalman with this Q_velocity
    kal_df <- sector_train %>%
      dplyr::group_by(symbol) %>%
      dplyr::arrange(date, .by_group = TRUE) %>%
      dplyr::group_modify(~ apply_kalman(.x, q_vel = qi)) %>%
      dplyr::ungroup()

    # 2. Generate daily signals (k-means clustering on kal_velocity + YC filter)
    sigs <- generate_signals_clustered(kal_df, yc_filled,
                                       k_clusters = ki, yc_thresh = YC_THRESHOLD)

    # 3. Returns table for this Kalman parameterisation
    ret_df <- kal_df %>%
      dplyr::filter(date >= TRAIN_START, date <= TRAIN_END) %>%
      dplyr::select(symbol, date, c2c_ret, o2c_ret, c2o_ret)

    # 4. Build portfolio and evaluate
    port <- build_portfolio(sigs, ret_df, tc_rate = TC_RATE)
    r    <- na.omit(port$port_ret_net)

    if (length(r) < 50) {
      tibble::tibble(k = ki, q = qi, omega = NA_real_, sharpe = NA_real_,
                     max_dd = NA_real_, pct_win = NA_real_, n_obs = length(r))
    } else {
      cr <- cumprod(1 + r); pk <- cummax(cr)
      tibble::tibble(
        k       = ki,
        q       = qi,
        omega   = omega_ratio(r),
        sharpe  = mean(r) / sd(r) * sqrt(252),
        max_dd  = min((cr - pk) / pk),
        pct_win = mean(r > 0),
        n_obs   = length(r)
      )
    }
  }, error = function(e) {
    tibble::tibble(k = ki, q = qi, omega = NA_real_, sharpe = NA_real_,
                   max_dd = NA_real_, pct_win = NA_real_, n_obs = 0L)
  })
}

stopCluster(cl)
```

## Omega Ratio Parameter Surface

```{r}
#| label: surface_plot_omega

surface_wide <- opt_results %>%
  mutate(log_q = log10(q)) %>%
  select(k, log_q, omega) %>%
  pivot_wider(names_from = log_q, values_from = omega)

z_mat <- as.matrix(surface_wide[, -1])
rownames(z_mat) <- surface_wide$k

plot_ly(
  x          = sort(unique(log10(opt_results$q))),
  y          = sort(unique(opt_results$k)),
  z          = z_mat,
  type       = "surface",
  colorscale = list(c(0,"#d32f2f"), c(0.5,"#ffeb3b"), c(1,"#1565c0")),
  showscale  = TRUE
) %>%
  plotly::layout(
    title = list(text = "Figure 2: Omega Ratio Surface — Parameter Grid (Training Period)"),
    scene = list(
      xaxis = list(title = "log\u2081\u2080(Kalman Q\u2082 velocity noise)"),
      yaxis = list(title = "Clusters (k)"),
      zaxis = list(title = "Omega Ratio")
    )
  )
```

## Optimal Parameter Selection

```{r}
#| label: optimal_params
#| include: false

best_row <- opt_results %>%
  filter(!is.na(omega)) %>%
  slice_max(omega, n = 1)

OPTIMAL_K <- best_row$k[1]
OPTIMAL_Q <- best_row$q[1]

cat("Optimal parameters (training Omega criterion):\n")
cat("  k (clusters)      :", OPTIMAL_K, "\n")
cat("  Q (velocity noise):", OPTIMAL_Q, "\n")
cat("  Training Omega    :", round(best_row$omega[1], 3), "\n")
cat("  Training Sharpe   :", round(best_row$sharpe[1], 3), "\n")
```

## Z-Scored Risk Metrics Across Parameter Grid

Z-scoring allows visual comparison of metrics that are on different scales. Values > 0 are above average; values > 1 are more than 1 SD above average.

```{r}
#| label: zscored_metrics

zscored <- opt_results %>%
  filter(!is.na(omega)) %>%
  mutate(
    z_omega   = as.vector(scale(omega)),
    z_sharpe  = as.vector(scale(sharpe)),
    z_max_dd  = as.vector(scale(-max_dd)),
    z_pct_win = as.vector(scale(pct_win))
  ) %>%
  arrange(desc(z_omega)) %>%
  select(k, q, omega, sharpe, max_dd, pct_win, z_omega, z_sharpe, z_max_dd, z_pct_win)

kable(zscored,
      digits  = 3,
      caption = "Table 3: Z-Scored Risk Metrics Across Parameter Grid (Training)",
      format  = "html",
      col.names = c("k (clusters)","Q","Omega","Sharpe","Max DD","% Win",
                    "z(Omega)","z(Sharpe)","z(Max DD)","z(% Win)")) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed")) %>%
  row_spec(1, bold = TRUE, background = "#e3f2fd")
```

---

# Training Period Results

::: {.callout-note title="In-Sample Performance Caveat"}
Results in this section are **in-sample** — the model was fit on this same data. In-sample performance is expected to look favorable due to model fitting, and should not be interpreted as an indication of forward-looking performance. The more meaningful evaluation is in Section 9 (Testing Period Results).
:::


```{r}
#| label: refit_optimal
#| cache: true
#| include: false

# Apply Kalman with OPTIMAL_Q to full training data
kalman_train_opt <- sector_train %>%
  group_by(symbol) %>%
  arrange(date, .by_group = TRUE) %>%
  group_modify(~ apply_kalman(.x, q_vel = OPTIMAL_Q)) %>%
  ungroup()

# Generate daily signals with OPTIMAL_K and OPTIMAL_Q
train_signals_opt <- generate_signals_clustered(
  kalman_train_opt, yc_filled,
  k_clusters = OPTIMAL_K, yc_thresh = YC_THRESHOLD
)

# Returns for training period
returns_train_opt <- kalman_train_opt %>%
  filter(date >= TRAIN_START, date <= TRAIN_END) %>%
  select(symbol, date, c2c_ret, o2c_ret, c2o_ret)

# Portfolio
portfolio_train <- build_portfolio(train_signals_opt, returns_train_opt, tc_rate = TC_RATE)

# Merge with SPY benchmark (spy_ret_train built in apply_kalman_train chunk)
train_combined <- portfolio_train %>%
  left_join(spy_ret_train, by = "date") %>%
  filter(!is.na(spy_c2c))
```

## In-Sample Performance Metrics

```{r}
#| label: train_metrics_table

# Count monthly rebalances where a sector enters a LONG position (month-end signal changes)
n_trades_train <- train_signals_opt %>%
  group_by(symbol) %>%
  arrange(date) %>%
  mutate(entry = signal == "LONG" & dplyr::lag(signal, default = "FLAT") == "FLAT") %>%
  ungroup() %>%
  summarise(trades = sum(entry)) %>%
  pull(trades)

train_metrics_strategy <- compute_metrics(
  port_ret_vec = train_combined$port_ret_net,
  spy_ret_vec  = train_combined$spy_c2c,
  label        = "Strategy (Train)"
)
train_metrics_spy <- compute_metrics(
  port_ret_vec = train_combined$spy_c2c,
  label        = "SPY Buy & Hold (Train)"
)

bind_rows(train_metrics_strategy, train_metrics_spy) %>%
  mutate(
    `% In Market` = c(
      round(mean(portfolio_train$n_positions > 0) * 100, 1),
      100
    ),
    `# Trades` = c(n_trades_train, 0)
  ) %>%
  kable(caption = "Table 4: In-Sample Performance Metrics (Training Period — June 2000 – Dec 2024)",
        format = "html") %>%
  kable_styling(bootstrap_options = c("striped","hover"))
```


```{r}
#| label: train_cum_chart

train_cr <- train_combined %>%
  mutate(
    strategy = (cumprod(1 + port_ret_net) - 1) * 100,
    spy      = (cumprod(1 + spy_c2c)      - 1) * 100
  )

plot_ly(train_cr, x = ~date) %>%
  add_lines(y = ~strategy, name = "Strategy (net)",
            line = list(color = "#1565c0", width = 1.5)) %>%
  add_lines(y = ~spy,      name = "SPY Buy & Hold",
            line = list(color = "#b71c1c", width = 1.5)) %>%
  plotly::layout(
    title     = "Figure 3: Cumulative Returns — Training Period (In-Sample)",
    xaxis     = list(title = ""),
    yaxis     = list(title = "Cumulative Return (%)", ticksuffix = "%"),
    hovermode = "x unified",
    legend    = list(x = 0.01, y = 0.99)
  )
```

---

# Testing Period Results

The testing period (January 2025 – March 2026) represents the **true out-of-sample evaluation**. The Kalman filter state carries forward naturally from training — `dlmFilter` processes data sequentially, so applying it to the concatenated price series involves no look-ahead. The optimal parameters (k = `r OPTIMAL_K`, Q = `r OPTIMAL_Q`) are fixed from training. No refitting or parameter adjustment occurs. The test set was not examined at any point prior to this evaluation.


```{r}
#| label: build_test_features
#| cache: true
#| include: false

# Apply Kalman to full price history (train + test concatenated).
# dlmFilter is sequential — filter state at t uses only y_1..y_t; no look-ahead.
full_sector_prices <- bind_rows(sector_train, sector_test) %>%
  arrange(symbol, date)

kalman_full_opt <- full_sector_prices %>%
  group_by(symbol) %>%
  arrange(date, .by_group = TRUE) %>%
  group_modify(~ apply_kalman(.x, q_vel = OPTIMAL_Q)) %>%
  ungroup()
```

```{r}
#| label: test_clustering_and_portfolio
#| include: false

# Generate daily signals for the test period using optimal parameters
# The Kalman state from training carries forward naturally via the full concatenation
kalman_test_only <- kalman_full_opt %>%
  filter(date >= TEST_START, date <= TEST_END)

test_signals <- generate_signals_clustered(
  kalman_test_only, yc_filled,
  k_clusters = OPTIMAL_K, yc_thresh = YC_THRESHOLD
)

# Returns for test period
returns_test <- kalman_full_opt %>%
  filter(symbol %in% SECTOR_SYMBOLS, date >= TEST_START, date <= TEST_END) %>%
  select(symbol, date, c2c_ret, o2c_ret, c2o_ret)

# Build test portfolio
portfolio_test <- build_portfolio(test_signals, returns_test, tc_rate = TC_RATE)

# SPY test benchmark returns
spy_ret_test <- spy_test %>%
  arrange(date) %>%
  mutate(spy_c2c = adjusted / dplyr::lag(adjusted) - 1) %>%
  filter(!is.na(spy_c2c)) %>%
  select(date, spy_c2c)

test_combined <- portfolio_test %>%
  left_join(spy_ret_test, by = "date") %>%
  filter(!is.na(spy_c2c))

cat("Test portfolio constructed.\n")
cat("Test period:", format(min(test_combined$date)), "to", format(max(test_combined$date)), "\n")
cat("Trading days:", nrow(test_combined), "\n")
cat("Avg daily positions:", round(mean(test_combined$n_positions), 2), "\n")
cat("% Days in market:", round(mean(test_combined$n_positions > 0) * 100, 1), "%\n")
```

## Out-of-Sample Performance Metrics

```{r}
#| label: test_metrics_table

n_trades_test <- test_signals %>%
  group_by(symbol) %>%
  arrange(date) %>%
  mutate(entry = signal == "LONG" & dplyr::lag(signal, default = "FLAT") == "FLAT") %>%
  ungroup() %>%
  summarise(trades = sum(entry)) %>%
  pull(trades)

test_metrics_strategy <- compute_metrics(
  port_ret_vec = test_combined$port_ret_net,
  spy_ret_vec  = test_combined$spy_c2c,
  label        = "Strategy (Test)"
)
test_metrics_spy <- compute_metrics(
  port_ret_vec = test_combined$spy_c2c,
  label        = "SPY Buy & Hold (Test)"
)

bind_rows(test_metrics_strategy, test_metrics_spy) %>%
  mutate(
    `% In Market` = c(
      round(mean(portfolio_test$n_positions > 0) * 100, 1),
      100
    ),
    `# Trades` = c(n_trades_test, 0),
    `Gross CAGR` = c(
      round((prod(1 + na.omit(test_combined$port_ret_gross))^(252/nrow(test_combined)) - 1) * 100, 2),
      NA
    )
  ) %>%
  kable(caption = "Table 5: Out-of-Sample Performance Metrics (Testing Period — Jan 2025 – Mar 2026)",
        format = "html") %>%
  kable_styling(bootstrap_options = c("striped","hover")) %>%
  footnote(general = "Strategy return = Open-to-Close of execution day. SPY = Close-to-Close. TC = 0.05% one-way.")
```

## Cumulative Return Chart

```{r}
#| label: cum_return_chart

test_cr <- test_combined %>%
  mutate(
    net   = (cumprod(1 + port_ret_net)   - 1) * 100,
    gross = (cumprod(1 + port_ret_gross) - 1) * 100,
    spy   = (cumprod(1 + spy_c2c)        - 1) * 100
  )

plot_ly(test_cr, x = ~date) %>%
  add_lines(y = ~net,   name = "Strategy (net)",
            line = list(color = "#1565c0", width = 2)) %>%
  add_lines(y = ~gross, name = "Strategy (gross)",
            line = list(color = "#42a5f5", width = 1.5, dash = "dash")) %>%
  add_lines(y = ~spy,   name = "SPY Buy & Hold",
            line = list(color = "#b71c1c", width = 2)) %>%
  plotly::layout(
    title     = "Figure 4: Cumulative Returns — Testing Period (Out-of-Sample)",
    xaxis     = list(title = ""),
    yaxis     = list(title = "Cumulative Return (%)", ticksuffix = "%"),
    hovermode = "x unified",
    legend    = list(x = 0.01, y = 0.99)
  )
```

## Performance Dashboard

```{r}
#| label: pa_charts

# Build xts for PerformanceAnalytics functions that need it
port_xts <- xts(test_combined$port_ret_net, order.by = test_combined$date)
spy_xts  <- xts(test_combined$spy_c2c,      order.by = test_combined$date)
both_xts <- na.omit(merge(port_xts, spy_xts))
colnames(both_xts) <- c("Strategy", "SPY")

# ── Plotly performance dashboard: daily returns + drawdown ──────────────────
dd_df <- test_combined %>%
  mutate(
    strat_cum = cumprod(1 + port_ret_net),
    spy_cum   = cumprod(1 + spy_c2c),
    strat_dd  = (strat_cum - cummax(strat_cum)) / cummax(strat_cum) * 100,
    spy_dd    = (spy_cum   - cummax(spy_cum))   / cummax(spy_cum)   * 100,
    daily_pct = port_ret_net * 100,
    bar_color = if_else(daily_pct >= 0, "gain", "loss")
  )

p_ret <- plot_ly(dd_df, x = ~date) %>%
  add_bars(y = ~daily_pct, color = ~bar_color,
           colors = c("gain" = "#1b5e20", "loss" = "#b71c1c"),
           showlegend = FALSE,
           hovertemplate = "%{x}<br>Daily: %{y:.2f}%<extra></extra>") %>%
  plotly::layout(
    yaxis     = list(title = "Daily Return (%)", ticksuffix = "%"),
    xaxis     = list(title = "", showticklabels = FALSE),
    hovermode = "x unified",
    bargap    = 0
  )

p_dd <- plot_ly(dd_df, x = ~date) %>%
  add_lines(y = ~strat_dd, name = "Strategy",
            line = list(color = "#1565c0", width = 1.5)) %>%
  add_lines(y = ~spy_dd,   name = "SPY",
            line = list(color = "#b71c1c", width = 1.5)) %>%
  plotly::layout(
    yaxis     = list(title = "Drawdown (%)", ticksuffix = "%"),
    xaxis     = list(title = ""),
    hovermode = "x unified",
    legend    = list(x = 0.01, y = 0.05)
  )

subplot(p_ret, p_dd, nrows = 2, shareX = TRUE, titleY = TRUE,
        heights = c(0.45, 0.55)) %>%
  plotly::layout(
    title = "Figure 5: Performance Dashboard — Daily Returns & Drawdown (Testing Period)"
  )
```

## Monthly Holdings Summary

```{r}
#| label: monthly_holdings

# Which sectors were held each month (test period)
monthly_holdings <- test_signals %>%
  filter(signal == "LONG") %>%
  mutate(ym = floor_date(date, "month")) %>%
  group_by(ym) %>%
  summarise(
    held_sectors = paste(sort(unique(symbol)), collapse = ", "),
    n_held       = n_distinct(symbol),
    .groups      = "drop"
  )

kable(monthly_holdings,
      col.names = c("Month","Sectors Held","# Sectors"),
      caption   = "Table 6: Monthly Holdings — Testing Period",
      format    = "html") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"))
```

## Sector Positions Over Time

```{r}
#| label: positions_chart

# Heatmap: which sectors were LONG on each day of the test period
pos_df <- test_signals %>%
  mutate(held = as.integer(signal == "LONG"))

pos_wide <- pos_df %>%
  select(date, symbol, held) %>%
  pivot_wider(names_from = symbol, values_from = held, values_fill = 0)

pos_mat  <- as.matrix(pos_wide[, -1])
rownames(pos_mat) <- as.character(pos_wide$date)

plot_ly(
  x          = colnames(pos_mat),
  y          = rownames(pos_mat),
  z          = t(pos_mat),
  type       = "heatmap",
  colorscale = list(c(0, "#f5f5f5"), c(1, "#1565c0")),
  showscale  = FALSE,
  hovertemplate = "<b>%{x}</b><br>%{y}<br>%{z:d} (1=LONG, 0=FLAT)<extra></extra>"
) %>%
  plotly::layout(
    title  = "Figure 6: Sector Positions Held — Testing Period",
    xaxis  = list(title = ""),
    yaxis  = list(title = "", autorange = "reversed",
                  tickfont = list(size = 9))
  )
```

## Daily P&L

```{r}
#| label: daily_pnl_chart

pnl_df <- test_combined %>%
  mutate(
    daily_pct = port_ret_net * 100,
    cum_pct   = (cumprod(1 + port_ret_net) - 1) * 100,
    bar_color = if_else(daily_pct >= 0, "gain", "loss")
  )

plot_ly(pnl_df, x = ~date) %>%
  add_bars(
    y             = ~daily_pct,
    color         = ~bar_color,
    colors        = c("gain" = "#1b5e20", "loss" = "#b71c1c"),
    yaxis         = "y",
    name          = "Daily Return",
    showlegend    = TRUE,
    hovertemplate = "%{x}<br>Daily: %{y:.2f}%<extra></extra>",
    bargap        = 0
  ) %>%
  add_lines(
    y             = ~cum_pct,
    yaxis         = "y2",
    name          = "Cumulative Return",
    line          = list(color = "#1565c0", width = 2),
    hovertemplate = "%{x}<br>Cumulative: %{y:.1f}%<extra></extra>"
  ) %>%
  plotly::layout(
    title     = "Figure 7: Daily P&L and Cumulative Return — Testing Period",
    xaxis     = list(title = ""),
    yaxis     = list(title = "Daily Return (%)", ticksuffix = "%", side = "left"),
    yaxis2    = list(title = "Cumulative Return (%)", ticksuffix = "%",
                     overlaying = "y", side = "right"),
    hovermode = "x unified",
    legend    = list(x = 0.01, y = 0.99),
    bargap    = 0
  )
```

---

# Correlation Analysis

Understanding the strategy's correlation to SPY is important for portfolio construction. A strategy with low or negative correlation to SPY adds diversification value.

```{r}
#| label: rolling_correlation

# 60-day rolling Pearson correlation between strategy and SPY returns
min_window <- 30  # require at least 30 observations for correlation

roll_cor_vec <- rollapply(
  cbind(test_combined$port_ret_net, test_combined$spy_c2c),
  width     = 60,
  FUN       = function(x) {
    cc <- stats::complete.cases(x)
    if (sum(cc) < min_window) return(NA_real_)
    stats::cor(x[cc, 1], x[cc, 2])
  },
  by.column = FALSE,
  align     = "right",
  fill      = NA
)

roll_cor_df <- tibble(
  date    = test_combined$date,
  cor_60d = as.numeric(roll_cor_vec)
)

plot_ly(
  data = roll_cor_df,
  x = ~date, y = ~cor_60d,
  type = "scatter", mode = "lines",
  line = list(color = "#1565c0", width = 2),
  name = "60-day Rolling Correlation"
) %>%
  add_segments(
    x = min(roll_cor_df$date, na.rm=TRUE), xend = max(roll_cor_df$date, na.rm=TRUE),
    y = 0, yend = 0,
    line = list(color = "grey60", dash = "dash"),
    showlegend = FALSE
  ) %>%
  plotly::layout(
    title  = "Figure 8: 60-Day Rolling Correlation — Strategy vs SPY (Testing Period)",
    xaxis  = list(title = ""),
    yaxis  = list(title = "Pearson Correlation", range = c(-1, 1)),
    hovermode = "x unified"
  )

cat("Average correlation to SPY (test period):",
    round(mean(roll_cor_df$cor_60d, na.rm=TRUE), 3), "\n")
```

---

# Stress Testing

::: {.callout-warning title="Backtesting Is Not a Scientific Experiment"}
Stress testing is not statistical proof. We cannot know in advance which historical periods will recur, and the market's response to a strategy changes once the strategy becomes widely known. These tests are presented as **robustness checks** — ways of probing whether the strategy's performance depends critically on specific favorable conditions. They are not forecasts of future resilience.
:::

```{r}
#| label: stress_test_computations
#| include: false

# ── Stress Test 1: Double Transaction Costs (0.10% one-way) ──────────────────
portfolio_stress_2x_tc <- build_portfolio(test_signals, returns_test, tc_rate = TC_RATE * 2)
stress_combined_2x <- portfolio_stress_2x_tc %>%
  left_join(spy_ret_test, by = "date") %>% filter(!is.na(spy_c2c))

# ── Stress Test 2: Exclude GFC (2008–2009) from TRAINING period ───────────────
returns_no_gfc  <- returns_train_opt %>%
  filter(!(date >= "2008-01-01" & date <= "2009-12-31"))
signals_no_gfc  <- train_signals_opt %>%
  filter(!(date >= "2008-01-01" & date <= "2009-12-31"))
portfolio_no_gfc <- build_portfolio(signals_no_gfc, returns_no_gfc, tc_rate = TC_RATE)
spy_no_gfc <- spy_ret_train %>%
  filter(!(date >= "2008-01-01" & date <= "2009-12-31"))
combined_no_gfc <- portfolio_no_gfc %>%
  left_join(spy_no_gfc, by = "date") %>% filter(!is.na(spy_c2c))

# ── Stress Test 3: Exclude COVID Shock (Feb–Jun 2020) from TRAINING ───────────
returns_no_covid <- returns_train_opt %>%
  filter(!(date >= "2020-02-01" & date <= "2020-06-30"))
signals_no_covid <- train_signals_opt %>%
  filter(!(date >= "2020-02-01" & date <= "2020-06-30"))
portfolio_no_covid <- build_portfolio(signals_no_covid, returns_no_covid, tc_rate = TC_RATE)
spy_no_covid <- spy_ret_train %>%
  filter(!(date >= "2020-02-01" & date <= "2020-06-30"))
combined_no_covid <- portfolio_no_covid %>%
  left_join(spy_no_covid, by = "date") %>% filter(!is.na(spy_c2c))
```

```{r}
#| label: stress_metrics_table

stress_table <- bind_rows(
  # Training period baseline
  compute_metrics(train_combined$port_ret_net, train_combined$spy_c2c,
                  label = "Training (Baseline)"),
  # Test period baseline
  compute_metrics(test_combined$port_ret_net,  test_combined$spy_c2c,
                  label = "Testing (Baseline)"),
  # 2x TC (test period)
  compute_metrics(stress_combined_2x$port_ret_net, stress_combined_2x$spy_c2c,
                  label = "Testing (2× TC)"),
  # Exclude GFC (training period analysis)
  compute_metrics(combined_no_gfc$port_ret_net, combined_no_gfc$spy_c2c,
                  label = "Training (excl. GFC 2008-09)"),
  # Exclude COVID (training period analysis)
  compute_metrics(combined_no_covid$port_ret_net, combined_no_covid$spy_c2c,
                  label = "Training (excl. COVID 2020)")
)

kable(stress_table,
      digits  = 2,
      caption = "Table 7: Stress Test Comparison — Strategy Performance Across Scenarios",
      format  = "html") %>%
  kable_styling(bootstrap_options = c("striped","hover")) %>%
  row_spec(2, bold = TRUE, background = "#e3f2fd") %>%
  footnote(general = "Testing scenarios apply the frozen model trained on full training data. Training scenarios exclude specific periods from the training backtest.")
```

```{r}
#| label: stress_cum_chart

sc_std <- test_combined      %>% mutate(cum = (cumprod(1 + port_ret_net) - 1) * 100)
sc_2x  <- stress_combined_2x %>% mutate(cum = (cumprod(1 + port_ret_net) - 1) * 100)

plot_ly(x = ~sc_std$date) %>%
  add_lines(y = ~sc_std$cum, name = "Standard TC (0.05%)",
            line = list(color = "#1565c0", width = 2)) %>%
  add_lines(y = ~sc_2x$cum,  name = "Double TC (0.10%)",
            line = list(color = "#e65100", width = 2, dash = "dash")) %>%
  plotly::layout(
    title     = "Figure 9: Transaction Cost Sensitivity — Testing Period",
    xaxis     = list(title = ""),
    yaxis     = list(title = "Cumulative Return (%)", ticksuffix = "%"),
    hovermode = "x unified",
    legend    = list(x = 0.01, y = 0.99)
  )
```

---

# Conclusion

## Summary of Results

This project implemented a Kalman-filter-based sector rotation strategy using k-means momentum clustering with a yield curve macro overlay. The strategy was trained on 24.5 years of daily SPDR sector ETF data (June 2000 – December 2024) and evaluated out-of-sample over 15 months (January 2025 – March 2026).

The Kalman filter proved effective at extracting noise-filtered trend velocity from daily prices. Clustering sectors monthly on this single velocity signal via k-means partitions the investment universe into economically interpretable momentum regimes — the Bull cluster captures sectors with the strongest trending behavior. The 10Y-2Y yield curve overlay provided a macro-grounded defensive tilt during inversion regimes, adding a fundamental anchor absent from purely technical clustering approaches.

## Limitations

**Single macro signal**: The yield curve overlay uses one FRED series (10Y-2Y spread) with a fixed threshold (-0.25%). Yield curve inversion is a strong but lagging recession predictor — the inversion-to-recession lag historically ranges from 6 to 24 months, meaning the defensive rotation may activate prematurely or persist too long.

**Cluster instability**: K-means can produce different cluster assignments across runs due to random centroid initialization (mitigated here with `nstart = 25`). More critically, sectors near cluster boundaries can flip between Bull and non-Bull assignments on small velocity changes, generating unnecessary trades. Soft clustering methods (Gaussian Mixture Models) that produce membership probabilities rather than hard assignments could reduce boundary sensitivity.

**Static optimization**: Optimal parameters ($k$ and $Q$) are estimated once on the full training period and applied statically to the test period. A walk-forward approach — re-optimizing quarterly on a rolling window — would be more realistic and reduce the risk of regime-change invalidation.

**Equal weighting**: Held sectors are equally weighted regardless of velocity magnitude within the Bull cluster. Velocity-proportional weighting would concentrate exposure on the strongest-trending sectors.

**Out-of-sample period length**: Fifteen months is a short evaluation window. A single bull or bear market regime during this period will dominate the results. Definitive conclusions about strategy robustness require multiple market cycles.

## Future Work

- **Walk-forward backtesting**: Re-optimize $k$ and $Q$ quarterly on rolling 36-month windows
- **Soft clustering**: Replace k-means with Gaussian Mixture Models for probabilistic Bull cluster membership; weight positions by membership probability
- **Multi-feature clustering**: Extend beyond single-feature velocity clustering by adding yield-spread sensitivity or realized volatility as additional clustering dimensions
- **Additional macro filters**: Credit spreads (HYG/LQD ratio), ISM PMI, or Fed funds futures to complement yield curve signal
- **Volatility scaling**: Scale position size inversely to realized volatility (as in Moskowitz, Ooi & Pedersen 2012) to stabilize portfolio risk through regimes

