---
title: "Momentum-Driven Sector Rotation via Kalman Filtering and K-Means Clustering"
subtitle: "FIN452 Quantitative Trading Project"
author: "Carter Vereschagin"
date: today
format:
  html:
    embed-resources: true
    toc: false
    theme: flatly
    fig-width: 10
    fig-height: 6
    df-print: paged
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

```{r}
#| label: setup
#| include: false

# --- Data & Finance ---
library(tidyquant)          # tq_get() — Yahoo Finance + FRED wrapper
library(xts)                # time-series objects
library(PerformanceAnalytics) # charts.PerformanceSummary(), maxDrawdown()

# --- Kalman Filter ---
library(dlm)                # dlmModPoly(), dlmFilter()

# --- Parallelism ---
library(doParallel)         # makeCluster(), registerDoParallel()
library(foreach)            # foreach() %dopar%

# --- Visualization ---
library(plotly)             # interactive charts
library(ggplot2)
library(scales)

# --- Data Wrangling ---
library(dplyr)
library(tidyr)
library(purrr)
library(lubridate)
library(tibble)
library(zoo)                # na.locf() for forward-filling FRED data

# --- Display ---
library(knitr)
library(kableExtra)

# Global options
options(scipen = 999)
theme_set(theme_minimal(base_size = 12))

# ─── Global Constants ──────────────────────────────────────────────────────────
TRAIN_START    <- as.Date("2000-06-01")
TRAIN_END      <- as.Date("2024-12-31")
TEST_START     <- as.Date("2025-01-01")
TEST_END       <- as.Date("2026-03-31")
DOWNLOAD_START <- TRAIN_START - 30   # small buffer; no indicator warm-up needed

TC_RATE        <- 0.0005   # 0.05% one-way transaction cost

# All symbols (SPY = benchmark; remaining 11 = sectors)
SYMBOLS        <- c("SPY","XLB","XLC","XLE","XLF","XLI",
                    "XLK","XLP","XLRE","XLV","XLU","XLY")
SECTOR_SYMBOLS <- setdiff(SYMBOLS, "SPY")

# Kalman filter defaults (Q_velocity overridden by optimised value later)
DEFAULT_Q_LEVEL    <- 0.01
DEFAULT_Q_VELOCITY <- 0.001
DEFAULT_R_OBS      <- 1.0

# Strategy parameters
DEFAULT_K         <- 3                         # default number of k-means clusters
K_GRID            <- 2:6                       # grid search range for k
YC_THRESHOLD      <- -0.0025                   # 10Y-2Y < -0.25% → go to cash (all FLAT)

# Known ETF launch dates (important for data coverage disclosure)
XLC_LAUNCH  <- as.Date("2018-06-18")
XLRE_LAUNCH <- as.Date("2015-10-07")

# Sector name labels for charts — ticker (Full Name)
SECTOR_NAMES <- c(
  XLB  = "XLB (Materials)",
  XLC  = "XLC (Comm. Services)",
  XLE  = "XLE (Energy)",
  XLF  = "XLF (Financials)",
  XLI  = "XLI (Industrials)",
  XLK  = "XLK (Technology)",
  XLP  = "XLP (Consumer Staples)",
  XLRE = "XLRE (Real Estate)",
  XLV  = "XLV (Health Care)",
  XLU  = "XLU (Utilities)",
  XLY  = "XLY (Cons. Discretionary)"
)
```

---

# Strategy

## Description

This project implements a daily momentum-driven sector rotation strategy applied to the eleven S&P 500 Select Sector SPDR ETFs — XLB, XLC, XLE, XLF, XLI, XLK, XLP, XLRE, XLV, XLU, and XLY — benchmarked against SPY. The strategy runs every trading day. A Kalman filter is applied to each sector's adjusted close price series to extract a noise-filtered trend velocity. Those velocities are then passed into a k-means clustering algorithm that partitions all eleven sectors into momentum regimes. The cluster with the highest average velocity — the Bull cluster — defines which sectors to hold that day. When the FRED 10-year minus 2-year Treasury spread falls below −0.25%, indicating yield curve inversion, the strategy exits all positions entirely and holds cash. Positions are sized equally among Bull cluster sectors and executed at the following morning's market open.

The training period spans June 2000 through December 2024 — 24.5 years including two major bear markets, the global financial crisis, a pandemic crash, and a rate hike cycle. The out-of-sample test period covers January 2025 through March 2026 and was completely sealed during model development. Two free parameters, the number of clusters $k$ and the Kalman velocity noise $Q_2$, are jointly optimized on training data using a 5 × 12 grid search evaluated by Omega ratio.

## Rationale

The core intuition is that equity sectors exhibit persistent momentum. Sectors trending strongly today tend to continue trending strongly in the near term, and sectors in downtrends tend to remain weak. Rather than attempting to forecast which sectors will outperform based on macroeconomic views, this strategy simply measures trend momentum as objectively as possible and holds the strongest-trending sectors each day.

The Kalman filter's role is noise reduction before clustering. Raw daily returns are dominated by single-day shocks — macro headlines, fund flows, temporary imbalances — that do not reflect genuine trend reversals. Clustering on raw returns generates excessive turnover as sectors bounce in and out of the Bull cluster on noise rather than signal. The Kalman velocity filters out these transient shocks by maintaining a running estimate of each sector's underlying trend speed, updating conservatively when a new price is consistent with the prior trend and more aggressively when a sustained directional move accumulates. Clustering on this denoised velocity produces more stable daily assignments and lower unnecessary turnover.

The yield curve overlay provides a macro-grounded defensive mechanism. Rather than trying to pick which sectors will hold up in a recessionary environment — a second prediction problem layered on top of the momentum signal — I simply exit the market entirely when the probability of a broad economic contraction is elevated. The 10Y-2Y spread has been a reliable leading indicator historically predicting US recessions, making it the cleanest and most widely-validated signal available.

## Research

Moskowitz & Grinblatt (1999, *Journal of Finance*) demonstrated that industry-level momentum is statistically significant and economically large, and that the profits are concentrated on the long side — validating a long-only implementation. Faber (2010, SSRN 1585517) tested relative-strength sector rotation on SPDR-equivalent data back to the 1920s and found it profitable across all lookback windows from one to twelve months, with a trend-following overlay reducing drawdowns substantially. Moskowitz, Ooi & Pedersen (2012, *Journal of Financial Economics*) established time-series momentum — each asset's own past return predicting its own future return — providing the theoretical basis for using each sector's individual velocity signal rather than only cross-sectional comparisons.

For the macro overlay, Estrella & Mishkin (1998, *Review of Economics and Statistics*) documented the yield curve's predictive power for US recessions, and the Chicago Fed (2018) confirmed that the 10Y-2Y spread has an essentially perfect historical track record as a leading recession indicator. This body of evidence grounds the yield curve filter in peer-reviewed research rather than ad hoc rule-setting.

---

# Model Implementation

## Data

```{r}
#| label: data_download
#| include: false
#| cache: true

# Download all 12 equity symbols
raw_prices <- tq_get(
  SYMBOLS,
  get  = "stock.prices",
  from = DOWNLOAD_START,
  to   = TEST_END
)

# Download FRED 10Y-2Y Treasury yield spread (T10Y2Y series)
# Positive = normal curve; negative = inverted (recession signal)
yc_raw <- tq_get(
  "T10Y2Y",
  get  = "economic.data",
  from = DOWNLOAD_START,
  to   = TEST_END
) %>%
  rename(yield_spread = price) %>%
  mutate(yield_spread = yield_spread / 100)   # convert from percentage points to decimal
```

I download daily OHLC price data for all eleven sector ETFs and SPY from Yahoo Finance, and the FRED T10Y2Y series for the yield curve overlay. XLC (Communication Services) launched on `r format(XLC_LAUNCH, "%B %d, %Y")` and XLRE (Real Estate) launched on `r format(XLRE_LAUNCH, "%B %d, %Y")`; prior to those dates those sectors are simply absent from the clustering universe, and k-means operates on the available subset.

```{r}
#| label: data_quality

coverage <- raw_prices %>%
  group_by(symbol) %>%
  summarise(
    first_date  = min(date),
    last_date   = max(date),
    n_rows      = n(),
    n_na_adj    = sum(is.na(adjusted)),
    n_na_open   = sum(is.na(open)),
    .groups     = "drop"
  ) %>%
  arrange(first_date)

kable(coverage,
      col.names = c("Symbol","First Date","Last Date","Rows","NA (Adjusted)","NA (Open)"),
      caption   = "Table 1: Data Coverage by Symbol",
      format    = "html") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed")) %>%
  row_spec(which(coverage$symbol %in% c("XLC","XLRE")), background = "#fff3cd")
```

I use the adjusted close from Yahoo Finance for the Kalman filter and C2C returns. Yahoo computes adjusted prices using the full forward history of dividends and splits, which introduces a minor look-ahead bias in the feature engineering. This is standard practice in academic finance and the impact is small for sector ETFs, which pay modest quarterly dividends and rarely split.

```{r}
#| label: data_split
#| include: false

# Training data (June 2000 – Dec 2024)
train_all    <- raw_prices %>% filter(date >= TRAIN_START, date <= TRAIN_END)
sector_train <- train_all %>% filter(symbol %in% SECTOR_SYMBOLS)
spy_train    <- train_all %>% filter(symbol == "SPY")

# Testing data (Jan 2025 – Mar 2026)
test_all    <- raw_prices %>% filter(date >= TEST_START, date <= TEST_END)
sector_test <- test_all %>% filter(symbol %in% SECTOR_SYMBOLS)
spy_test    <- test_all %>% filter(symbol == "SPY")

# Yield curve split — FRED data has gaps (weekends/holidays); forward-fill to trading days
# The full series is kept unsplit here; filtering happens inside signal generation
yc_filled <- yc_raw %>%
  arrange(date) %>%
  mutate(yield_spread = zoo::na.locf(yield_spread, na.rm = FALSE))

cat("Training rows (sectors):", nrow(sector_train), "\n")
cat("Testing rows (sectors):", nrow(sector_test), "\n")
cat("Training date range:", format(min(train_all$date)), "–", format(max(train_all$date)), "\n")
cat("Testing date range:",  format(min(test_all$date)),  "–", format(max(test_all$date)),  "\n")
```

The training and test splits are created here and the test set is not accessed again until the Performance section. All model fitting, parameter selection, and normalization are derived exclusively from training data.

## Indicators

```{r}
#| label: helper_functions
#| include: false

# ─── Kalman Filter ─────────────────────────────────────────────────────────────
run_kalman <- function(price_vec, q_level = DEFAULT_Q_LEVEL,
                        q_velocity = DEFAULT_Q_VELOCITY, r_obs = DEFAULT_R_OBS) {
  # Local level + slope state-space model
  # State: [level, velocity]; Observation: price
  # IMPORTANT: uses dlmFilter (causal), NOT dlmSmooth (uses future data)
  mod  <- dlmModPoly(order = 2, dV = r_obs, dW = c(q_level, q_velocity))
  filt <- dlmFilter(price_vec, mod)
  # filt$m is (n+1) x 2: row 1 = prior state; rows 2:(n+1) = posterior after each obs
  tibble(
    level    = filt$m[-1, 1],  # filtered level aligned with price_vec
    velocity = filt$m[-1, 2]   # filtered velocity (daily rate of change of level)
  )
}

# ─── Omega Ratio ───────────────────────────────────────────────────────────────
# Non-parametric risk-reward measure; accounts for all moments of the distribution
# Omega > 1: expected gains exceed expected losses at threshold t
# More robust than Sharpe when return distribution is non-normal
omega_ratio <- function(returns, threshold = 0) {
  r <- returns[!is.na(returns)]
  excess <- r - threshold
  gains  <- sum(excess[excess > 0])
  losses <- abs(sum(excess[excess < 0]))
  if (losses == 0) return(Inf)
  gains / losses
}

# ─── Daily Signal Generation ──────────────────────────────────────────────────
# Runs k-means on Kalman velocity every trading day (single feature, k clusters).
# Bull cluster = cluster with highest mean kal_velocity → LONG signal.
# Yield curve override: if 10Y-2Y < threshold → ALL sectors FLAT (go to cash).
# Signal at Close_T → executed at Open_{T+1} via 1-day lag in build_portfolio.
# Returns one row per sector per trading day.
generate_signals_clustered <- function(kal_df, yc_df,
                                       k_clusters = DEFAULT_K,
                                       yc_thresh  = YC_THRESHOLD) {
  # Forward-fill yield curve to every trading day
  all_trading_days <- sort(unique(kal_df$date))
  yc_daily <- tibble(date = all_trading_days) %>%
    left_join(yc_df %>% select(date, yield_spread), by = "date") %>%
    arrange(date) %>%
    mutate(yield_spread = zoo::na.locf(yield_spread, na.rm = FALSE))

  # Join yield curve into Kalman data
  kal_sector <- kal_df %>%
    filter(symbol %in% SECTOR_SYMBOLS) %>%
    left_join(yc_daily, by = "date")

  # For each trading day: k-means on kal_velocity, identify Bull cluster
  result_list <- lapply(all_trading_days, function(d) {
    df_d <- kal_sector %>% filter(date == d)
    if (nrow(df_d) < 2) return(NULL)

    yc_inv <- isTRUE(!is.na(df_d$yield_spread[1]) && df_d$yield_spread[1] < yc_thresh)

    # Yield curve inverted → all flat (cash), skip clustering
    if (yc_inv) {
      return(df_d %>%
        mutate(cluster = NA_integer_, yc_inverted = TRUE, signal = "FLAT") %>%
        select(symbol, date, signal, cluster, kal_velocity, yield_spread, yc_inverted))
    }

    # Guard: cannot have more clusters than observations
    k_actual <- min(k_clusters, nrow(df_d))
    km       <- kmeans(matrix(df_d$kal_velocity, ncol = 1),
                       centers = k_actual, nstart = 25, iter.max = 50)
    df_d$cluster <- as.integer(km$cluster)

    cluster_means <- tapply(df_d$kal_velocity, df_d$cluster, mean)
    bull_cluster  <- as.integer(names(which.max(cluster_means)))

    df_d %>%
      mutate(
        yc_inverted = FALSE,
        signal      = if_else(cluster == bull_cluster, "LONG", "FLAT")
      ) %>%
      select(symbol, date, signal, cluster, kal_velocity, yield_spread, yc_inverted)
  })

  bind_rows(result_list)
}

# ─── Portfolio Construction ────────────────────────────────────────────────────
# Signal at Close_T → executed at Open_{T+1} via 1-day lag.
# Return type depends on position transition:
#   FLAT→LONG (entry) : O2C — bought at open, held to close
#   LONG→LONG (hold)  : C2C — held from previous close to this close
#   LONG→FLAT (exit)  : C2O — held from previous close, sold at open
build_portfolio <- function(signals_df, returns_df, tc_rate = TC_RATE) {
  # Lag signal: today's exec_signal = yesterday's cluster signal
  daily <- signals_df %>%
    arrange(symbol, date) %>%
    group_by(symbol) %>%
    mutate(exec_signal = dplyr::lag(signal, default = "FLAT")) %>%
    ungroup() %>%
    left_join(returns_df %>% select(symbol, date, o2c_ret, c2c_ret, c2o_ret),
              by = c("symbol", "date")) %>%
    filter(!is.na(c2c_ret))

  # Compute equal weights based on exec_signal
  daily <- daily %>%
    group_by(date) %>%
    mutate(
      n_long = sum(exec_signal == "LONG"),
      weight = if_else(exec_signal == "LONG" & n_long > 0L, 1 / n_long, 0)
    ) %>%
    ungroup()

  # Determine position transition and select correct return type
  # prev_exec = what we were doing yesterday (did we already hold this position?)
  daily <- daily %>%
    group_by(symbol) %>%
    arrange(date, .by_group = TRUE) %>%
    mutate(
      prev_exec   = dplyr::lag(exec_signal, default = "FLAT"),
      prev_weight = dplyr::lag(weight, default = 0),
      position_ret = case_when(
        prev_exec == "FLAT" & exec_signal == "LONG" ~ o2c_ret,  # entry: Open→Close
        prev_exec == "LONG" & exec_signal == "LONG" ~ c2c_ret,  # hold:  Close→Close
        prev_exec == "LONG" & exec_signal == "FLAT" ~ c2o_ret,  # exit:  Close→Open
        TRUE                                        ~ 0
      ),
      # Exit days: position was held at prev_weight, now closing; use prev_weight for return
      eff_weight = if_else(prev_exec == "LONG" & exec_signal == "FLAT",
                           prev_weight, weight),
      tc_cost = abs(weight - prev_weight) * tc_rate
    ) %>%
    ungroup()

  daily %>%
    group_by(date) %>%
    summarise(
      port_ret_gross = sum(eff_weight * replace_na(position_ret, 0), na.rm = TRUE),
      total_tc       = sum(tc_cost, na.rm = TRUE),
      n_positions    = sum(weight > 0),
      .groups        = "drop"
    ) %>%
    mutate(port_ret_net = port_ret_gross - total_tc)
}

# ─── Performance Metrics ───────────────────────────────────────────────────────
compute_metrics <- function(port_ret_vec, spy_ret_vec = NULL, n_sectors = 11,
                            label = "", tc_rate = TC_RATE) {
  r    <- port_ret_vec[!is.na(port_ret_vec)]
  n    <- length(r)
  ann  <- 252

  cagr    <- prod(1 + r)^(ann / n) - 1
  sr      <- if (isTRUE(sd(r) > 0)) mean(r) / sd(r) * sqrt(ann) else NA_real_
  om      <- omega_ratio(r)
  cum_r   <- cumprod(1 + r)
  peak    <- cummax(cum_r)
  dds     <- (cum_r - peak) / peak
  max_dd  <- min(dds)
  in_dd   <- dds < -1e-8
  dd_rle  <- rle(in_dd)
  dd_lens <- dd_rle$lengths[dd_rle$values]
  max_dd_len <- if (length(dd_lens) > 0) max(dd_lens) else 0L
  pct_win    <- mean(r > 0, na.rm = TRUE)

  cor_spy <- if (!is.null(spy_ret_vec)) {
    s <- na.omit(cbind(r, spy_ret_vec[seq_along(r)]))
    cor(s[, 1], s[, 2])
  } else NA_real_

  tibble(
    Period          = label,
    `CAGR (%)`      = round(cagr * 100, 2),
    `Sharpe`        = round(sr, 3),
    `Omega`         = round(om, 3),
    `Max DD (%)`    = round(max_dd * 100, 2),
    `Max DD (days)` = max_dd_len,
    `% Win`         = round(pct_win * 100, 1),
    `Cor SPY`       = round(cor_spy, 3)
  )
}
```

I apply a Kalman filter (forward-only — not a Kalman smoother that uses future data) to each sector's daily adjusted close price. The model is a local level plus slope (second-order polynomial) state-space, tracking two hidden states: the sector's underlying price **level** and its **velocity** — the rate of change of that level. Each new daily price updates both estimates in proportion to how surprising it was relative to the prior.

The key tuning parameter is $Q_2$, the velocity process noise. A small $Q_2$ produces a slow-reacting, long-memory velocity; a large $Q_2$ makes it more reactive to recent moves. This parameter is optimized jointly with $k$ in the grid search. $Q_1$ (level noise) and $R$ (observation noise) are fixed.

The velocity output is used as the sole clustering feature because it is a noise-filtered measure of each sector's trend speed — stable enough that a single bad day does not flip the signal, but responsive enough to track genuine trend changes.

```{r}
#| label: apply_kalman_train
#| cache: true
#| include: false

apply_kalman <- function(df, q_vel = DEFAULT_Q_VELOCITY) {
  kal <- run_kalman(df$adjusted, q_velocity = q_vel)
  df %>%
    mutate(
      kal_level    = kal$level,
      kal_velocity = kal$velocity,
      # Return types needed for portfolio construction
      c2c_ret = adjusted / dplyr::lag(adjusted) - 1,
      o2c_ret = close    / open - 1,
      c2o_ret = open     / dplyr::lag(close)    - 1
    )
}

kalman_train <- sector_train %>%
  group_by(symbol) %>%
  arrange(date, .by_group = TRUE) %>%
  group_modify(~ apply_kalman(.x)) %>%
  ungroup()

# Return table for portfolio construction (training period only)
returns_train <- kalman_train %>%
  filter(date >= TRAIN_START, date <= TRAIN_END) %>%
  select(symbol, date, c2c_ret, o2c_ret, c2o_ret)

# SPY benchmark returns
spy_ret_train <- spy_train %>%
  arrange(date) %>%
  mutate(spy_c2c = adjusted / dplyr::lag(adjusted) - 1) %>%
  filter(!is.na(spy_c2c)) %>%
  select(date, spy_c2c)
```

## Signals

Each trading day, k-means clusters the eleven sectors by Kalman velocity into $k$ groups. The cluster with the highest mean velocity is the Bull cluster — those sectors receive a LONG signal; all others are FLAT. The cluster count $k$ is a free parameter optimized jointly with $Q_2$.

When the FRED 10Y-2Y spread falls below −0.25%, the yield curve override fires: clustering is skipped and the portfolio goes entirely to cash. This removes market exposure during elevated recession risk rather than attempting to pick resilient sectors in a downturn.

```{r}
#| label: default_ranking_inspection

# Default model: DEFAULT_K clusters, DEFAULT_Q_VELOCITY
train_signals_default <- generate_signals_clustered(
  kalman_train, yc_filled,
  k_clusters = DEFAULT_K, yc_thresh = YC_THRESHOLD
)

# Sector frequency in Bull cluster by year
rank_freq <- train_signals_default %>%
  mutate(year = year(date)) %>%
  group_by(symbol, year) %>%
  summarise(
    pct_selected = mean(signal == "LONG", na.rm = TRUE),
    .groups = "drop"
  )

hm_wide <- rank_freq %>%
  pivot_wider(names_from = year, values_from = pct_selected, values_fill = 0)
hm_mat  <- as.matrix(hm_wide[, -1])
rownames(hm_mat) <- SECTOR_NAMES[hm_wide$symbol]

plot_ly(
  x          = colnames(hm_mat),
  y          = rownames(hm_mat),
  z          = hm_mat,
  type       = "heatmap",
  colorscale = list(c(0,"#f5f5f5"), c(0.5,"#42a5f5"), c(1,"#1565c0")),
  zmin = 0, zmax = 1,
  hovertemplate = "%{y} | %{x}<br>Selected: %{z:.0%}<extra></extra>"
) %>%
  plotly::layout(
    title  = "Figure 1: Bull-Cluster Selection Frequency by Year (Training, Default Parameters)",
    xaxis  = list(title = ""),
    yaxis  = list(title = ""),
    margin = list(l = 60)
  )
```

Figure 1 shows Bull-cluster selection frequency by year under default parameters ($k$ = `r DEFAULT_K`, $Q_2$ = `r DEFAULT_Q_VELOCITY`). Warmer colors indicate more frequent selection.

The complete daily decision process is summarized in the diagram below.

```{mermaid}
%%{init: {'theme': 'base', 'themeVariables': {'primaryColor': '#e3f2fd', 'primaryTextColor': '#0d1b2a', 'primaryBorderColor': '#1565c0', 'lineColor': '#1565c0', 'secondaryColor': '#fff3e0', 'tertiaryColor': '#ffffff'}}}%%
flowchart TD
    A([Daily Close]):::start --> B["Apply Kalman Filter<br/>dlmFilter — forward pass only"]
    B --> C["Extract kal_velocity<br/>noise-filtered trend slope"]
    C --> D{"Yield curve inverted?<br/>10Y-2Y < -0.25%"}
    D -- Yes --> E["ALL FLAT<br/>Go to cash"]:::exit
    D -- No --> F["K-Means Clustering<br/>k clusters on kal_velocity"]
    F --> G["Identify Bull Cluster<br/>highest mean velocity"]:::bull
    G --> H["LONG sectors in Bull Cluster<br/>equal weight"]:::bull
    E --> I["Apply TC: 0.05%<br/>per unit weight change"]
    H --> I
    I --> J([Execute at Open T+1]):::start
    J --> A
    classDef start fill:#1565c0,stroke:#0d47a1,color:#ffffff
    classDef bull  fill:#c8e6c9,stroke:#2e7d32,color:#1b5e20
    classDef exit  fill:#ffecb3,stroke:#f57f17,color:#4e3400
```

| Condition | Signal | Execution Timing | Return Captured |
|---|---|---|---|
| Yield curve inverted (10Y-2Y < -0.25%) | ALL FLAT | Exit all positions at Open$_{T+1}$ | None (hold cash) |
| Yield curve normal AND sector in Bull cluster | LONG | Buy at Open$_{T+1}$ | Open-to-Close on $T+1$ |
| Yield curve normal AND sector NOT in Bull cluster | FLAT | Sell (if held) at Open$_{T+1}$ | None (cash = 0%) |

## Trades

Signals are generated at each day's close and executed at the following day's open. A trade only occurs when the signal changes — a sector that stays in the Bull cluster incurs no transaction cost on hold days. Return type varies by transition:

| Transition | Return Type | Formula | Rationale |
|---|---|---|---|
| FLAT → LONG (entry) | Open-to-Close (O2C) | Close / Open − 1 | Bought at open, held to close. No overnight exposure on entry day. |
| LONG → LONG (hold) | Close-to-Close (C2C) | Adj Close / Lag(Adj Close) − 1 | Already held at previous close. Earns full overnight gap plus intraday move. |
| LONG → FLAT (exit) | Close-to-Open (C2O) | Open / Lag(Close) − 1 | Signal flipped overnight. Position sold at open next day — earns only the overnight gap, not the intraday move. |

Transaction costs are 0.05% one-way ($|\Delta w_i| \times 0.0005$), covering estimated bid-ask spread and commission. Results are shown gross and net.

## Training Period

The strategy has two free parameters — $k$ (number of clusters) and $Q_2$ (Kalman velocity noise) — optimized jointly on training data by evaluating the Omega ratio across a 5 × 12 = 60 combination grid. $k$ ranges from 2 to 6; $Q_2$ is evaluated at 12 log-spaced values from $1 \times 10^{-5}$ to $1 \times 10^{-1}$ (0.00001 to 0.1). Because these values span five orders of magnitude, the optimization surface is plotted on a log₁₀ scale — so the x-axis runs from −5 to −1, where each tick represents a tenfold change in $Q_2$.

I optimize for Omega ratio rather than cumulative return because cumulative return would simply reward parameters that kept the portfolio most invested during the 2009–2021 bull market, penalizing the yield curve cash periods that are the strategy's core risk management mechanism. Omega ratio measures whether the strategy generates more dollar-weighted upside than downside at zero return, accounting for all moments of the distribution rather than just mean and variance as Sharpe does.

```{r}
#| label: param_grid_setup
#| include: false

k_grid     <- K_GRID                                                                              # 2, 3, 4, 5, 6
q_grid     <- c(0.00001, 0.00005, 0.0001, 0.0003, 0.0005, 0.001, 0.003, 0.005, 0.01, 0.03, 0.05, 0.1)  # 12 log-spaced values
param_grid <- expand.grid(k = k_grid, q = q_grid)                                                # 5 × 12 = 60 combinations
```

```{r}
#| label: parallel_optimization
#| cache: true
#| include: false

cl <- makeCluster(parallel::detectCores() - 1)
registerDoParallel(cl)

clusterExport(cl, varlist = c(
  "run_kalman", "apply_kalman", "generate_signals_clustered", "build_portfolio",
  "omega_ratio",
  "sector_train", "yc_filled",
  "TRAIN_START", "TRAIN_END", "TC_RATE",
  "SECTOR_SYMBOLS", "YC_THRESHOLD",
  "DEFAULT_Q_LEVEL", "DEFAULT_R_OBS", "DEFAULT_K"
), envir = .GlobalEnv)

opt_results <- foreach(
  idx      = 1:nrow(param_grid),
  .combine = "rbind",
  .packages = c("dlm", "dplyr", "tidyr", "tibble", "purrr", "lubridate", "zoo")
) %dopar% {

  ki <- param_grid$k[idx]
  qi <- param_grid$q[idx]

  tryCatch({
    # 1. Apply Kalman with this Q_velocity
    kal_df <- sector_train %>%
      dplyr::group_by(symbol) %>%
      dplyr::arrange(date, .by_group = TRUE) %>%
      dplyr::group_modify(~ apply_kalman(.x, q_vel = qi)) %>%
      dplyr::ungroup()

    # 2. Generate daily signals (k-means clustering on kal_velocity + YC filter)
    sigs <- generate_signals_clustered(kal_df, yc_filled,
                                       k_clusters = ki, yc_thresh = YC_THRESHOLD)

    # 3. Returns table for this Kalman parameterisation
    ret_df <- kal_df %>%
      dplyr::filter(date >= TRAIN_START, date <= TRAIN_END) %>%
      dplyr::select(symbol, date, c2c_ret, o2c_ret, c2o_ret)

    # 4. Build portfolio and evaluate
    port <- build_portfolio(sigs, ret_df, tc_rate = TC_RATE)
    r    <- na.omit(port$port_ret_net)

    if (length(r) < 50) {
      tibble::tibble(k = ki, q = qi, omega = NA_real_, sharpe = NA_real_,
                     max_dd = NA_real_, pct_win = NA_real_, n_obs = length(r))
    } else {
      cr <- cumprod(1 + r); pk <- cummax(cr)
      tibble::tibble(
        k       = ki,
        q       = qi,
        omega   = omega_ratio(r),
        sharpe  = mean(r) / sd(r) * sqrt(252),
        max_dd  = min((cr - pk) / pk),
        pct_win = mean(r > 0),
        n_obs   = length(r)
      )
    }
  }, error = function(e) {
    tibble::tibble(k = ki, q = qi, omega = NA_real_, sharpe = NA_real_,
                   max_dd = NA_real_, pct_win = NA_real_, n_obs = 0L)
  })
}

stopCluster(cl)
```

```{r}
#| label: surface_plot_omega

surface_wide <- opt_results %>%
  mutate(log_q = log10(q)) %>%
  select(k, log_q, omega) %>%
  pivot_wider(names_from = log_q, values_from = omega)

z_mat <- as.matrix(surface_wide[, -1])
rownames(z_mat) <- surface_wide$k

plot_ly(
  x          = sort(unique(log10(opt_results$q))),
  y          = sort(unique(opt_results$k)),
  z          = z_mat,
  type       = "surface",
  colorscale = list(c(0,"#d32f2f"), c(0.5,"#ffeb3b"), c(1,"#1565c0")),
  showscale  = TRUE
) %>%
  plotly::layout(
    title = list(text = "Figure 2: Omega Ratio Surface — Parameter Grid (Training Period)"),
    scene = list(
      xaxis = list(title = "log\u2081\u2080(Kalman Q\u2082 velocity noise)"),
      yaxis = list(title = "Clusters (k)"),
      zaxis = list(title = "Omega Ratio")
    )
  )
```

Figure 2 shows the Omega ratio surface across all 60 parameter combinations. A broad plateau indicates robustness — the strategy performs similarly across a range of parameter values and is not sensitive to the exact grid location of the optimum. A sharp isolated spike would suggest overfitting to a specific historical period.

```{r}
#| label: optimal_params
#| include: false

# Global optimum (k=2) sits at a steep isolated peak on the Omega surface —
# high in-sample Omega but surrounded by sharp drop-offs, indicating fragility.
# k=3 lies in a broader plateau region, is more robust to small parameter
# perturbations, and produces a more selective Bull cluster (3 regimes vs 2).
# OPTIMAL_K is fixed at 3; OPTIMAL_Q is selected as the best Q within k=3
# so it remains data-driven from the plateau rather than the spike.
best_row <- opt_results %>%
  filter(!is.na(omega), k == 3) %>%
  slice_max(omega, n = 1)

OPTIMAL_K <- 3L
OPTIMAL_Q <- best_row$q[1]

cat("Optimal parameters (training Omega criterion, k constrained to 3):\n")
cat("  k (clusters)         :", OPTIMAL_K, "\n")
cat("  Q2 (velocity noise)  :", formatC(OPTIMAL_Q, format = "e", digits = 1),
    sprintf("  [log10 = %.1f]", log10(OPTIMAL_Q)), "\n")
cat("  Training Omega       :", round(best_row$omega[1], 3), "\n")
cat("  Training Sharpe      :", round(best_row$sharpe[1], 3), "\n")
```

The global Omega maximum occurs at $k$ = 2, but that point sits at a steep, isolated peak on the surface — surrounded by sharp drop-offs that indicate fragility rather than robustness. $k$ = 3 lies in a broader plateau region of the surface, producing similar Omega values across a wider range of $Q_2$ values and partitioning sectors into three interpretable momentum regimes (strong uptrend, neutral, downtrend) rather than a coarse binary split. Accordingly, $k$ is fixed at 3 and $Q_2$ is selected as the best value within that row of the surface: $Q_2$ = `r formatC(OPTIMAL_Q, format = "e", digits = 1)` (log₁₀ = `r round(log10(OPTIMAL_Q), 1)`). These parameters are frozen and applied without adjustment to the test period.

```{r}
#| label: refit_optimal
#| cache: true
#| include: false

# Apply Kalman with OPTIMAL_Q to full training data
kalman_train_opt <- sector_train %>%
  group_by(symbol) %>%
  arrange(date, .by_group = TRUE) %>%
  group_modify(~ apply_kalman(.x, q_vel = OPTIMAL_Q)) %>%
  ungroup()

# Generate daily signals with OPTIMAL_K and OPTIMAL_Q
train_signals_opt <- generate_signals_clustered(
  kalman_train_opt, yc_filled,
  k_clusters = OPTIMAL_K, yc_thresh = YC_THRESHOLD
)

# Returns for training period
returns_train_opt <- kalman_train_opt %>%
  filter(date >= TRAIN_START, date <= TRAIN_END) %>%
  select(symbol, date, c2c_ret, o2c_ret, c2o_ret)

# Portfolio
portfolio_train <- build_portfolio(train_signals_opt, returns_train_opt, tc_rate = TC_RATE)

# Merge with SPY benchmark (spy_ret_train built in apply_kalman_train chunk)
train_combined <- portfolio_train %>%
  left_join(spy_ret_train, by = "date") %>%
  filter(!is.na(spy_c2c))
```

The in-sample results below are presented for transparency and model calibration only.

```{r}
#| label: train_metrics_table

# Count daily entry trades — days where a sector's signal flips from FLAT to LONG
n_trades_train <- train_signals_opt %>%
  group_by(symbol) %>%
  arrange(date) %>%
  mutate(entry = signal == "LONG" & dplyr::lag(signal, default = "FLAT") == "FLAT") %>%
  ungroup() %>%
  summarise(trades = sum(entry)) %>%
  pull(trades)

train_metrics_strategy <- compute_metrics(
  port_ret_vec = train_combined$port_ret_net,
  spy_ret_vec  = train_combined$spy_c2c,
  label        = "Strategy (Train)"
)
train_metrics_spy <- compute_metrics(
  port_ret_vec = train_combined$spy_c2c,
  label        = "SPY Buy & Hold (Train)"
)

bind_rows(train_metrics_strategy, train_metrics_spy) %>%
  mutate(
    `% In Market` = c(
      round(mean(portfolio_train$n_positions > 0) * 100, 1),
      100
    ),
    `# Trades` = c(n_trades_train, 0)
  ) %>%
  kable(caption = "Table 4: In-Sample Performance Metrics (Training Period — June 2000 – Dec 2024)",
        format = "html") %>%
  kable_styling(bootstrap_options = c("striped","hover"))
```

```{r}
#| label: train_cum_chart

train_cr <- train_combined %>%
  mutate(
    strategy = (cumprod(1 + port_ret_net) - 1) * 100,
    spy      = (cumprod(1 + spy_c2c)      - 1) * 100
  )

plot_ly(train_cr, x = ~date) %>%
  add_lines(y = ~strategy, name = "Strategy (net)",
            line = list(color = "#1565c0", width = 1.5)) %>%
  add_lines(y = ~spy,      name = "SPY Buy & Hold",
            line = list(color = "#b71c1c", width = 1.5)) %>%
  plotly::layout(
    title     = "Figure 3: Cumulative Returns — Training Period (In-Sample)",
    xaxis     = list(title = ""),
    yaxis     = list(title = "Cumulative Return (%)", ticksuffix = "%"),
    hovermode = "x unified",
    legend    = list(x = 0.01, y = 0.99)
  )
```

## Performance

The testing period — January 2025 through March 2026 — is the true out-of-sample evaluation. The Kalman filter state carries forward naturally from training due to it processing the data sequentially; applying it to the full concatenated price series involves no look-ahead. The optimal parameters ($k$ = `r OPTIMAL_K`, $Q_2$ = `r OPTIMAL_Q`) are fixed from training. No refitting or parameter adjustment occurs. The test set was not examined at any point prior to this section.

```{r}
#| label: build_test_features
#| cache: true
#| include: false

# Apply Kalman to full price history (train + test concatenated).
# dlmFilter is sequential — filter state at t uses only y_1..y_t; no look-ahead.
full_sector_prices <- bind_rows(sector_train, sector_test) %>%
  arrange(symbol, date)

kalman_full_opt <- full_sector_prices %>%
  group_by(symbol) %>%
  arrange(date, .by_group = TRUE) %>%
  group_modify(~ apply_kalman(.x, q_vel = OPTIMAL_Q)) %>%
  ungroup()
```

```{r}
#| label: test_clustering_and_portfolio
#| include: false

# Generate daily signals for the test period using optimal parameters
# The Kalman state from training carries forward naturally via the full concatenation
kalman_test_only <- kalman_full_opt %>%
  filter(date >= TEST_START, date <= TEST_END)

test_signals <- generate_signals_clustered(
  kalman_test_only, yc_filled,
  k_clusters = OPTIMAL_K, yc_thresh = YC_THRESHOLD
)

# Returns for test period
returns_test <- kalman_full_opt %>%
  filter(symbol %in% SECTOR_SYMBOLS, date >= TEST_START, date <= TEST_END) %>%
  select(symbol, date, c2c_ret, o2c_ret, c2o_ret)

# Build test portfolio
portfolio_test <- build_portfolio(test_signals, returns_test, tc_rate = TC_RATE)

# SPY test benchmark returns
spy_ret_test <- spy_test %>%
  arrange(date) %>%
  mutate(spy_c2c = adjusted / dplyr::lag(adjusted) - 1) %>%
  filter(!is.na(spy_c2c)) %>%
  select(date, spy_c2c)

test_combined <- portfolio_test %>%
  left_join(spy_ret_test, by = "date") %>%
  filter(!is.na(spy_c2c))

cat("Test portfolio constructed.\n")
cat("Test period:", format(min(test_combined$date)), "to", format(max(test_combined$date)), "\n")
cat("Trading days:", nrow(test_combined), "\n")
cat("Avg daily positions:", round(mean(test_combined$n_positions), 2), "\n")
cat("% Days in market:", round(mean(test_combined$n_positions > 0) * 100, 1), "%\n")
```

```{r}
#| label: test_metrics_table

n_trades_test <- test_signals %>%
  group_by(symbol) %>%
  arrange(date) %>%
  mutate(entry = signal == "LONG" & dplyr::lag(signal, default = "FLAT") == "FLAT") %>%
  ungroup() %>%
  summarise(trades = sum(entry)) %>%
  pull(trades)

test_metrics_strategy <- compute_metrics(
  port_ret_vec = test_combined$port_ret_net,
  spy_ret_vec  = test_combined$spy_c2c,
  label        = "Strategy (Test)"
)
test_metrics_spy <- compute_metrics(
  port_ret_vec = test_combined$spy_c2c,
  label        = "SPY Buy & Hold (Test)"
)

bind_rows(test_metrics_strategy, test_metrics_spy) %>%
  mutate(
    `% In Market` = c(
      round(mean(portfolio_test$n_positions > 0) * 100, 1),
      100
    ),
    `# Trades` = c(n_trades_test, 0),
    `Gross CAGR` = c(
      round((prod(1 + na.omit(test_combined$port_ret_gross))^(252/nrow(test_combined)) - 1) * 100, 2),
      NA
    )
  ) %>%
  kable(caption = "Table 5: Out-of-Sample Performance Metrics (Testing Period — Jan 2025 – Mar 2026)",
        format = "html") %>%
  kable_styling(bootstrap_options = c("striped","hover")) %>%
  footnote(general = "Strategy return = Open-to-Close of execution day. SPY = Close-to-Close. TC = 0.05% one-way.")
```

```{r}
#| label: cum_return_chart

test_cr <- test_combined %>%
  mutate(
    net   = (cumprod(1 + port_ret_net)   - 1) * 100,
    gross = (cumprod(1 + port_ret_gross) - 1) * 100,
    spy   = (cumprod(1 + spy_c2c)        - 1) * 100
  )

plot_ly(test_cr, x = ~date) %>%
  add_lines(y = ~net,   name = "Strategy (net)",
            line = list(color = "#1565c0", width = 2)) %>%
  add_lines(y = ~gross, name = "Strategy (gross)",
            line = list(color = "#42a5f5", width = 1.5, dash = "dash")) %>%
  add_lines(y = ~spy,   name = "SPY Buy & Hold",
            line = list(color = "#b71c1c", width = 2)) %>%
  plotly::layout(
    title     = "Figure 4: Cumulative Returns — Testing Period (Out-of-Sample)",
    xaxis     = list(title = ""),
    yaxis     = list(title = "Cumulative Return (%)", ticksuffix = "%"),
    hovermode = "x unified",
    legend    = list(x = 0.01, y = 0.99)
  )
```

```{r}
#| label: pa_charts

dd_df <- test_combined %>%
  mutate(
    strat_cum = cumprod(1 + port_ret_net),
    spy_cum   = cumprod(1 + spy_c2c),
    strat_dd  = (strat_cum - cummax(strat_cum)) / cummax(strat_cum) * 100,
    spy_dd    = (spy_cum   - cummax(spy_cum))   / cummax(spy_cum)   * 100,
    daily_pct = port_ret_net * 100,
    bar_color = if_else(daily_pct >= 0, "gain", "loss")
  )

# Figure 5a: Daily Returns
plot_ly(dd_df, x = ~date) %>%
  add_bars(y = ~daily_pct, color = ~bar_color,
           colors = c("gain" = "#1b5e20", "loss" = "#b71c1c"),
           showlegend = FALSE,
           hovertemplate = "%{x}<br>Daily: %{y:.2f}%<extra></extra>") %>%
  plotly::layout(
    title     = "Figure 5a: Daily Returns — Strategy (Testing Period)",
    yaxis     = list(title = "Daily Return (%)", ticksuffix = "%"),
    xaxis     = list(title = ""),
    hovermode = "x unified",
    bargap    = 0
  )
```

```{r}
#| label: drawdown_chart

# Figure 5b: Drawdown
dd_df2 <- test_combined %>%
  mutate(
    strat_cum = cumprod(1 + port_ret_net),
    spy_cum   = cumprod(1 + spy_c2c),
    strat_dd  = (strat_cum - cummax(strat_cum)) / cummax(strat_cum) * 100,
    spy_dd    = (spy_cum   - cummax(spy_cum))   / cummax(spy_cum)   * 100
  )

plot_ly(dd_df2, x = ~date) %>%
  add_lines(y = ~strat_dd, name = "Strategy",
            line = list(color = "#1565c0", width = 2)) %>%
  add_lines(y = ~spy_dd,   name = "SPY",
            line = list(color = "#b71c1c", width = 2)) %>%
  plotly::layout(
    title     = "Figure 5b: Drawdown — Strategy vs SPY (Testing Period)",
    yaxis     = list(title = "Drawdown (%)", ticksuffix = "%"),
    xaxis     = list(title = ""),
    hovermode = "x unified",
    legend    = list(x = 0.01, y = 0.05)
  )
```

```{r}
#| label: positions_chart
#| fig-height: 4

# Tile chart: each row = one sector, each column = one day, blue = LONG
pos_tile <- test_signals %>%
  mutate(
    held        = signal == "LONG",
    sector_name = SECTOR_NAMES[symbol],
    sector_name = factor(sector_name, levels = rev(sort(unique(sector_name))))
  )

p_tile <- ggplot(pos_tile, aes(x = date, y = sector_name, fill = held)) +
  geom_tile(width = 1, height = 0.85) +
  scale_fill_manual(
    values = c("TRUE" = "#1565c0", "FALSE" = "#eeeeee"),
    labels = c("TRUE" = "LONG", "FALSE" = "FLAT"),
    name   = ""
  ) +
  scale_x_date(date_breaks = "2 months", date_labels = "%b %Y") +
  labs(title = "Figure 6: Daily Sector Holdings — Testing Period",
       x = "", y = "") +
  theme_minimal(base_size = 11) +
  theme(
    legend.position  = "top",
    panel.grid       = element_blank(),
    axis.text.x      = element_text(angle = 45, hjust = 1),
    plot.title       = element_text(size = 12)
  )

ggplotly(p_tile, tooltip = c("x", "sector_name", "fill")) %>%
  layout(
    legend = list(orientation = "h", x = 0.35, y = -0.2),
    margin = list(b = 80)
  )
```

I also examine rolling correlation to SPY, which measures the strategy's diversification value relative to a passive index. A strategy with low or negative correlation adds something beyond what SPY already provides.

```{r}
#| label: rolling_correlation

# 60-day rolling Pearson correlation between strategy and SPY returns
min_window <- 30  # require at least 30 observations for correlation

roll_cor_vec <- rollapply(
  cbind(test_combined$port_ret_net, test_combined$spy_c2c),
  width     = 60,
  FUN       = function(x) {
    cc <- stats::complete.cases(x)
    if (sum(cc) < min_window) return(NA_real_)
    stats::cor(x[cc, 1], x[cc, 2])
  },
  by.column = FALSE,
  align     = "right",
  fill      = NA
)

roll_cor_df <- tibble(
  date    = test_combined$date,
  cor_60d = as.numeric(roll_cor_vec)
)

plot_ly(
  data = roll_cor_df,
  x = ~date, y = ~cor_60d,
  type = "scatter", mode = "lines",
  line = list(color = "#1565c0", width = 2),
  name = "60-day Rolling Correlation"
) %>%
  add_segments(
    x = min(roll_cor_df$date, na.rm=TRUE), xend = max(roll_cor_df$date, na.rm=TRUE),
    y = 0, yend = 0,
    line = list(color = "grey60", dash = "dash"),
    showlegend = FALSE
  ) %>%
  plotly::layout(
    title  = "Figure 8: 60-Day Rolling Correlation — Strategy vs SPY (Testing Period)",
    xaxis  = list(title = ""),
    yaxis  = list(title = "Pearson Correlation", range = c(-1, 1)),
    hovermode = "x unified"
  )

cat("Average correlation to SPY (test period):",
    round(mean(roll_cor_df$cor_60d, na.rm=TRUE), 3), "\n")
```

## Stress Testing

To test robustness, I ran three stress tests on the results. First, doubling transaction costs to 0.10% one-way tests sensitivity to the TC assumption. Second and third, excluding the GFC (2008–2009) and COVID shock (February–June 2020) from the training backtest shows whether the strategy's training-period performance depends critically on those extreme episodes.

```{r}
#| label: stress_test_computations
#| include: false

# ── Stress Test 1: Double Transaction Costs (0.10% one-way) ──────────────────
portfolio_stress_2x_tc <- build_portfolio(test_signals, returns_test, tc_rate = TC_RATE * 2)
stress_combined_2x <- portfolio_stress_2x_tc %>%
  left_join(spy_ret_test, by = "date") %>% filter(!is.na(spy_c2c))

# ── Stress Test 2: Exclude GFC (2008–2009) from TRAINING period ───────────────
returns_no_gfc  <- returns_train_opt %>%
  filter(!(date >= "2008-01-01" & date <= "2009-12-31"))
signals_no_gfc  <- train_signals_opt %>%
  filter(!(date >= "2008-01-01" & date <= "2009-12-31"))
portfolio_no_gfc <- build_portfolio(signals_no_gfc, returns_no_gfc, tc_rate = TC_RATE)
spy_no_gfc <- spy_ret_train %>%
  filter(!(date >= "2008-01-01" & date <= "2009-12-31"))
combined_no_gfc <- portfolio_no_gfc %>%
  left_join(spy_no_gfc, by = "date") %>% filter(!is.na(spy_c2c))

# ── Stress Test 3: Exclude COVID Shock (Feb–Jun 2020) from TRAINING ───────────
returns_no_covid <- returns_train_opt %>%
  filter(!(date >= "2020-02-01" & date <= "2020-06-30"))
signals_no_covid <- train_signals_opt %>%
  filter(!(date >= "2020-02-01" & date <= "2020-06-30"))
portfolio_no_covid <- build_portfolio(signals_no_covid, returns_no_covid, tc_rate = TC_RATE)
spy_no_covid <- spy_ret_train %>%
  filter(!(date >= "2020-02-01" & date <= "2020-06-30"))
combined_no_covid <- portfolio_no_covid %>%
  left_join(spy_no_covid, by = "date") %>% filter(!is.na(spy_c2c))
```

```{r}
#| label: stress_metrics_table

stress_table <- bind_rows(
  # Training period baseline
  compute_metrics(train_combined$port_ret_net, train_combined$spy_c2c,
                  label = "Training (Baseline)"),
  # Test period baseline
  compute_metrics(test_combined$port_ret_net,  test_combined$spy_c2c,
                  label = "Testing (Baseline)"),
  # 2x TC (test period)
  compute_metrics(stress_combined_2x$port_ret_net, stress_combined_2x$spy_c2c,
                  label = "Testing (2× TC)"),
  # Exclude GFC (training period analysis)
  compute_metrics(combined_no_gfc$port_ret_net, combined_no_gfc$spy_c2c,
                  label = "Training (excl. GFC 2008-09)"),
  # Exclude COVID (training period analysis)
  compute_metrics(combined_no_covid$port_ret_net, combined_no_covid$spy_c2c,
                  label = "Training (excl. COVID 2020)")
)

kable(stress_table,
      digits  = 2,
      caption = "Table 7: Stress Test Comparison — Strategy Performance Across Scenarios",
      format  = "html") %>%
  kable_styling(bootstrap_options = c("striped","hover")) %>%
  row_spec(2, bold = TRUE, background = "#e3f2fd") %>%
  footnote(general = "Testing scenarios apply the frozen model trained on full training data. Training scenarios exclude specific periods from the training backtest.")
```

```{r}
#| label: stress_cum_chart

sc_test_base  <- test_combined      %>% mutate(cum = (cumprod(1 + port_ret_net) - 1) * 100,
                                               spy_cum = (cumprod(1 + spy_c2c) - 1) * 100)
sc_test_2x    <- stress_combined_2x %>% mutate(cum = (cumprod(1 + port_ret_net) - 1) * 100)
sc_train_base <- train_combined     %>% mutate(cum = (cumprod(1 + port_ret_net) - 1) * 100,
                                               spy_cum = (cumprod(1 + spy_c2c) - 1) * 100)
sc_no_gfc     <- combined_no_gfc    %>% mutate(cum = (cumprod(1 + port_ret_net) - 1) * 100)
sc_no_covid   <- combined_no_covid  %>% mutate(cum = (cumprod(1 + port_ret_net) - 1) * 100)

# Figure 9a: Transaction cost sensitivity (test period)
plot_ly() %>%
  add_lines(x = sc_test_base$date, y = sc_test_base$cum,
            name = "Baseline (0.05% TC)",
            line = list(color = "#1565c0", width = 2)) %>%
  add_lines(x = sc_test_2x$date,   y = sc_test_2x$cum,
            name = "Double TC (0.10%)",
            line = list(color = "#e65100", width = 2, dash = "dash")) %>%
  add_lines(x = sc_test_base$date, y = sc_test_base$spy_cum,
            name = "SPY",
            line = list(color = "#757575", width = 1.5, dash = "dot")) %>%
  plotly::layout(
    title     = "Figure 9a: TC Sensitivity — Baseline vs 2\u00d7 TC (Testing Period)",
    xaxis     = list(title = ""),
    yaxis     = list(title = "Cumulative Return (%)", ticksuffix = "%"),
    hovermode = "x unified",
    legend    = list(x = 0.01, y = 0.99)
  )
```

```{r}
#| label: stress_gfc_chart

# Figure 9b: GFC exclusion (training period)
plot_ly() %>%
  add_lines(x = sc_train_base$date, y = sc_train_base$cum,
            name = "Training Baseline",
            line = list(color = "#1565c0", width = 2)) %>%
  add_lines(x = sc_no_gfc$date,     y = sc_no_gfc$cum,
            name = "Excl. GFC (2008–2009)",
            line = list(color = "#6a1b9a", width = 2, dash = "dash")) %>%
  add_lines(x = sc_train_base$date, y = sc_train_base$spy_cum,
            name = "SPY",
            line = list(color = "#757575", width = 1.5, dash = "dot")) %>%
  plotly::layout(
    title     = "Figure 9b: GFC Exclusion — Training Baseline vs Excl. GFC",
    xaxis     = list(title = ""),
    yaxis     = list(title = "Cumulative Return (%)", ticksuffix = "%"),
    hovermode = "x unified",
    legend    = list(x = 0.01, y = 0.99)
  )
```

```{r}
#| label: stress_covid_chart

# Figure 9c: COVID exclusion (training period)
plot_ly() %>%
  add_lines(x = sc_train_base$date, y = sc_train_base$cum,
            name = "Training Baseline",
            line = list(color = "#1565c0", width = 2)) %>%
  add_lines(x = sc_no_covid$date,   y = sc_no_covid$cum,
            name = "Excl. COVID (Feb–Jun 2020)",
            line = list(color = "#00695c", width = 2, dash = "dash")) %>%
  add_lines(x = sc_train_base$date, y = sc_train_base$spy_cum,
            name = "SPY",
            line = list(color = "#757575", width = 1.5, dash = "dot")) %>%
  plotly::layout(
    title     = "Figure 9c: COVID Exclusion — Training Baseline vs Excl. COVID",
    xaxis     = list(title = ""),
    yaxis     = list(title = "Cumulative Return (%)", ticksuffix = "%"),
    hovermode = "x unified",
    legend    = list(x = 0.01, y = 0.99)
  )
```

---

## Limitations

The yield curve overlay uses one FRED series with a fixed threshold. Yield curve inversion is a strong but lagging recession predictor — the inversion-to-recession lag historically ranges from 6 to 18 months, meaning the defensive rotation may activate prematurely or persist too long relative to actual market drawdowns.

K-means produces hard cluster assignments, and sectors near cluster boundaries can flip between Bull and non-Bull on small velocity changes, generating unnecessary trades even with `nstart = 25` to stabilize centroid initialization. Soft clustering methods like Gaussian Mixture Models would provide probabilistic membership and could reduce this boundary sensitivity.

The parameters are optimized once on the full training period and applied statically to the test period. A walk-forward approach — re-optimizing quarterly on a rolling window — would be more realistic and reduce the risk of regime-change invalidation as market conditions evolve.

Held sectors are equally weighted regardless of velocity magnitude within the Bull cluster. Velocity-proportional weighting would concentrate exposure on the strongest-trending sectors within the cluster rather than distributing it uniformly.

The 15-month test window is short. A single bull or bear market regime during this period will dominate the results. Definitive conclusions about strategy robustness require multiple market cycles.

## Future Work

- **Walk-forward backtesting**: Re-optimize $k$ and $Q_2$ quarterly on rolling 36-month windows to adapt to regime shifts
- **Soft clustering**: Replace k-means with Gaussian Mixture Models; weight positions by Bull cluster membership probability
- **Realized volatility as a second clustering feature**: The current clustering operates on Kalman velocity alone, and relies on $Q_2$ to filter temporal noise in the velocity estimate. However, $Q_2$ addresses smoothness over time only — the observation noise $R$ is fixed at 1.0 across all eleven sectors, meaning the Kalman filter applies uniform noise assumptions regardless of each sector's actual return dispersion. Adding realized volatility (e.g., 20-day rolling standard deviation) as a second feature would let the cluster geometry distinguish clean uptrends from erratic ones with similar mean velocity. This requires z-scoring both features before clustering to prevent realized volatility from dominating the Euclidean distance metric, and would likely shift the optimal $Q_2$ upward since some noise-filtering responsibility transfers to the volatility dimension explicitly.
- **Additional macro filters**: Credit spreads (HYG/LQD ratio), ISM PMI, or Fed funds futures to complement the yield curve signal
- **Volatility scaling**: Scale position size inversely to realized volatility, as in Moskowitz, Ooi & Pedersen (2012), to stabilize portfolio risk through regimes
